### 2018-02-23 pacmpl-zipper.tex
Introduction to the problem with recursion

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[acmsmall,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmJournal{PACMPL}
\acmVolume{1}
\acmNumber{CONF} % CONF = POPL or ICFP or OOPSLA
\acmArticle{1}
\acmYear{2018}
\acmMonth{1}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmauthoryear}   %% For author/year citations

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\newcommand{\s}{$\hphantom{~}$}
\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\vs}{\vspace{0.2cm}\\}

\begin{document}

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with single affiliation.
\author{First2 Last2}
%\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2}
  \department{Department2}             %% \department is recommended
  \institution{Institution2}           %% \institution is required
  \streetaddress{Street2 Address2}
  \city{City2}
  \state{State2}
  \postcode{Post-Code2}
  \country{Country2}                   %% \country is recommended
}
\email{first2.last2@inst2.com}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Many of the commonly used today libraries for datatype-generic programming offer fixed-point view on datatypes to express the recursive structure. The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are represented as combinations of binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view. The representation allows for generic functions that do not need access to the recursive positions in the datatype structure, but it raises issues when it needs to deal with recursion. We present a technique that uses closed type families to allow us to handle recursive occurences. Moreover, we show, by giving an advanced example, that our approach allows the \textsf{generics-sop}'s view for families of mutually recursive datatypes.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Datatype-generic programming, Sums of products, Recursion, Closed type families, Zippers, Mutually recursive datatypes, Haskell}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

The classical way to generically view data is to represent constructors by nested binary sum types while constructor arguments are represented by nested binary product types~\citep{VanNoort2008, MuRec2009, Cheney2002, Magalhaes2010}\nocite{Rodriguez2008}. \citet{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types. They call their view SOP that stands for a ``sum of products''---this is implemented in the \textsf{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}~\citep{Yorgey2012}, \emph{constraint kinds}, \emph{kind polymorphism}, and \emph{GADTs}. Using these Haskell's features, the library provides the generic view as well as a rich interface of high-level traversal combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which together encourage generic function definition in more concise and high-level style compared to the classical binary sum-of-products views.

Many generic functions require access to the recursive knots in the structure of datatypes. Some of the most general examples are \emph{maps}~\citep{Magalhaes2010} and \emph{folds}~\citep{Meijer1991, MuRec2009}, more advanced one is a \emph{zipper}~\citep{Huet1997, MuRec2009, HiJeLo2004}. For handling recursion, several generic programming approachs use different forms of fixed-point operator and represent the underlying polynomial functor~\citep{VanNoort2008, MuRec2009, Jansson1997, Loeh2011}\nocite{MagLoeh2012}. The SOP view allows to easily define functions that do not need to specially treat recursive occurences. But for those that do need it, as we show in our paper, there is a problem, so it normally does not allow. This paper introduces a method that makes it possible to define such functions within the SOP view, though the definitions are still quite complicated.

\subsection{The SOP view}

We first explain the terminology we borrow from the SOP and use throughout the paper. We show the core definitions of the SOP universe in Fig.~\ref{fig:sop}. The main idea is that a datatype is isomorphic to the sum-of-products of its code whose kind is a promoted list of lists of types \texttt{[[*]]} (here, the use of \textsf{DataKinds} extension enables datatype definitions to be promoted to kinds). This is specified using the \texttt{Code} type family. An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that provides, depending on the constructor chosen, an appropriate sequence of constructor arguments; the outer list, an $n$-ary sum, corresponds to a choice between different constructors.

A type of promoted lists has no inhabitants, so the universe provides the definitions to operate on $n$-ary sums (\texttt{NS}) and on $n$-ary products (\texttt{NP}) as on terms. The datatypes \texttt{NS} and \texttt{NP} are defined as GADTs and are indexed~\citep{HiJeLo2004} by a promoted datatype of lists. The universe defines these with built-in functor application, that is, each element of the term list is given by applying a particular type constructor (``interpretation function'') to a corresponding type in the type-level index list. The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic. The index list allows to be a list of arbitrary types of kind \texttt{k}, as the interpretation function \texttt{f} maps \texttt{k} to \texttt{*}.

A common instantiation of \texttt{f}, which is enough for our goals, is the identity functor \texttt{I}
\texttt{
\vs
newtype I (a :: *) = I \{unI :: a\}
\vs
}
that is type-level equivalent of \texttt{id} function, for which the product becomes a heterogeneous list of types.

An example value of a product looks thus:
\texttt{
\vs
I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
\vspace{0.2cm}
}

The sum constructors resemble Peano numbers, so the choice from datatype's sum of products matches the index of a particular constructor in the index list and gives the constructor being chosen. The constructor \texttt{S} skips the first element of an $n$-element index list producing an index into a list that has $n+1$ elements, while \texttt{Zero} contains the payload of type \texttt{f x}. The below is an example value of a sum.
\texttt{
\vs
S (S (Z (I 3))) :: NS I '[Char, Bool, Int, Bool]
}

\begin{figure}[t]
\begin{tabular}{cc}
\begin{tabular}[t]{l}
\tt \textbf{type} Rep a = SOP f (Code a)
\vs
\tt \textbf{class} All SListI (Code a) \Ra\\
\tt\s\s\s\s\s\s Generic (a :: *) \textbf{where}\\
\tt\s\s \textbf{type} Code a :: [[*]]\\
\tt\s\s from :: a \ra Rep a\\
\tt\s\s to\s\s\s :: Rep a \ra a
\end{tabular}
&
\begin{tabular}[t]{l}
\tt \textbf{data} NS (f :: k \ra *) (xs :: [k]) \textbf{where}\\
\tt\s\s Z :: f x \ra NS f (x ': xs)\\
\tt\s\s S :: NS f xs \ra NS f (x ': xs)
\vs
\tt \textbf{data} NP (f :: k \ra *) (xs :: [k]) \textbf{where}\\
\tt\s\s Nil\s\s :: NP f '[]\\
\tt\s\s (:*) :: f x \ra NP f xs \ra NP f (x ': xs)
\vs
\end{tabular}\\
\hline
\end{tabular}
\caption{The SOP view on data.}
\label{fig:sop}
\end{figure}

\subsection{Problem with handling recursion}

Let us illustrate the problem by giving a short example. The \textsf{QuickCheck}, a library for automatic testing of Haskell program properties, using \textsf{GHC.Generics}, defines the helper function \textsf{subterms} that obtains all the immediate subterms of a term that are of the same type as the term itself, that is, all the recursive positions in the term structure. To implement such the function using the SOP view, we have to say something like this:
\texttt{
\vs
subterms :: Generic a \Ra a \ra [a]\\
subterms t = subtermsNS (Proxy :: Proxy a) (unSOP \$ from t)
\vs
subtermsNS :: Proxy a \ra NS (NP I) xss \ra [a]\\
subtermsNS p (S ns) = subtermsNS p ns\\
subtermsNS p (Z np) = subtermsNP p np
\vs
subtermsNP :: Proxy a \ra NP I xs \ra [a]\\
subtermsNP p (I x :* xs)\\
\s\s | typeEq p x = witness p x : subtermsNP p xs\\
\s\s | otherwise\s\s = subtermsNP p xs\\
subtermsNP \_ Nil = []
}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{biblio}

%% Appendix
%\appendix
%\section{Appendix}

%Text of appendix \ldots

\end{document}

### 2018-03-07 sigplanproc-zipper.tex
Rewritten Introduction and Section 2

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For double-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review,anonymous]{acmart}\settopmatter{printfolios=true}
%% For single-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For single-blind review submission, w/ CCS and ACM Reference
%\documentclass[sigplan,review]{acmart}\settopmatter{printfolios=true}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[sigplan]{acmart}\settopmatter{}

%% Conference information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
\acmConference[PL'18]{ACM SIGPLAN Conference on Programming Languages}{January 01--03, 2018}{New York, NY, USA}
\acmYear{2018}
\acmISBN{} % \acmISBN{978-x-xxxx-xxxx-x/YY/MM}
\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
\startPage{1}

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
\citestyle{acmauthoryear}  %% For author/year citations
%\citestyle{acmnumeric}     %% For numeric citations
%\setcitestyle{nosort}      %% With 'acmnumeric', to disable automatic
                            %% sorting of references within a single citation;
                            %% e.g., \cite{Smith99,Carpenter05,Baker12}
                            %% rendered as [14,5,2] rather than [2,5,14].
%\setcitesyle{nocompress}   %% With 'acmnumeric', to disable automatic
                            %% compression of sequential references within a
                            %% single citation;
                            %% e.g., \cite{Baker12,Baker14,Baker16}
                            %% rendered as [2,3,4] rather than [2-4].

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Note: Authors migrating a paper from PACMPL format to traditional
%% SIGPLAN proceedings format must update the '\documentclass' and
%% topmatter commands above; see 'acmart-sigplanproc-template.tex'.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% Some recommended packages.
\usepackage{booktabs}   %% For formal tables:
                        %% http://ctan.org/pkg/booktabs
\usepackage{subcaption} %% For complex figures with subfigures/subcaptions
                        %% http://ctan.org/pkg/subcaption

\newcommand{\s}{$\hphantom{~}$}
\newcommand{\ind}{\s\s\s\s}
\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\vs}{\vspace{0.2cm}\\}

\begin{document}

%% Title information
\title[Short Title]{Full Title}         %% [Short Title] is optional;
                                        %% when present, will be used in
                                        %% header instead of Full Title.
%\titlenote{with title note}             %% \titlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'
%\subtitle{Subtitle}                     %% \subtitle is optional
%\subtitlenote{with subtitle note}       %% \subtitlenote is optional;
                                        %% can be repeated if necessary;
                                        %% contents suppressed with 'anonymous'

%% Author information
%% Contents and number of authors suppressed with 'anonymous'.
%% Each author should be introduced by \author, followed by
%% \authornote (optional), \orcid (optional), \affiliation, and
%% \email.
%% An author may have multiple affiliations and/or emails; repeat the
%% appropriate command.
%% Many elements are not rendered, but should be provided for metadata
%% extraction tools.

%% Author with single affiliation.
\author{First1 Last1}
%\authornote{with author1 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position1}
  \department{Department1}              %% \department is recommended
  \institution{Institution1}            %% \institution is required
  \streetaddress{Street1 Address1}
  \city{City1}
  \state{State1}
  \postcode{Post-Code1}
  \country{Country1}                    %% \country is recommended
}
\email{first1.last1@inst1.edu}          %% \email is recommended

%% Author with single affiliation.
\author{First2 Last2}
%\authornote{with author2 note}          %% \authornote is optional;
                                        %% can be repeated if necessary
\orcid{nnnn-nnnn-nnnn-nnnn}             %% \orcid is optional
\affiliation{
  \position{Position2}
  \department{Department2}             %% \department is recommended
  \institution{Institution2}           %% \institution is required
  \streetaddress{Street2 Address2}
  \city{City2}
  \state{State2}
  \postcode{Post-Code2}
  \country{Country2}                   %% \country is recommended
}
\email{first2.last2@inst2.com}         %% \email is recommended

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
Many of the commonly used today libraries for datatype-generic programming offer a fixed-point view on datatypes to express the recursive structure. Some approachs based on sums of products, however, do not use a fixed point. Those views therefore allow for generic functions that do not need access to the recursive positions in the datatype structure, but raise issues when it needs to deal with recursion. A known unwelcome solution is use of overlapping instances. We present a technique that uses closed type families to allow us to eliminate overlapping for handling recursion. Moreover, we show, by giving an advanced example, that our idiom allows for families of mutually recursive datatypes.

The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are represented as combinations of binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view. We have chosen this approach as a case study to demonstrate our solution.
\end{abstract}

%% 2012 ACM Computing Classification System (CSS) concepts
%% Generate at 'http://dl.acm.org/ccs/ccs.cfm'.
\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10011007.10011006.10011008</concept_id>
<concept_desc>Software and its engineering~General programming languages</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003456.10003457.10003521.10003525</concept_id>
<concept_desc>Social and professional topics~History of programming languages</concept_desc>
<concept_significance>300</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Software and its engineering~General programming languages}
\ccsdesc[300]{Social and professional topics~History of programming languages}
%% End of generated code

%% Keywords
%% comma separated list
\keywords{Datatype-generic programming, Sums of products, Recursion, Overlapping instances, Closed type families, Zippers, Mutually recursive datatypes, Haskell}  %% \keywords are mandatory in final camera-ready submission

%% \maketitle
%% Note: \maketitle command must come after title commands, author
%% commands, abstract environment, Computing Classification System
%% environment and commands, and keywords command.
\maketitle

\section{Introduction}

A classical way to generically view data is to represent constructors by nested binary sum types while constructor arguments are represented by nested binary product types~\citep{VanNoort2008, MuRec2009, Cheney2002, Magalhaes2010, Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}. \citet{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types. They call their view SOP that stands for a ``sum of products''---this is implemented in the \textsf{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds} and \emph{kind polymorphism}~\citep{Yorgey2012}, \emph{constraint kinds}, and \emph{GADTs}~\cite{Schrijvers2009}. Using these Haskell's features, the library provides the generic view as well as a rich interface of high-level traversal combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which together encourage generic function definition in more concise and high-level style compared to the classical binary sum-of-products views.

Many generic functions require access to the recursive knots in the structure of datatypes. Some of the most general examples are \emph{maps}~\citep{Magalhaes2010} and \emph{folds}~\citep{Meijer1991, MuRec2009}, more advanced one is a \emph{zipper}~\citep{Huet1997, MuRec2009, HiJeLo2004, Adams2010}. For handling recursion, several generic programming approachs use different forms of fixed-point operator and represent the underlying polynomial functor~\citep{VanNoort2008, MuRec2009, Jansson1997, Loeh2011}\nocite{MagLoeh2012}. The SOP view allows to easily define functions that do not need to specially treat recursive occurences. But for those that do need it, it normally does not allow.

One possible solution is to construct a more more specific universe modifying the SOP core by explicitly encoding recursive positions using a fixed point. The fixed-point approach is not fundamentally incompatible with SOP~\citep{VriLoeh2014}, but doing so makes that more complicated. Besides, such a decision may lead to need for additional conversions between the views. Our method makes it possible to define functions that need access to recursive occurences within the SOP view without modifying it anyhow. However, the definitions are still quite verbose.

Another known solution is use of overlapping instances. This is unwelcome, as involving overlapping instances complicates reasoning about the program semantics, makes it unstable because code with overlapping instances can indirectly alter the behavior, when resolving instance selection, if one adds more specific instances into scope by importing a new module. Furthermore, GHC does not reject ambiguous overlaps where neither of instances is more specific than the other and does not even report an error before attempting to use it at a type class function call site. This is also crucial in the security setting when code is compiled as safe, as GHC does not capture unsafe overlaps by detecting declaration of overlapping instances in a module and marking it unsafe~\citep{sh-overlapping}. Hence, the undesirable effect of overlapping instances becomes a problem in the setting of Safe Haskell. \citet{Kiselyov2004} show how to localize overlapping for access operations in a systematic way. Using \textit{closed type families}~\citep{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlapping for handling recursion.

We believe that our idea is suitable for a number of different sum-of-products approachs that do not exploit a fixed point---so there the described problem appears. We have chosen the \textsf{generics-sop} approach as a case study because we think it is a widely applicable library based on the interesting ideas for generic programming, and it relies on the recent GHC's extensions.

To demonstrate our approach, we give several examples. Another interesting result of our work is an interface we provide for generic functions: the type class instances with them do not need to be manually declared, as we define the instances on the generic representation of datatypes.

\paragraph{Contributions} This paper makes the following contributions:
\begin{itemize}
\item We introduce an idiom that allows to exclude use of overlapping instances for handling recursion in a number of sum-of-products approachs to datatype-generic programming that do not express recursive positions through a fixed point.
\item We present our idea by giving several examples of generic functions, particularly the generic zipper for mutually recursive datatypes, using the \textsf{generics-sop} view.
\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
\end{itemize}

\paragraph{Organization} The paper is organized as follows. In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and describe the problem. In Section~\ref{sec:handling-recursion}, we demonstrate our solution using closed type families and accompany it with several examples of generic functions. In Section~\ref{sec:generic-zipper}, we present an advanced example of the generic zipper for mutually recursive datatypes. In Section~\ref{sec:related-work}, we review related work, and we conclude in Section~\ref{sec:conclusion}.

\section{The SOP universe and the problem}
\label{sec:sop-problem}

In this section, we first review the SOP view on data describing its basic concepts to introduce the terminology we are using. After this introduction, we discuss the problem with handling recursion by generic functions and illustrate it with a short example.

\subsection{The SOP view}

We first explain the terminology we borrow from SOP\nocite{Loeh2015}\nocite{generics-sop} and use throughout the paper. The main idea of the SOP view is that a datatype is isomorphic to the sum of products of its code whose kind is a list of lists of types---it makes use of the \textsf{DataKinds} extension that enables datatype definitions to be promoted to kinds. SOP expresses this using a \texttt{Code} type family:
\texttt{
\vs
\textbf{type family} Code (a :: *) :: [[*]]
\vs
}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that provides a sequence of constructor arguments depending on the constructor chosen; the outer list, an $n$-ary sum, corresponds to a choice between different constructors.

For example, the code for the datatype of binary trees
\texttt{
\vs
data Tree a = Leaf a\\
\ind\ind\ind | Node (Tree a) (Tree a)
\vs
}
is supposed to be as follows:
\texttt{
\vs
\textbf{type instance} Code (Tree a) =\\
\ind '[ '[a]\\
\ind\s , '[Tree a, Tree a]\\
\ind\s ]
\vspace{0.2cm}
}

\begin{figure}[t]
\begin{tabular}{l}
\tt \textbf{data} NP (f :: k \ra *) (xs :: [k]) \textbf{where}\\
\tt\s\s Nil\s\s :: NP f '[]\\
\tt\s\s (:*) :: f x \ra NP f xs \ra NP f (x ': xs)
\vs
\tt \textbf{data} NS (f :: k \ra *) (xs :: [k]) \textbf{where}\\
\tt\s\s Z :: f x \ra NS f (x ': xs)\\
\tt\s\s S :: NS f xs \ra NS f (x ': xs)
\vs
\hline
\end{tabular}
\caption{Datatypes for $n$-ary sums and products.}
\label{fig:ns-np}
\end{figure}

A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms. As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are indexed~\citep{HiJeLo2004} by a promoted datatype of lists. The universe defines these with built-in functor application, where the functor is some type constructor. Each element of \texttt{NS} and \texttt{NP} is given by applying a functor to a corresponding type in the type-level index list. The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic. The index list allows to be a list of arbitrary types of kind \texttt{k}, as the functor \texttt{f} maps \texttt{k} to \texttt{*}.

As a basic instantiation of \texttt{f}, SOP defines type-level equivalent of \texttt{id}, an identity functor \texttt{I}:
\texttt{
\vs
\textbf{newtype} I (a :: *) = I \{unI :: a\}
\vs
}
When instantiated with \texttt{I}, the product becomes a direct heterogeneous list of types.

An example value of a product looks thus:
\texttt{
\vs
I 5 :* I True :* Nil :: NP I '[Int, Bool]
\vspace{0.2cm}
}

The sum constructors are similar to Peano numbers, so the choice from a datatype's sum of products matches the index of a particular constructor in the index list and gives the constructor being chosen. The constructor \texttt{S} skips the first element of an $n$-element index list producing an index into a list that has $n+1$ elements, while \texttt{Z} contains the payload of type \texttt{f x}.

For example, the following chooses the third element of a sum:
\texttt{
\vs
S (S (Z (I 3))) :: NS I '[Char, Bool, Int, Bool]
\vspace{0.2cm}
}

With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
\texttt{
\vs
\textbf{type} Rep a = SOP I (Code a)
\vs
\textbf{class} All SListI (Code a) \Ra\\
\s\s\s\s\s\s Generic (a :: *) \textbf{where}\\
\s\s \textbf{type} Code a :: [[*]]\\
\s\s from :: a \ra Rep a\\
\s\s to\s\s\s :: Rep a \ra a
\vs
}
The \texttt{SOP} is a newtype for an \texttt{NS} of an \texttt{NP}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code.

The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be always satisfied and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the type family \texttt{All} specifying that a constraint holds for all elements of a list of types. Using constraints as types of a special kind \texttt{Constraint}, hence they can appear in type families, is allowed by the \textsf{ConstraintKinds} language extension.

The \texttt{Generic} class definition in the \textsf{generics-sop} library uses a \textit{generic generic programming} technique~\citep{MagLoeh2014} to derive the instance for a particular datatype automatically using the internal \textsf{GHC.Generics} representation. Alternatively, the \textsf{generics-sop} library allows to generate the \texttt{Generic} instances using Template Haskell~\citep{SheardSPJ2002}.

\subsection{Problem with handling recursion}

Let us illustrate the problem by giving a short example. The \textsf{QuickCheck}~\citep{Claessen2011}, a library for automatic testing of Haskell program properties, using the \textsf{GHC.Generics} view, defines a helper function \texttt{subterms} that obtains all the immediate subterms of a term that are of the same type as the term itself, that is, all the recursive positions in the term structure. To implement such a function using the SOP view, we have to say something like this:
\texttt{
\vs
subterms :: Generic a \Ra a \ra [a]\\
subterms t = subtermsNS (Proxy :: Proxy a)\\
\ind\ind\ind\ind\ind\ind (unSOP \$ from t)
\vs
subtermsNS :: Proxy a \ra NS (NP I) xss \ra [a]\\
subtermsNS p (S ns) = subtermsNS p ns\\
subtermsNS p (Z np) = subtermsNP p np
\vs
subtermsNP :: Proxy a \ra NP I xs \ra [a]\\
subtermsNP p (I x :* xs)\\
\s\s | typeEq p x = witness p x : subtermsNP p xs\\
\s\s | otherwise\s\s = subtermsNP p xs\\
subtermsNP \_ Nil = []
}

\section{Handling recursion with closed type families}
\label{sec:handling-recursion}

\section{The generic Zipper}
\label{sec:generic-zipper}

\section{Related work}
\label{sec:related-work}

\section{Conclusion}
\label{sec:conclusion}

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
  This material is based upon work supported by the
  \grantsponsor{GS100000001}{National Science
    Foundation}{http://dx.doi.org/10.13039/100000001} under Grant
  No.~\grantnum{GS100000001}{nnnnnnn} and Grant
  No.~\grantnum{GS100000001}{mmmmmmm}.  Any opinions, findings, and
  conclusions or recommendations expressed in this material are those
  of the author and do not necessarily reflect the views of the
  National Science Foundation.
\end{acks}

%% Bibliography
\bibliography{biblio}

%% Appendix
%\appendix
%\section{Appendix}

%Text of appendix \ldots

\end{document}

### 2018-03-27 sigplanproc-zipper.tex
Section 3: Solution

@@
 \newcommand{\vs}{\vspace{0.2cm}\\}
+\newcommand{\fa}{$\forall$}
 
@@
 Some approachs based on sums of products, however, do not use a fixed point.
-Those views therefore allow for generic functions that do not need access to the recursive positions in the datatype structure, but raise issues when it needs to deal with recursion.
+Those views therefore allow for generic functions that do not require to look at the recursive knots in the datatype representation, but raise issues when it needs to deal with recursion.
 A known unwelcome solution is use of overlapping instances.
 We present a technique that uses closed type families to allow us to eliminate overlapping for handling recursion.
-Moreover, we show, by giving an advanced example, that our idiom allows for families of mutually recursive datatypes.
+Moreover, we show that our idiom allows for families of mutually recursive datatypes.
 
-The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are represented as combinations of binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view.
-We have chosen this approach as a case study to demonstrate our solution.
+The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are expressed by combinations of separate binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view.
+We choose this approach as a case study to demonstrate our solution.
 \end{abstract}
@@
 \section{Introduction}
+\label{sec:introduction}
 
-A classical way to generically view data is to represent constructors by nested binary sum types while constructor arguments are represented by nested binary product types~\citep{VanNoort2008, MuRec2009, Cheney2002, Magalhaes2010, Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
+A classical way to generically view data is to represent constructors as chains of nested binary sums, and constructor arguments as chains of nested binary products~\citep{VanNoort2008, MuRec2009, Cheney2002, Magalhaes2010, Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
 \citet{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types.
 They call their view SOP that stands for a ``sum of products''---this is implemented in the \textsf{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds} and \emph{kind polymorphism}~\citep{Yorgey2012}, \emph{constraint kinds}, and \emph{GADTs}~\cite{Schrijvers2009}.
-Using these Haskell's features, the library provides the generic view as well as a rich interface of high-level traversal combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which together encourage generic function definition in more concise and high-level style compared to the classical binary sum-of-products views.
+Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level traversal combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.
 
-Many generic functions require access to the recursive knots in the structure of datatypes.
+There are many generic functions that deal with the recursive knots, when traverse the structure of datatypes.
 Some of the most general examples are \emph{maps}~\citep{Magalhaes2010} and \emph{folds}~\citep{Meijer1991, MuRec2009}, more advanced one is a \emph{zipper}~\citep{Huet1997, MuRec2009, HiJeLo2004, Adams2010}.
-For handling recursion, several generic programming approachs use different forms of fixed-point operator and represent the underlying polynomial functor~\citep{VanNoort2008, MuRec2009, Jansson1997, Loeh2011}\nocite{MagLoeh2012}.
-The SOP view allows to easily define functions that do not need to specially treat recursive occurences.
+For handling recursion, several generic programming approachs use different forms of a fixed-point operator and express datatypes in the form of polynomial functors closed under fixed points~\citep{VanNoort2008, MuRec2009, Jansson1997, Loeh2011}\nocite{MagLoeh2012}.
+The SOP view allows to easily define functions that do not need to specially process recursive occurences.
 But for those that do need it, it normally does not allow.
 
-One possible solution is to construct a more more specific universe modifying the SOP core by explicitly encoding recursive positions using a fixed point.
-The fixed-point approach is not fundamentally incompatible with SOP~\citep{VriLoeh2014}, but doing so makes that more complicated.
+One possible solution is to construct a more specific universe modifying the SOP core by explicitly encoding recursive positions using a fixed point.
+The fixed-point approach can be combined with SOP~\citep{VriLoeh2014}, but doing so makes that more complicated.
 Besides, such a decision may lead to need for additional conversions between the views.
@@
 Another known solution is use of overlapping instances.
-This is unwelcome, as involving overlapping instances complicates reasoning about the program semantics, makes it unstable because code with overlapping instances can indirectly alter the behavior, when resolving instance selection, if one adds more specific instances into scope by importing a new module.
-Furthermore, GHC does not reject ambiguous overlaps where neither of instances is more specific than the other and does not even report an error before attempting to use it at a type class function call site.
-This is also crucial in the security setting when code is compiled as safe, as GHC does not capture unsafe overlaps by detecting declaration of overlapping instances in a module and marking it unsafe~\citep{sh-overlapping}.
+This is unwelcome, as involving overlapping instances complicates reasoning about the program semantics, makes it unstable because code with overlapping instances can indirectly change the behavior, if one adds more specific instances into scope by importing a new module.
+Furthermore, GHC does not reject ambiguous overlaps where neither of instance declarations is more specific than the other and does not even warn about ambiguity before attempting to use it at a type class function call site.
+This is also crucial in the security setting when code is compiled as safe, as GHC does not capture unsafe overlaps by detecting the presence of overlapping declarations and marking the module unsafe~\citep{sh-overlapping}.
 Hence, the undesirable effect of overlapping instances becomes a problem in the setting of Safe Haskell.
@@
 We believe that our idea is suitable for a number of different sum-of-products approachs that do not exploit a fixed point---so there the described problem appears.
-We have chosen the \textsf{generics-sop} approach as a case study because we think it is a widely applicable library based on the interesting ideas for generic programming, and it relies on the recent GHC's extensions.
+We choose the \textsf{generics-sop} approach as a case study because we think it is a widely applicable library based on the interesting ideas for generic programming, and it relies on GHC's recent extensions.
 
 To demonstrate our approach, we give several examples.
-Another interesting result of our work is an interface we provide for generic functions: the type class instances with them do not need to be manually declared, as we define the instances on the generic representation of datatypes.
+Another interesting result of our work, which contributes to \textsf{generics-sop}, is an interface we provide for generic functions: the type class instances with them do not need to be manually declared, as we define the instances on the generic representation of datatypes.
 
@@
 \paragraph{Organization} The paper is organized as follows.
-In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and describe the problem.
-In Section~\ref{sec:handling-recursion}, we demonstrate our solution using closed type families and accompany it with several examples of generic functions.
+In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and demonstrate the problem by a short example.
+In Section~\ref{sec:handling-recursion}, we show our solution using closed type families and accompany it with one more example of a generic function adapting the idea.
 In Section~\ref{sec:generic-zipper}, we present an advanced example of the generic zipper for mutually recursive datatypes.
@@
 \vs
-data Tree a = Leaf a\\
+\textbf{data} Tree a = Leaf a\\
 \ind\ind\ind | Node (Tree a) (Tree a)
@@
 }
-is supposed to be as follows:
+is expected to be as follows:
 \texttt{
@@
 The universe defines these with built-in functor application, where the functor is some type constructor.
-Each element of \texttt{NS} and \texttt{NP} is given by applying a functor to a corresponding type in the type-level index list.
+Thus, each component of \texttt{NS} and \texttt{NP} is given by mapping a corresponding type in the type-level list.
 The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic.
-The index list allows to be a list of arbitrary types of kind \texttt{k}, as the functor \texttt{f} maps \texttt{k} to \texttt{*}.
+The index list allows to contain types of arbitrary kind \texttt{k}, as applying the functor \texttt{f} turns \texttt{k} to \texttt{*}.
 
-As a basic instantiation of \texttt{f}, SOP defines type-level equivalent of \texttt{id}, an identity functor \texttt{I}:
+As basic instantiation types of \texttt{f}, SOP defines an identity functor \texttt{I} that is type-level equivalent for \texttt{id}, and a constant functor \texttt{K}, for \texttt{const}:
 \texttt{
 \vs
-\textbf{newtype} I (a :: *) = I \{unI :: a\}
+\textbf{newtype} I (a :: *)\ind\ind\s\s = I \{unI :: a\}\\
+\textbf{newtype} K (a :: *) (b :: k) = K \{unK :: a\}
 \vs
 }
-When instantiated with \texttt{I}, the product becomes a direct heterogeneous list of types.
-
-An example value of a product looks thus:
+Whereas instantiated with \texttt{I}, the product is a direct heterogeneous list of types, with \texttt{K a}, it is essentially homogeneous. An example value of a product for \texttt{I} looks thus:
 \texttt{
@@
 
-The sum constructors are similar to Peano numbers, so the choice from a datatype's sum of products matches the index of a particular constructor in the index list and gives the constructor being chosen.
-The constructor \texttt{S} skips the first element of an $n$-element index list producing an index into a list that has $n+1$ elements, while \texttt{Z} contains the payload of type \texttt{f x}.
-
+A clear idea underlies the sum definition.
+The constructor \texttt{S} of a sum skips the first element of an $n$-element index list giving an index in a list that has $n+1$ elements, while \texttt{Z} stores the payload of type \texttt{f x}.
 For example, the following chooses the third element of a sum:
@@
 S (S (Z (I 3))) :: NS I '[Char, Bool, Int, Bool]
-\vspace{0.2cm}
+\vs
 }
+The sum constructors are similar to Peano numbers, so the choice from a datatype's sum of products matches the index of its particular constructor in the index list and gives the product representing arguments of that constructor.
 
-With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
+With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism:
 \texttt{
@@
 }
-The \texttt{SOP} is a newtype for an \texttt{NS} of an \texttt{NP}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code.
+The sum of products \texttt{SOP f} is a newtype for \texttt{NS (NP f)}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code.
+The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.
 
-The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be always satisfied and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the type family \texttt{All} specifying that a constraint holds for all elements of a list of types.
-Using constraints as types of a special kind \texttt{Constraint}, hence they can appear in type families, is allowed by the \textsf{ConstraintKinds} language extension.
+The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be satisfied, and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the constraint \texttt{All} specifying that a particular constraint holds for each element in a list of types.
+Using constraints as types of a special kind \texttt{Constraint}, hence they can appear as type parameters, is allowed by the \textsf{ConstraintKinds} language extension.
 
-The \texttt{Generic} class definition in the \textsf{generics-sop} library uses a \textit{generic generic programming} technique~\citep{MagLoeh2014} to derive the instance for a particular datatype automatically using the internal \textsf{GHC.Generics} representation.
-Alternatively, the \textsf{generics-sop} library allows to generate the \texttt{Generic} instances using Template Haskell~\citep{SheardSPJ2002}.
+The \texttt{Generic} class definition in the \textsf{generics-sop} library uses a \textit{generic generic programming} technique~\citep{MagLoeh2014} to create the instance for a particular datatype automatically using the internal \textsf{GHC.Generics} representation.
+Alternatively, \textsf{generics-sop} allows to produce the instances using Template Haskell~\citep{SheardSPJ2002}.
 
@@
 Let us illustrate the problem by giving a short example.
-The \textsf{QuickCheck}~\citep{Claessen2011}, a library for automatic testing of Haskell program properties, using the \textsf{GHC.Generics} view, defines a helper function \texttt{subterms} that obtains all the immediate subterms of a term that are of the same type as the term itself, that is, all the recursive positions in the term structure.
+The \textsf{QuickCheck}~\citep{Claessen2011}, a library for automatic testing of Haskell code, using the \textsf{GHC.Generics} view, defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
 To implement such a function using the SOP view, we have to say something like this:
@@
 subterms :: Generic a \Ra a \ra [a]\\
-subterms t = subtermsNS (Proxy :: Proxy a)\\
-\ind\ind\ind\ind\ind\ind (unSOP \$ from t)
+subterms t = subtermsNS (unSOP \$ from t)
 \vs
-subtermsNS :: Proxy a \ra NS (NP I) xss \ra [a]\\
-subtermsNS p (S ns) = subtermsNS p ns\\
-subtermsNS p (Z np) = subtermsNP p np
+subtermsNS :: NS (NP I) xss \ra [a]\\
+subtermsNS (S ns) = subtermsNS ns\\
+subtermsNS (Z np) = subtermsNP np
 \vs
-subtermsNP :: Proxy a \ra NP I xs \ra [a]\\
-subtermsNP p (I x :* xs)\\
-\s\s | typeEq p x = witness p x : subtermsNP p xs\\
-\s\s | otherwise\s\s = subtermsNP p xs\\
+subtermsNP :: \fa a xs. NP I xs \ra [a]\\
+subtermsNP p (I y :* ys)\\
+\s\s | typeEq @a y = witnessEq y : subtermsNP ys\\
+\s\s | otherwise\s\s\s = subtermsNP ys\\
 subtermsNP \_ Nil = []
+\vs
 }
+The function \texttt{subterms} translates the term to its representation unwrapping the sum of products from \texttt{SOP} and passes that to the auxiliary function \texttt{subtermsNS} that merely traverses the sum until it reaches the product.
+Once it gets the product, it passes that further to \texttt{subtermsNP}.
+
+The definition of \texttt{subtermsNP} shows the clear idea: it traverses the product adding the element to the result list if its type is the same as a given term's, otherwise skipping that.
+We use GHC's \textsf{TypeAppications} extension, which allows \emph{visible type applications}, here to fix that type.
+Note that using this with the type variables appearing in the function signature requires explicit universal quantification.
+
+Now, we need a way to check type equality, and in the case of equal types, to witness that the element is of the appropriate type that admits adding that to the result list of subterms.
+There is a solution: we can follow \textsf{QuickCheck}'s example and implement this using overlapping instances as it does with \textsf{GHC.Generics}.
+
+The definition of \texttt{subtermsNP} we have shown serves as a clear illustration of the idea and as a template for a more satisfactory solution that we present in Section~\ref{sec:handling-recursion}.
+But right now, the simplest way to implement this is to rewrite \texttt{subtermsNP} as follows:
+\texttt{
+\vs
+\textbf{class} Subterms a (xs :: [*]) \textbf{where}\\
+\s\s subtermsNP :: NP I xs \ra [a]
+\vs
+\textbf{instance} Subterms a xs \Ra\\
+\ind\ind\s Subterms a (x ': xs) \textbf{where}\\
+\s\s subtermsNP (\_ :* xs) = subtermsNP xs\\
+\textbf{instance} \{-\# OVERLAPS \#-\} Subterms a xs \Ra\\
+\ind\ind\s Subterms a (a ': xs) \textbf{where}\\
+\s\s subtermsNP (I x :* xs)\\
+\s\s\ind = x : subtermsNP xs\\
+\textbf{instance} Subterms a '[] \textbf{where}\\
+\s\s subtermsNP \_ = []
+\vs
+}
+This uses a type class with the overloaded function.
+It needs also to declare that all the products in the datatype code are instances of \texttt{Subterms a} by adding a constraint to the \texttt{subterms} and \texttt{subtermsNS} signatures:
+\texttt{
+\vs
+subterms :: (Generic a,\\
+\ind\ind\ind\s All (Subterms a) (Code a))\\
+\ind\ind\s \Ra a \ra [a]\\
+subtermsNS :: All (Subterms a) xss\\
+\ind\ind\s\s\s \Ra NS (NP I) xss \ra [a]
+\vspace{0.2cm}
+}
+
+Though this will work, and there are a number of the Hackage packages that do it in such a manner, we want to write programs free of overlapping because of complexity it introduces into code as we have discussed in Section~\ref{sec:introduction}.
+Fortunately, we can do without overlapping, and we will demonstrate the solution in the next section.
 
@@
 \label{sec:handling-recursion}
+
+In Section~\ref{sec:sop-problem}, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances.
+We are now going to improve the solution and remove overlapping.
+
+Closed type families are the Haskell language extension introduced by~\citet{Eisenberg2014}.
+Equations for the \emph{closed type family} are disallowed to de defined outside its declaration.
+Under this extension, we can give the following definition of type-level equality:
+\texttt{
+\vs
+\textbf{type family} Equal a x :: Bool \textbf{where}\\
+\s\s Equal a a = 'True\\
+\s\s Equal a x = 'False
+\vs
+}
+The equations in a closed type family are matched in a top-to-bottom order.
+Since the order is always strictly defined, overlapping equations can be used to define sound type-level functions like this type equality.
+
+We now come back to our running example from Section~\ref{sec:sop-problem}. With type equality, we can witness the coersion between the equal types by defining a type class:
+\texttt{
+\vs
+\textbf{class} Proof (eq :: Bool)\\
+\ind\ind\ind (a :: *) (b :: *) \textbf{where}\\
+\s\s witnessEq :: b -> Maybe a
+\vs
+\textbf{instance} Proof 'False a b \textbf{where}\\
+\s\s witnessEq \_ = Nothing\\
+\textbf{instance} Proof 'True a a \textbf{where}\\
+\s\s witnessEq\s\s\s = Just
+\vs
+}
+For convenience, we provide a synonym for this. As we need the constraint can be partially applied, we define this as a class with a superclass constraint, rather than as a type synonym:
+\texttt{
+\vs
+\textbf{class} Proof (Equal a b) a b \Ra ProofEq a b\\
+\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
+\vs
+}
+The following implementation of \texttt{subtermsNP} using this proof resembles our first definition given in Section~\ref{sec:sop-problem}:
+\texttt{
+\vs
+subtermsNP :: \fa a xs. All (ProofEq a) xs\\
+\ind\ind\s\s\s \Ra NP I xs \ra [a]\\
+subtermsNP (I (y :: x) :* ys)\\
+\s\s = case witnessEq @(Equal a x) y of\\
+\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
+\ind\s\s  Nothing \ra subtermsNP  ys\\
+subtermsNP \_ Nil = []
+\vs
+}
+We also make use of \textsf{ScopedTypeVariables} in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
+A proper \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
+
+All generic functions accessing recursive knots in the underlying datatype structure can be defined in this way.
+We give another detailed example of one of such functions to show how to adapt our idiom to different scenarios.
+
+\subsection{Generic show}
+
+The function \texttt{show} is one of common examples of useful functions that traverse a datatype's recursive structure.
+It is known that this function can be defined in a generic way for an arbitrary datatype.
+We follow the implementation of the generic function \texttt{gshow} in \citet{basic-sop} for the most part, but improve it in respect of handling recursion.
+The original function \texttt{gshow} from \textsf{basic-sop} is not a substitute for the version of \texttt{show} that can be generated for a particular datatype through \texttt{\textbf{deriving} Show}, because it does not consider recursion to place parentheses at points of recursive calls---but we remove this lack of expressiveness.
+
+The following exploits the idea of \emph{pattern matching}.
+As before, we consider two cases.
+In the first case when the position we are matching is not recursive, we only require it to be the \texttt{Show} instance, and invoke its \texttt{show}, whereas in the case of the recursion point, we surround it with parentheses and recursively apply our generic function we define next.
+We again use the defined type equality to model a form of pattern matching on types:
+\texttt{
+\vs
+\textbf{class} CaseShow (eq :: Bool)\\
+\ind\ind\ind\s\s\s (a :: *) (b :: *) \textbf{where}\\
+\s\s caseShow' :: b \ra String
+\vs
+\textbf{instance} Show b \Ra CaseShow 'False a b \textbf{where}\\
+\s\s caseShow'\s\s\s = show\\
+\textbf{instance} GShow a \Ra CaseShow 'True a a \textbf{where}\\
+\s\s caseShow' t = "(" ++ gshow t ++ ")"
+\vs
+\textbf{class} CaseShow (Equal a b) a b \Ra\\
+\ind\s\s CaseRecShow a b\\
+\textbf{instance} CaseShow (Equal a b) a b \Ra\\
+\ind\ind\s CaseRecShow a b
+\vs
+caseShow :: \fa a b. CaseRecShow a b \Ra b \ra String\\
+caseShow t = caseShow' @(Equal a b) @a t
+\vs
+}
+As before, we provide the synonyms in order to reduce the constraints, as well as one for the matching function.
+
+The function \texttt{gshow} involves meta-information to show a datatype constructor's and its record fields' names.
+It uses the functions that \textsf{generics-sop} provides for handling metadata which it defines separate from the universe, as follows:
+\texttt{
+\vs
+\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
+\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
+\vs
+gshow :: \fa a. GShow a \Ra a \ra String\\
+gshow t = gshow' @a (constructorInfo\\
+\ind\ind\ind\ind\ind \$ datatypeInfo\\
+\ind\ind\ind\ind\ind \$ Proxy @a) (from t)
+\vs
+}
+The \textsf{generics-sop} library has features for deriving this meta-information automatically.
+We again define the synonym for the set of the constraints to appear in the function signature.
+The constraint \texttt{All2} is an analogue of \texttt{All} for a list of lists of types.
+The functions from \textsf{generics-sop} use a proxy to fix a type value, where we use \textsf{TypeApplications}---a later language extension than those the library relies on.
+
+The auxiliary function \texttt{gshow'} that works on the metadata encoding and generic representation uses \textsf{generics-sop}'s traversal combinators for collapsing and mapping:
+\texttt{
+\vs
+gshow' :: \fa a xss.\\
+\ind\ind\s\s (All2 (CaseRecShow a) xss, SListI xss)\\
+\ind\s\s\s \Ra NP ConstructorInfo xss\\
+\ind\s\s\s \ra SOP I xss \ra String\\
+gshow' cs (SOP sop)\\
+\ind = hcollapse\\
+\ind\s\s \$ hczipWith allp (goConstructor @a) cs sop\\
+\s\s \textbf{where}\\
+\ind allp = Proxy @(All (CaseRecShow a))
+\vs
+}
+When a sum structure actually is homogeneous (i.~e., has type \texttt{NS~(K~a)~xs}), it can be collapsed to a single component of type \texttt{a}.
+The respective instantiation of the function \texttt{hcollapse} that generalizes collapsing homogeneous structures is
+\texttt{
+\vs
+hcollapse  :: NS  (K a) xs  \ra  a
+\vs
+}
+The \texttt{hczipWith} function is generalized \texttt{zipWith} that operates with a constrained function on heterogeneous structures, where the proxy fixes the constraint:
+\texttt{
+\vs
+hczipWith :: All c xs\\
+\ind\ind\s\s \Ra proxy c\\
+\ind\ind\s\s \ra (\fa a. c a \Ra f a \ra g a \ra h a)\\
+\ind\ind\s\s \ra NP f xs \ra NS g xs \ra NS h xs
+\vs
+}
+Finally, we show functions, which process constructors and record fields and make use of \texttt{caseShow} (Figure~\ref{fig:cons-fields}).
+The function \texttt{goConstructor} repeats in general the shape of \texttt{gshow'}, where \texttt{hcmap} generalizes \texttt{map} for \texttt{NP}, \texttt{hczipWith} on two products returns \texttt{NP}, and on two cases, \texttt{hcollapse} collapses the result \texttt{NP} to a list of strings.
+
+The function \texttt{gshow} now can be used to generically show data, for example, a value of type \texttt{Tree~Bool}, where \texttt{Tree~a} (Section~\ref{sec:sop-problem}) is now assumed to be an instance of \texttt{Generic}, and of \texttt{HasDatatypeInfo}:
+\texttt{
+\vs
+**Main> \textbf{let} tree = Node (Leaf True) (Leaf False)\\
+*Main> gshow tree
+\vs
+"Node (Leaf True) (Leaf False)"
+\vs
+}
+And here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \citet{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
+\texttt{
+\vs
+\textbf{instance} Show T \textbf{where}\\
+\s\s show = gshow
+\vs
+}
+This is a consequence of that in \textsf{basic-sop}, the \texttt{gshow} function does not treat recursive positions separately, and therefore requires the \texttt{Show} constraint for all knots in the datatype structure.
+
+\begin{figure}[t]
+\begin{tabular}{l}
+\tt goConstructor :: \fa a xs. All (CaseRecShow a) xs\\
+\tt\ind\ind\ind\s\s \Ra ConstructorInfo xs \ra NP I xs\\
+\tt\ind\ind\ind\s\s \ra K String xs\\
+\tt goConstructor (Constructor n) args\\
+\tt\ind = K \$ unwords (n : args')\\
+\tt\s\s \textbf{where}\\
+\tt\ind args' :: [String]\\
+\tt\ind args' = hcollapse\\
+\tt\ind\ind\ind \$ hcmap (p @a)\\
+\tt\ind\ind\ind\ind\ind (K . caseShow @a . unI)\\
+\tt\ind\ind\ind\ind\ind args\\
+\tt goConstructor (Record n ns) args\\
+\tt\ind = K \$ n ++ " \{" ++ intercalate ", " args'\\
+\tt\ind\ind\ind ++\s\s "\}"\\
+\tt\s\s \textbf{where}\\
+\tt\ind args' :: [String]\\
+\tt\ind args' = hcollapse\\
+\tt\ind\ind\ind \$ hczipWith (p @a)\\
+\tt\ind\ind\ind\ind\ind\ind (goField @a) ns args\\
+\tt goConstructor (Infix n \_ \_)\\
+\tt\ind\ind\ind\s\s (I arg1 :* I arg2 :* Nil)\\
+\tt\ind = K \$\s\s\s\s caseShow @a arg1\\
+\tt\ind\ind\s\s ++ " " ++ n ++ " "\\
+\tt\ind\ind\s\s ++ caseShow @a arg2
+\vs
+\tt p :: Proxy (CaseRecShow a)\\
+\tt p = Proxy
+\vs
+\hline\\
+\tt goField :: \fa a x. (CaseRecShow a) x\\
+\tt\ind\ind \Ra FieldInfo x \ra I x \ra K String x\\
+\tt goField (FieldInfo field) (I y)\\
+\tt\ind = K \$ field ++ " = " ++ caseShow @a y
+\vs
+\hline
+\end{tabular}
+\caption{Processing constructors and record fields.}
+\label{fig:cons-fields}
+\end{figure}
 

### 2018-04-24 lncs-paper.tex
Section 4: Zipper

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{url}
\usepackage{amssymb}

\urlstyle{same}
% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
%\renewcommand\UrlFont{\color{blue}\rmfamily}

\newcommand{\s}{$\hphantom{~}$}
\newcommand{\ind}{\s\s\s\s}
\newcommand{\hs}{\hspace{0.06cm}}
\newcommand{\ths}{\hspace{0.01cm}}
\newcommand{\tths}{\hspace{0.003cm}}
\newcommand{\nths}{\hspace{-0.01cm}}
\newcommand{\nhs}{\hspace{-0.06cm}}
\newcommand{\vs}{\vspace{0.2cm}\\}

\newcommand{\Ra}{$\Rightarrow$\s}
\newcommand{\ra}{$\rightarrow$\s}
\newcommand{\fa}{$\forall$}
\newcommand{\tteq}{{\scriptsize$\thicksim$}\s}
\newcommand{\comp}{{\tt>\nhs>\nhs>}}
\newcommand{\kleisli}{{\tt>\nhs=\nhs>}}
\newcommand{\comps}{\comp\s}
\newcommand{\kleislis}{\kleisli\s}
\newcommand{\ann}{:\nhs:\s}

\begin{document}
%
\title{Handling Recursion in Generic Programming Using Closed Type Families}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{First Author\inst{1} \and
Second Author\inst{2}}
%
%\authorrunning{F. Author and S. Author}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany\\
\email{lncs@springer.com} \and
ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
\email{abc@uni-heidelberg.de}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
Many of the extensively used today libraries for datatype-generic programming offer a fixed-point view on datatypes to express the recursive structure. Some approachs based on sums of products, however, do not use a fixed point. Those views therefore allow for generic functions that do not require to look at the recursive knots in the datatype representation, but raise issues when it needs to deal with recursion. A known unwelcome solution is use of overlapping instances. We present a technique that uses closed type families to allow us to eliminate overlap for handling recursion. Moreover, we show that our idiom allows for families of mutually recursive datatypes.

The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are expressed by separate binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view. We choose this approach as a case study to demonstrate our solution.

\keywords{Datatype-generic programming \and Sums of products \and Recursion \and Overlapping instances \and Closed type families \and Zippers \and Mutually recursive datatypes \and Haskell.}
\end{abstract}
%
%
%
\section{Introduction}
\label{sec:introduction}

A classical way to generically view data is to represent constructors as chains of nested binary sums, and constructor arguments as chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}. De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types. They call their view SOP that stands for a ``sum of products''---this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds} and \emph{kind polymorphism}~\cite{Yorgey2012}, \emph{constraint kinds}, and \emph{GADTs}~\cite{Schrijvers2009}. Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.

There are many generic functions that deal with the recursive knots, when traverse the structure of datatypes. Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}, more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}. For handling recursion, several generic programming approachs express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}\nocite{MagLoeh2012}. The SOP view allows to easily define functions that do not need to specially process recursive occurences. But for those that do need it, it normally does not allow.

One possible solution is to construct a more specific universe modifying the SOP core by explicitly encoding recursive positions using a fixed point. The fixed-point approach can be combined with SOP~\cite{VriLoeh2014}, but doing so makes that more complicated. Besides, such a decision may lead to need for additional conversions between the views. Our method makes it possible to define functions that need access to recursive occurences within the SOP view without modifying it anyhow. However, the definitions are still quite verbose.

Another known solution is use of overlapping instances. This is unwelcome, as involving overlapping instances complicates reasoning about the code semantics, makes that unstable because it can affect the program behavior, if one adds a module with more specific instances. Morris and Jones~\cite{Morris2010} extensively discuss problems that may arise from using overlapping instances, and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap, while type classes in \textsf{ilab} are open. But openness leaves a possibility for a rise of troubles---thus, defining instances in different modules still can cause issues. The overlap problem is also crucial in the security setting when code is compiled as safe, as GHC does not capture unsafe overlaps by detecting the presence of overlapping declarations and marking the module unsafe~\cite{sh-overlapping}. Hence, the undesirable effect of overlapping instances becomes a problem in the setting of Safe Haskell. Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way. Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.

We believe that our idea is suitable for a number of different sum-of-products approachs that do not exploit a fixed point---so there the described problem appears. We choose the \textsf{generics-sop} approach as a case study because we think it is a widely applicable library based on the interesting ideas for generic programming, and it relies on GHC's recent extensions.

To demonstrate our solution, we give several examples including the generic zipper for families of mutually recursive datatypes. The approach we introduce has also advantages that contribute to \textsf{generics-sop} in addition to the main result of our work. There are examples of helpful generic functions defined in \textsf{basic-sop}~\cite{basic-sop} using the SOP view. Those definitions, however, suffer from that they require manual (boilerplate) instance declarations---as a consequence of that they do not take recursion points in a data structure into account, as we will show. We surmount this inconvenience in an interface that we provide for generic functions: the type class instances with them do not need to be manually declared. The zipper definition we offer is also very flexible for use with different families of datatypes in comparison with known generic solutions for a zipper, as it just takes a family in the form of a list of its constituent types when starts traversing a structure, and needs no boilerplate to be written for a family in addition.

\paragraph{Contributions} This paper makes the following contributions:
\begin{itemize}
\item We introduce an idiom that allows to exclude use of overlapping instances for handling recursion in a number of sum-of-products approachs to datatype-generic programming that do not express recursive positions through a fixed point.
\item We present our idea by giving several examples of generic functions, particularly the generic zipper for mutually recursive datatypes, using the \textsf{generics-sop} view.
\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
\item We develop the generic zipper that can be simply and flexibly used with different families of mutually recursive datatypes.
\end{itemize}

\paragraph{Organization} The paper is organized as follows. In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and demonstrate the problem by a short example. In Section~\ref{sec:handling-recursion}, we show our solution using closed type families and accompany it with one more example of a generic function adapting the idea. In Section~\ref{sec:generic-zipper}, we present an advanced example of the generic zipper for mutually recursive datatypes. In Section~\ref{sec:related-work}, we review related work, and we conclude in Section~\ref{sec:conclusion}.

\section{The SOP universe and the problem}
\label{sec:sop-problem}

In this section, we first review the SOP view on data describing its basic concepts to introduce the terminology we are using. Then we discuss the problem with handling recursion by generic functions and illustrate it with a short example.

\subsection{The SOP view}
\label{sec:sop-view}

We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper. The main idea of the SOP view is to use $n$-ary sums and products to view a datatype as its isomorphic code whose kind is a list of lists of types. Using the \textsf{DataKinds} extension that enables datatype definitions to be promoted to kinds, SOP expresses this code with a type family:
\texttt{
\vs
\indent\textbf{type family} Code (a \ann *) \ann [[*]]
\vs
}
An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments depending on the chosen constructor; the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.

For example, consider a datatype of binary trees:
\texttt{
\vs
\indent\textbf{data} Tree a = Leaf a | Node (Tree a) (Tree a)
\vs
}
This datatype is isomorphic to the following code:
\texttt{
\vs
\indent\textbf{type instance} Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
\vspace{0.2cm}
}

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt \textbf{data} NP (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
\tt\s\s Nil\s\s \ann NP f '[]\\
\tt\s\s (:*) \ann f x \ra NP f xs \ra NP f (x ': xs)
\vs
\tt \textbf{data} NS (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
\tt\s\s Z \ann f x \ra NS f (x ': xs)\\
\tt\s\s S \ann NS f xs \ra NS f (x ': xs)
\end{tabular}
\caption{Datatypes for $n$-ary sums and products.}
\label{fig:ns-np}
\end{figure}

A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms. As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types. The universe defines these with embedded functor application, where the functor is some type constructor. Thus, each component of \texttt{NS} and \texttt{NP} corresponds to one type (its index) in the type-level list and has type \texttt{f x}, for an index type \texttt{x}. The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic. The index list allows to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.

As basic instantiation types of \texttt{f}, SOP defines an identity functor \texttt{I} that is type-level equivalent for \texttt{id}, and a constant functor \texttt{K}, for \texttt{const}:
\texttt{
\vs
\indent\textbf{newtype} I (a \ann *)\ind\ind\s\s\nhs = I \{unI \ann a\}\\
\indent\textbf{newtype} K (a \ann *) (b \ann k) = K \{unK \ann a\}
\vs
}
Whereas instantiated with \texttt{I}, \texttt{NP} is a direct heterogeneous list of types, with \texttt{K a}, it is essentially homogeneous. An example value of an \texttt{NP} for \texttt{I} looks thus:
\texttt{
\vs
\indent I 5 :* I True :* I 'x' :* Nil \ann NP I '[Int, Bool, Char]
\vspace{0.2cm}
}

A clear idea underlies the sum definition. The constructor \texttt{S} of a sum skips the first element of an $n$-element index list giving an index in a list that has $n+1$ elements, while \texttt{Z} stores the payload of type \texttt{f x}. For example, the following chooses the third element of a sum:
\texttt{
\vs
\indent S (S (Z (I 3))) \ann NS I '[Char, Bool, Int, Bool]
\vs
}
The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and gives the product representing arguments of that constructor.

With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
\texttt{
\vs
\indent\textbf{type} Rep a = SOP I (Code a)
\vs
\indent\textbf{class} All SListI (Code a) \Ra Generic (a \ann *) \textbf{where}\\
\indent\s\s \textbf{type} Code a \ann [[*]]\\
\indent\s\s from \ann a \ra Rep a\\
\indent\s\s to\s\s\s \ann Rep a \ra a
\vs
}
The sum of products \texttt{SOP f} is a newtype for \texttt{NS (NP f)}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code. The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.

The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be satisfied, and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the constraint \texttt{All} specifying that a particular constraint holds for each member of a list of types\footnote{The definition of \texttt{All} is also kind polymorphic. Here, it applies the constraint to each inner list of types in the the outer list of lists.}. Using constraints as types of a special kind \texttt{Constraint}, hence they can appear as type parameters, is allowed by the \textsf{ConstraintKinds} language extension.

The \texttt{Generic} class definition in the \textsf{generics-sop} library allows to create the instance for a particular datatype automatically using internal \textsf{GHC.Generics}.

\subsection{Problem with handling recursion}
\label{sec:recursion-problem}

Let us illustrate the problem by giving a short example. The \textsf{QuickCheck}~\cite{Claessen2011}, a library for automatic testing of Haskell code, using the \textsf{GHC.Generics} view, defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure. To implement such a function using the SOP view, we have to say something like this:
\texttt{
\vs
\indent subterms \ann Generic a \Ra a \ra [a]\\
\indent subterms t = subtermsNS (unSOP \$ from t)
\vs
\indent subtermsNS \ann NS (NP I) xss \ra [a]\\
\indent subtermsNS (S ns) = subtermsNS ns\\
\indent subtermsNS (Z np) = subtermsNP np
\vs
\indent subtermsNP \ann \fa a xs. NP I xs \ra [a]\\
\indent subtermsNP p (I y :* ys)\\
\indent\s\s | typeEq @a y \hs\hs\ths= witnessEq y : subtermsNP ys\\
\indent\s\s | otherwise\s\s\s\hs\hs\ths = subtermsNP ys\\
\indent subtermsNP \_ Nil = []
\vs
}
The function \texttt{subterms} translates the term to its representation unwrapping the sum of products from \texttt{SOP} and passes that to the auxiliary function \texttt{subtermsNS} that merely traverses the sum until it reaches the product. Once it gets the product, it passes that further to \texttt{subtermsNP}.

The definition of \texttt{subtermsNP} shows the clear idea: it traverses the product adding the element to the result list if its type is the same as a given term's, otherwise skipping that. We use GHC's \textsf{TypeAppications} extension, which allows \emph{visible type applications}, here to fix that type. Note that using this with the type variables appearing in the function signature requires explicit universal quantification.

Now, we need a way to check type equality, and in the case of equal types, to witness that the element is of the appropriate type that admits adding that to the result list of subterms. There is a solution: we can follow \textsf{QuickCheck}'s example and implement this using overlapping instances as it does with \textsf{GHC.Generics}.

The definition of \texttt{subtermsNP} we have shown serves as a clear illustration of the idea and as a template for a more satisfactory solution that we present in Section~\ref{sec:handling-recursion}. But right now, the simplest way to implement this is to rewrite \texttt{subtermsNP} as follows:
\texttt{
\vs
\indent\textbf{class} Subterms a (xs \ann [*]) \textbf{where}\\
\indent\s\s subtermsNP \ann NP I xs \ra [a]
\vs
\indent\textbf{instance} Subterms a xs \Ra Subterms a (x ': xs) \textbf{where}\\
\indent\s\s subtermsNP (\_\s\s\s :* xs) \hs\ths= subtermsNP xs\\
\indent\textbf{instance} \{-\# OVERLAPS \#-\}\\
\indent\ind\ind\s Subterms a xs \Ra Subterms a (a ': xs) \textbf{where}\\
\indent\s\s subtermsNP (I x :* xs) = x : subtermsNP xs\\
\indent\textbf{instance} Subterms a '[] \textbf{where}\\
\indent\s\s subtermsNP \_ = []
\vs
}
This uses a type class with the overloaded function. It needs also to declare that all the products in the datatype code are instances of \texttt{Subterms a} by adding a constraint to the \texttt{subterms} and \texttt{subtermsNS} signatures:
\texttt{
\vs
\indent subterms\s\s\s \ann\hs (Generic a, All (Subterms a) (Code a))\\
\indent\ind\ind\s\s\s \Ra a \ra [a]\\
\indent subtermsNS \ann\hs All (Subterms a) xss\\
\indent\ind\ind\s\s\s \Ra NS (NP I) xss \ra [a]
\vspace{0.2cm}
}

Though this will work, and there are a number of the Hackage packages that do it in such a manner, we want to write programs free of overlap because of complexity it introduces into code as we have discussed in Section~\ref{sec:introduction}. Fortunately, we can do without overlap, and we will demonstrate the solution in the next section.

\section{Handling recursion with closed type families}
\label{sec:handling-recursion}

In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are now going to improve the solution and remove overlap.

Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}. Equations for the \emph{closed type family} are disallowed to de defined outside its declaration. Under this extension, we can give the following definition of type-level equality:
\texttt{
\vs
\indent\textbf{type family} Equal a x \ann Bool \textbf{where}\\
\indent\s\s Equal a a = 'True\\
\indent\s\s Equal a x = 'False
\vs
}
The equations in a closed type family are matched in a top-to-bottom order. Since the order is always strictly defined, overlapping equations can be used to define sound type-level functions like this type equality.

We now return to our running example from Section~\ref{sec:recursion-problem}. With type equality, we can witness the coersion between the equal types by defining a type class:
\texttt{
\vs
\indent\textbf{class} Proof (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s witnessEq \ann b \ra Maybe a
\vs
\indent\textbf{instance} Proof 'False a b \textbf{where}\\
\indent\s\s witnessEq \_ \hs\nths= Nothing\\
\indent\textbf{instance} Proof 'True\s\s a a \textbf{where}\\
\indent\s\s witnessEq\s\s\s = Just
\vs
}
We then need to provide the proof for all elements in \texttt{NP}. This can be done by means of the \texttt{All} combinator. For using that, we need the constraint can be partially applied, and so use a synonym for \texttt{Proof (Equal a b) a b} that we define as a class with a superclass constraint, rather than as a type synonym:
\texttt{
\vs
\indent\textbf{class}\ind Proof (Equal a b) a b \Ra ProofEq a b\\
\indent\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
\vs
}
The following implementation of \texttt{subtermsNP} using this proof resembles our first definition given in the previous section:
\texttt{
\vs
\indent subtermsNP \ann \fa a xs. All (ProofEq a) xs \Ra NP I xs \ra [a]\\
\indent subtermsNP (I (y \ann x) :* ys)\\
\indent\s\s = case witnessEq @(Equal a x) y of\\
\indent\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
\indent\ind\s\s  Nothing \ra subtermsNP  ys\\
\indent subtermsNP \_ Nil = []
\vs
}
We also make use of \textsf{ScopedTypeVariables} in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list. A proper \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.

All generic functions accessing recursive knots in the underlying datatype structure can be defined in this way. We give another detailed example of one of such functions to show how to adapt our idiom to different scenarios.

\subsection{Generic show}
\label{sec:gshow}

The function \texttt{show} is one of common examples of useful functions that traverse a datatype's recursive structure. It is known that this function can be defined in a generic way for an arbitrary datatype. De Vries and L\"{o}h define the generic function \texttt{gshow} in the \textsf{basic-sop}~\cite{basic-sop} package based on the SOP view. We follow their implementation of \texttt{gshow} for the most part, but improve it in respect of handling recursion. The original function \texttt{gshow} from \textsf{basic-sop} is not a substitute for the version of \texttt{show} that can be generated for a particular datatype through \texttt{\textbf{deriving} Show}, because it does not consider recursion to place parentheses at points of recursive calls---but we remove this lack of expressiveness.

The following exploits the idea of \emph{pattern matching}. As before, we consider two cases. In the first case when the position we are matching is not recursive, we only require it to be the \texttt{Show} instance, and invoke its \texttt{show}, whereas in the case of the recursion point, we surround it with parentheses and apply our generic function \texttt{gshow}. We again use the defined type equality to model a form of pattern matching on types:
\texttt{
\vs
\indent\textbf{class} CaseShow (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
\indent\s\s caseShow' \ann b \ra String
\vs
\indent\textbf{instance} Show\s\s b \Ra CaseShow 'False a b \textbf{where}\\
\indent\s\s caseShow'\s\s\s = show\\
\indent\textbf{instance} GShow a \Ra CaseShow 'True\s\s a a \textbf{where}\\
\indent\s\s caseShow' t = "(" ++ gshow t ++ ")"
\vs
}
We provide the synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, in the same way as before, as well as the synonym for the matching function:
\texttt{
\vs
\indent caseShow \ann \fa a b. CaseRecShow a b \Ra b \ra String\\
\indent caseShow t = caseShow' @(Equal a b) @a t
\vspace{0.2cm}
}

The function \texttt{gshow} involves meta-information provided by \textsf{generics-sop}'s class \texttt{HasDatatypeInfo} to show a datatype constructor's and its record fields' names. The \textsf{generics-sop} library has features for deriving this meta-information automatically.
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String
\vs
}
We define \texttt{gshow} mutually recursive with \texttt{caseShow}. The function is constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types. The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in Appendix~1.

The function \texttt{gshow} can now be used to generically show data, for example, a value of type \texttt{Tree~Bool}, where \texttt{Tree~a} (Section~\ref{sec:sop-view}) is now assumed to be an instance of \texttt{Generic}, and of \texttt{HasDatatypeInfo}:
\texttt{
\vs
\indent *Main> \textbf{let} tree = Node (Leaf True) (Leaf False)\\
\indent *Main> gshow tree
\vs
\indent "Node (Leaf True) (Leaf False)"
\vs
}
And here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
\texttt{
\vs
\indent\textbf{instance} Show T \textbf{where}\\
\indent\s\s show = gshow
\vs
}
This is a consequence of that in \textsf{basic-sop}, the \texttt{gshow} function does not treat recursive positions separately, and therefore requires the \texttt{Show} constraint for all knots in the datatype structure.

\section{The generic Zipper}
\label{sec:generic-zipper}

The zipper is a data structure that enables efficient editing and navigation within the tree-like structure of a datatype by representing a current location in that structure. The location is a focus, which can be one of recursive nodes in the tree, along with its context that consists of surroundings of the focal subtree.

The classical zipper described by Huet~\cite{Huet1997} can be generically produced for regular datatypes~\cite{HiJeLo2004}, which are a subset of datatypes that can be viewed as a least fixed point of some polynomial expression on types. Yakushev~et~al.~\cite{MuRec2009} generalize the definition of the generic zipper for an arbitrary family of mutually recursive datatypes. Those known solutions require a datatype to be expressed using forms of a fixed-point operator, since the zipper operates on recursion points. Using closed type families, it is possible to define the generic zipper using a representation that does not exploit a fixed point. In this section, we implement the generic zipper interface that includes functions for manipulating locations for mutually recursive datatypes using the SOP view.

\subsection{Interface and using}

Having the location of the zipper, which holds the current focus on one of recursive knots in the structure---for the zipper for mutually recursive datatypes, this means recursion points of the total structure of a family---and its context, we may produce a new location by moving that focus up, down, left, or right. We are first going to show the generic zipper interface and an example of how it can be used, and then proceed to implement that interface. The interface we provide for using the generic zipper is displayed in Figure~\ref{fig:zipper-interface}. This comprises the functions for \emph{movement}, \emph{starting} and \emph{ending navigation}, and \emph{updating} the focus, which are defined over the location structure.
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\textbf{Movement functions}
\vs
\tt goUp\ind \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\tt goDown\s\s \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\tt goLeft\s\s \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\tt goRight \ann Loc a fam c \ra Maybe (Loc a fam c)
\vspace{0.1cm}
\vs
\textbf{Starting navigation}
\vs
\tt enter\s\s\s \ann\hs \fa fam c a. (Generic a, In a fam, Zipper a fam c)\\
\tt\ind\ind \Ra a \ra Loc a fam c
\vspace{0.1cm}
\vs
\textbf{Ending navigation}
\vs
\tt leave\s\s\s \ann Loc a fam c \ra a
\vspace{0.1cm}
\vs
\textbf{Updating}
\vs
\tt update\s\s \ann (\fa b. c b \Ra b \ra b) \ra Loc a fam c \ra Loc a fam c
\end{tabular}
\caption{Generic zipper interface.}
\label{fig:zipper-interface}
\end{figure}

The functions \texttt{goUp}, \texttt{goDown}, \texttt{goLeft}, and \texttt{goRight} produce a location with the focus moved \emph{up} to the parent of the focal subtree, \emph{down} to its leftmost child, \emph{left} and \emph{right} to the left and right sibling, respectively, if it is possible. Movement may fail, as specified by the \texttt{Maybe} monad, if we cannot go further in a chosen direction.

The function signature of \texttt{enter} specifies the constraints necessary for starting navigation in a structure. A datatype of the structure needs to have the generic representation; the \texttt{In} constraint is defined over a generalization of \texttt{Equal} from Section~\ref{sec:handling-recursion}, which checks a type for membership of a family; the \texttt{Zipper} constraint collects specific constraints that refer to the implementation of movement operations. The universal quantifier here sets the instantiation order of the type variables for type applications that will be a part of our usage pattern for the zipper. Navigation in a tree starts at the root, and the type variable \texttt{a} refers to the root type that keeps fixed during the navigation, since it is necessary for leaving a tree and returning the root, while type in focus of the location may vary and is one of types in the type list \texttt{fam} (of kind \texttt{[*]}) specifying a family. The \texttt{leave} function ends navigation moving up to the root and returns its modified value.

The function \texttt{update} modifies the focal subtree with a constrained function, where the type in focus is existential in \texttt{Loc}, and the datatype of locations guarantees that the constraint \texttt{c} (actually, it has kind \texttt{* \ra Constraint}) holds for all datatypes in the family \texttt{fam} and therefore for all recursive nodes that can be in focus, whence the function can be applied.

Let us show how to use this with an example. Consider the following pair of mutually recursive datatypes for a rose tree and a forest, where the forest is a list of trees, and the tree is defined as a value in the node and a forest of its children:
\texttt{
\vs
\indent data RoseTree a = RTree a (Forest a)
\vs
\indent data Forest\s\s\s a = Empty | Forest (RoseTree a) (Forest a)
\vspace{0.2cm}
}

We can define a class that provides a function for updating the trees, for example:
\texttt{
\vs
\indent class UpdateTree a b where\\
\indent\s\s replaceBy \ann RoseTree a \ra b \ra b\\
\indent\s\s replaceBy \_\s\s\s = id\\
\indent instance UpdateTree a (RoseTree a) where\\
\indent\s\s replaceBy t \_ = t\\
\indent instance UpdateTree a (Forest a)
\vs
}
This replaces the tree node with a given tree, and for the forests, this leaves the nodes without change. This is also scalable if we want to extend the family.

For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and exploit the flipped function composition \comp~and Kleisli composition \kleisli, using the monad interface of \texttt{Maybe} that wraps the result type of the movement functions.
\texttt{
\vs
\indent (\comp) \ann (a \ra b) \ra (b \ra c) \ra (a \ra c)\\
\indent (\kleisli) \ann Monad m \Ra (a \ra m b) \ra (b \ra m c) \ra (a \ra m c)
\vspace{0.2cm}
}

We now define the family of our datatypes simply as follows:
\texttt{
\vs
\indent type TreeFam a = '[RoseTree a, Forest a]
\vs
}
And the example below shows how we can use the zipper operations with our updating function to change a mutually recursive expression:
\texttt{
\vs
\indent *Main> let forest\\
\indent\ind\ind\s = Forest (RTree 'a' \$ Forest (RTree 'b' Empty) Empty)\\
\indent\ind\ind\ind\ind\s\s (Forest (RTree 'x' Empty) Empty)
\vs
\indent *Main> let t = RoseTree 'c' Empty
\vs
\indent *Main> enter @(TreeFam Char) @(UpdateTree Char)\\
\indent\ind\ind\s\comps goDown \kleislis goRight \kleislis goDown\\
\indent\ind\ind\s\kleislis update (replaceBy t)\\
\indent\ind\ind\s\comps leave \comps return \$ forest
\vs
}
This yields the following result:
\texttt{
\vs
\indent Forest (RTree 'a' \$ Forest (RTree 'b' Empty) Empty)\\
\indent\ind\s\s\s (Forest (RTree 'c' Empty) Empty)
\vspace{0.2cm}
}

We also may use the zipper merely for regular datatypes, when we want to move only to recursive knots of a single datatype, just assigning a one-element type-level list to the family at the \texttt{enter} call. Or we may enumerate all types, whose nodes we want to visit, in that list. This is flexible, as this requires no boilerplate declarations. However, we demand an updating operation to be a type class function to distinguish types of the nodes, and enforce a constraint, which may be single or complex, on all types in the family, because the type of the focus is existential, as already said.

\subsection{Locations}

The location structure consists of the focal subtree, which is one of the mutually recursive nodes of total structure of the family of datatypes, and its surrounding context. We give the following definition of the datatype \texttt{Loc} for locations:
\texttt{
\vs
\indent data Loc (r \ann *) (fam \ann [*]) (c \ann * \ra Constraint) where\\
\indent\s\s Loc \ann\hs Family r a fam c \ra Contexts r a fam c\\
\indent\ind\s\s\ra Loc r fam c
\vs
}
The subtree in focus is wrapped by the \texttt{Family} datatype that we use to prove its required properties, while we do not know the current type of the node in focus. For being able to move up in the tree, we keep the type that will be obtained by uniting the current focus and context (essentially, the type of the focal node's parent in the tree): that type is unknown outside the location, and is reflected in the type parameters of \texttt{Family} and the \texttt{Contexts} datatype of the context by the type variable \texttt{a} that is existential in the \texttt{Loc} datatype; the type of the focal subtree itself is existential in \texttt{Family}. The other type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} mean the root type of the tree, the list of types constituting the family, and the constraint constructor imposing restrictions on the family's types, respectively.

\subsubsection{Families}

We define the \texttt{Family} datatype as a GADT, with embedded constraints on its constructor:
\texttt{
\vs
\indent data Family (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind (c \ann * \ra Constraint) where\\
\indent\s\s Family \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
\indent\ind\ind\s\Ra b \ra Family r a fam c
\vs
}
Though we do not know the type of the focus that is the existential type \texttt{b} in the family, we can obtain enough information about that type to provide the zipper functionality, which is captured by the restrictions above. We guarantee that the type \texttt{b} is a member of the family, and satisfies a set of constraints of the whole zipper interface, collected in the \texttt{ZipperI} constraint: in particular, it ensures that \texttt{a} is the type of the parent node for the node of focus of type \texttt{b} in the tree.

We define the \texttt{In} constraint by means of a type family \texttt{InFam} that is a generalization of \texttt{Equal} defined in Section~\ref{sec:handling-recursion}. The following checks a type for membership of a type-level list that we use to express a family:
\texttt{
\vs
\indent type family InFam (a \ann k) (fam \ann [k]) \ann Bool where\\
\indent\s\s InFam a (a ': fam) = 'True\\
\indent\s\s InFam a (x ': fam) = InFam a fam\\
\indent\s\s InFam a '[]\ind\ind = 'False
\vs
}
The \texttt{In} constraint is defined as
\texttt{
\vs
\indent type In a fam = InFam a fam \tteq 'True
\vspace{0.2cm}
}

Together with the datatype of families, we define a class \texttt{ProofIn} that provides a proof of membership of a family, which generalizes the proof of type equality from Section~\ref{sec:handling-recursion}. The definition of the \texttt{ProofIn} class and its auxiliary class \texttt{ProofFam} is given in Figure~\ref{fig:proof-in}. The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt class ProofFam (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
\tt\ind\ind\ind\s\s\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
\tt\s\s witness \ann b \ra Maybe (Family r a fam c)
\vs
\tt instance ProofFam 'False r a b fam c where\\
\tt\s\s witness \_ \hs\nths= Nothing\\
\tt instance (Generic b, In b fam, ZipperI r a b fam c)\\
\tt\ind\Ra ProofFam 'True r a b fam c where\\
\tt\s\s witness\s\s\s = Just . Family
\vs
\tt class ProofFam (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c\\
\tt instance ProofFam (InFam b fam) r a b fam c\\
\tt\ind\Ra ProofIn r a b fam c
\end{tabular}
\caption{Proof of membership of a family of datatypes.}
\label{fig:proof-in}
\end{figure}

\subsubsection{Contexts}

When we focus on a particular node in a data structure, having the surrounding context of that node is enough to reconstruct the entire structure. The context for the tree's location has the shape of the original structure of that tree but with one hole at the place of its focus, whence it is sometimes called a \emph{one-hole context}. The context can be expressed as a stack (\texttt{Contexts} below), where each frame \texttt{Context} corresponds to the particular node with a hole: it ascends from the focal node keeping its siblings, the siblings of its parent, and until it reaches the root node, through all levels in the tree. So the stack of contexts essentially reflects the track of the movement in the structure.
\texttt{
\vs
\indent data Contexts (r \ann *) (a \ann *) (fam \ann [*])\\
\indent\ind\ind\ind\s\s (c \ann * \ra Constraint) where\\
\indent\s\s CNil \ann\hs Contexts a a fam c\\
\indent\s\s Ctxs \ann\hs (Generic a, In a fam, ZipperI r x a fam c)\\
\indent\ind\s\s\s\Ra Context fam a \ra Contexts r x fam c\\
\indent\ind\s\s\s\ra Contexts r a fam c
\vs
}
This has the shape similar to the previously defined \texttt{Loc} and \texttt{Family} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Family} in the \texttt{Loc} constructor. The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame here indicates that the constraint for the zipper holds after plugging the hole, so all the properties, conversely, can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack---and the \texttt{CNil} constructor for an empty context with the \texttt{r} and \texttt{a} types equal is the inductive basis in that proof. Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype, as we do not need that.

McBride~\cite{McBride2001} studies a relation between the one-hole context definition and \emph{partial differentiation} from calculus: he shows that the type of the context for an arbitrary (regular) type can be derived mechanically from that type by means of a list of differentiation \emph{rules} that serve as formulaic instructions for computing the type in type-level programming. Yakushev~et~al.~\cite{MuRec2009} then demonstrate that his method can be generalized for mutually recursive datatypes. We adapt that technique to \textsf{generics-sop}, and now need a few auxiliary type-level functions to implement the computation of the context type. Those functions, defined recursively via type families, provide algebraic operations for lists of types (which we regard as sums and products of types): addition and multiplication. Specifically, we define addition \texttt{.++} of two sums of products (SOP) of types, multiplication \texttt{.*} of a SOP by a single type, and multiplication \texttt{.**} of a SOP by a product of types, as shown in Figure~\ref{fig:type-arithmetic}.
\begin{figure}[h]
\centering
\normalsize
\begin{tabular}{l}
\textbf{SOP addition}
\vs
\tt type family (.++) (xs \ann [[*]]) (ys \ann [[*]]) \ann [[*]] where\\
\tt\s\s (x ': xs) .++ ys = x ': (xs .++ ys)\\
\tt\s\s '[]\ind\s\s\s .++ ys = ys\\
\tt infixr 6 .++
\vspace{0.3cm}\\
\textbf{SOP-by-type multiplication}
\vs
\tt type family (.*) (x \ann *) (ys \ann [[*]]) \ann [[*]] where\\
\tt\s\s x .* (ys ': yss) = (x ': ys) ': (x .* yss)\\
\tt\s\s x .* '[]\ind\ind\s = '[]\\
\tt infixr 7 .*
\vspace{0.3cm}\\
\textbf{SOP-by-product multiplication}
\vs
\tt type family (.**) (xs \ann [*]) (ys \ann [[*]]) \ann [[*]] where\\
\tt\s\s (x ': xs) .** yss = x .* (xs .** yss)\\
\tt\s\s '[]\ind\s\s\s .** yss = yss\\
\tt infixr 7 .**
\end{tabular}
\caption{Algebraic operations on type-level sums and products.}
\label{fig:type-arithmetic}
\end{figure}

The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum. Again, kind \texttt{[*]} here denotes products, and \texttt{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers in these definitions becomes more clear if one realizes that an empty sum \texttt{'[] \ann [[*]]} corresponds to $1$, and an empty product \texttt{'[] \ann [*]} corresponds to $0$. We also specify, through the \texttt{infixr} declaration, that multiplication has a higher priority than addition.

Using the defined operations, we can implement differentiation of a product of types. The definition of differentiation shown in Figure~\ref{fig:diff-prod} resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$. When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \texttt{Hole}. Reflecting this type in the context is helpful when we traverse the context representation to plug the hole. We use the empty type \texttt{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice. The sum \texttt{'[ '[]]} represents a unit type that is exactly $1$. We also use type-level \texttt{If} that returns its second argument for \texttt{'True}, and the third one otherwise. We do not give its definition here, as it is straightforward.
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt data Hole = Hole\\
\tt data End
\vs
\tt type family DiffProd (fam \ann [*]) (xs \ann [*]) \ann [[*]] where\\
\tt\s\s DiffProd fam '[]\ind\s\s\s = '[]\\
\tt\s\s DiffProd fam '[x]\ind\s\s = If (InFam x fam) '[ '[Hole]] '[]\\
\tt\s\s DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]\ind\s '[]\\
\tt\s\s DiffProd fam (x ': xs)\\
\tt\ind = Hole .* xs .** DiffProd fam '[End, x]  .++\\
\tt\ind\ind\ind\s\s x\s\s .*\s\s DiffProd fam xs
\end{tabular}
\caption{Differentiation of a product of types.}
\label{fig:diff-prod}
\end{figure}

The following completes the computation of the context type:
\texttt{
\vs
\indent data ConsN = F | N ConsN | None\\
\indent\s\s deriving Eq
\vs
\indent type family ToContext (n \ann ConsN) (fam \ann [*])\\
\indent\ind\ind\ind\ind\ind\s\s (code \ann [[*]]) \ann [[*]] where\\
\indent\s\s ToContext n fam '[] = '[]\\
\indent\s\s ToContext n fam (xs ': xss)\\
\indent\ind = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
\vs
}
The type family \texttt{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code. Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype. We store the index in the datatype \texttt{ConsN} adding that to the head of each product but wrapped by \texttt{Proxy} because we use \texttt{ConsN} promoted to a kind in the type family, while the products contain types of kind \texttt{*}. The constructors \texttt{F} and \texttt{N} denote ``first'' and ``next'', respectively, and the special constructor \texttt{None} will be used further to indicate failure of matching indices.

We finally define \texttt{Context} as a newtype that wraps the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper interface functions.
\texttt{
\vs
\indent type CtxCode  fam a = ToContext 'F fam (Code a)
\vs
\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
\vs
}
The \texttt{CtxCode} type is the computed type of the context for the given code of a datatype. We will use this type synonym further when defining constraints for functions providing the generic zipper interface.

\subsection{Implementing the zipper interface}

We now can implement the interface functions of the zipper, which we have previously described. We only demonstrate the implementation of the \texttt{goDown} function here. This shows the idea of how we can use our idiom for defining the zipper functions, and the source code with the full implementation of the zipper interface is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.

To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context. The following definition of \texttt{goDown} uses two auxiliary functions: \texttt{toFirst} and \texttt{toFirstCtx}.
\texttt{
\vs
\indent goDown \ann Loc a fam c \ra Maybe (Loc a fam c)\\
\indent goDown (Loc (Family t) cs)\\
\indent\s\s = case toFirst t of\\
\indent\ind\s\s Just t' \ra Just \$ Loc t' (Ctxs (toFirstCtx t) cs)\\
\indent\ind\s\s\_\ind\s\s\s\hs\ra Nothing
\vs
}
The function \texttt{toFirst} returns its result in the \texttt{Maybe} monad, and may return \texttt{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down. The function \texttt{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.

\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\tt toFirst \ann\hs\fa fam c r a. (Generic a, ToFirst r a fam c)\\
\tt\ind\ind\Ra a \ra Maybe (Family r a fam c)\\
\tt toFirst t = appToNP @AllProof toFirstNP \$ unSOP \$ from t
\vspace{0.3cm}\\
\bf Proof
\vs
\tt class\ind All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs\\
\tt instance All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs
\vs
\tt type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
\vspace{0.3cm}\\
\bf Processing products
\vs
\tt toFirstNP \ann\hs\fa fam c r a xs. All (ProofIn r a fam c) xs\\
\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Family r a fam c)\\
\tt toFirstNP (I (x \ann b) :* xs)\\
\tt\s\s = witness @(InFam b fam) x `mplus` toFirstNP xs\\
\tt toFirstNP Nil = Nothing
\end{tabular}
\caption{Implementation of \texttt{toFirst}.}
\label{fig:toFirst}
\end{figure}

\subsubsection{toFirst}

We first implement the function \texttt{toFirst}. Its full definition is displayed in Figure~\ref{fig:toFirst}. The \texttt{toFirst} function uses the higher-oredered function \texttt{appToNP} that unwraps the product from \texttt{NS} and applies the given function to that product. The function \texttt{toFirstNP} is defined recursively using the proof we have defined for families: it traverses the product until it finds the first recursive node by means of \texttt{witness} that for the node \texttt{x} of unknown type \texttt{b}, witnesses its membership of the family, or else returns \texttt{Nothing}. To provide the proof for the representation code of a datatype, we define the proof for all products in a sum, and pass this proof through explicit type application to \texttt{appToNP} which takes a constrained function. The \texttt{appToNP} function is defined similarly to \texttt{subtermsNS} from Section~\ref{sec:recursion-problem}, and we omit its definition here.

\subsubsection{toFirstCtx}

The definition of \texttt{toFirstCtx} is more complicated, as it performs the computation of the context. The implementation comprises several steps including type- and term-level programming. In the following, we systematically construct the context representation from the given generic representation of a datatype.

At first, for a datatype's given constructor represented by \texttt{NP}, we build its matching constructor of the context. All constructors of the context have the same shape as the constructors of its respective datatype but with a hole at one of points of recursion---we are now computing the context with the first recursive node deleted. Assuming that we have the product type for the context computed, we can compute the product by matching on that type:
\texttt{
\vs
\indent class FromFstRec (ys \ann [*]) (xs \ann [*]) where\\
\indent\s\s fromFstRec \ann NP I xs \ra NP I ys
\vs
\indent instance FromFstRec (Hole ': xs) (x ': xs) where\\
\indent\s\s fromFstRec (\_ \hs\nths:* xs) = I Hole :* xs\\
\indent instance FromFstRec ys xs\\
\indent\ind\Ra FromFstRec (x ': ys) (x ': xs) where\\
\indent\s\s fromFstRec (x :* xs) = x\ind\s\s :* fromFstRec xs\\
\indent instance FromFstRec '[] '[] where\\
\indent\s\s fromFstRec \_\ind\ind\s \hs\nths= Nil
\vs
}
Note that the first type parameter in the \texttt{FromFstRec} class is the index type list of the result \texttt{NP}---this order of type variables remains for type application through which we will supply the computed type.

Once we have constructed the product, we have to build the sum representing the choice of that product. If we have the index of the chosen constructor for the datatype, the one that we find for its context can be computed as follows:
\texttt{
\vs
\indent type family CtxConsN (xss \ann [[*]]) (n \ann ConsN) \ann ConsN where\\
\indent\s\s CtxConsN '[]\ind\ind\ind\ind\ind\s\s\s n = 'None\\
\indent\s\s CtxConsN ((Proxy n \s': xs) ': xss) n = 'F\\
\indent\s\s CtxConsN ((Proxy n' ': xs) ': xss) n = 'N (CtxConsN xss n)
\vs
}
The list \texttt{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.

To construct the sum with the computed index and product, we again need a proof to witness that the product is type-consistent with the constructor choice. To choose the constructor from the context code, we can adopt \textsf{generics-sop}'s \emph{injections}\footnote{The actual definition of the type of injections in \textsf{generics-sop} slightly differs from this. We adapt that for presentation.}:
\texttt{
\vs
\indent injections \ann \fa xs f. SListI xs \Ra NP (Injection f xs) xs
\vs
\indent newtype Injection f xs a = Inj \{apInj \ann f a \ra K (NS f xs) a\}
\vs
}
For \texttt{f} instantiated to \texttt{NP I} and \texttt{xs} to be the sum of products reflecting a representation code, \texttt{injections} creates a product of all constructor choices that inject appropriate constructor arguments into each sum.

We now can witness the choice using the proof of type equality we have defined in Section~\ref{sec:handling-recursion}. In the following definition, \texttt{f} is supposed to be instantiated to \texttt{Injection yss} where \texttt{yss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \texttt{ConsN} instead of defining a type class function. However, when using a singleton, the recursive definition of \texttt{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
\texttt{
\vs
\indent class\ind Proof (Equal a b) (f a) (f b) \Ra ProofF f a b\\
\indent instance Proof (Equal a b) (f a) (f b) \Ra ProofF f a b
\vs
\indent class ConsNInj (n \ann ConsN) (xs \ann [*]) where\\
\indent\s\s consNInj \ann All (ProofF f xs) yss \Ra NP f yss \ra f xs
\vs
\indent instance ConsNInj n xs \Ra ConsNInj ('N n) xs where\\
\indent\s\s consNInj (\_ :* yss) = consNInj @n yss\\
\indent\s\s consNInj Nil\ind\s\s\s\hs \hs\ths\tths= impossible\\
\indent instance ConsNInj 'F xs where\\
\indent\s\s consNInj ((ys \ann f ys) :* \_)\\
\indent\ind = fromMaybe impossible \$ witnessEq @(Equal xs ys) ys\\
\indent\s\s consNInj Nil = impossible\\
\indent instance ConsNInj 'None xs where\\
\indent\s\s consNInj \_\s\s\s \hs\nths= impossible
\vs
\indent impossible \ann a\\
\indent impossible =  error "impossible"
\vs
}
As long as the constructor index and product are computed correctly, the proof should never fail (\texttt{impossible}).

Finally, we define a function that applies the injection and returns the constructed context for the given product, representing its constructor arguments, and index of that constructor:
\texttt{
\vs
\indent type AppInj n xs code = (ConsNInj n xs, SListI code,\\
\indent\ind\ind\ind\ind\ind\ind\s All (ProofF (Injection code) xs) code)
\vs
\indent appInjCtx \ann \fa a n xs fam. AppInj n xs (CtxCode fam a)\\
\indent\ind\ind\s\s \Ra NP I xs \ra Context fam a\\
\indent appInjCtx np\\
\indent\s\s = Ctx \$ SOP \$ unK \$ apInj (consNInj @n injections) np
\vspace{0.2cm}
}

The definition of \texttt{toFirstCtx} can now be given using \texttt{appInjCtx}. The following uses a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurence in the given datatype's product list with type \texttt{Hole}. Its definition is straightforward, and is omitted here.
\texttt{
\vs
\indent toFirstCtx \ann \fa fam c a. (Generic a, ToFirstCtx fam a)\\
\indent\ind\ind\s\s\s \Ra a \ra Context fam a\\
\indent toFirstCtx t = toFirstCtxNS @'F \$ unSOP \$ from t
\vs
\indent toFirstCtxNS \ann \fa n fam a xss. ToFirstCtx' fam a n xss\\
\indent\ind\ind\ind\s \Ra NS (NP I) xss \ra Context fam a\\
\indent toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns\\
\indent toFirstCtxNS (Z (np \ann NP I xs))\\
\indent\s\s = appInjCtx @a @(CtxConsN (CtxCode fam a) n) \$\\
\indent\ind\ind I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
\vs
\indent type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
\vs
}
The \texttt{ToFirstCtx'} constraint here is a complex constraint for use of \texttt{appInjCtx} and \texttt{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints. Its definition is shown in Appendix~2. We omit this here for reasons of space.

We have established the mechanism of translation, which is cumbersome but verifying, via the proof, that the constructed context will be correct for any given datatype and family. The other functions in the zipper interface can be implemented according to this technique.

\section{Related work}
\label{sec:related-work}

Related work

\section{Conclusion}
\label{sec:conclusion}

\section*{Appendix 1}

In this appendix, we show the implementation of the generic function \texttt{gshow}, which uses \texttt{caseShow}, described in Section~\ref{sec:gshow}. The \texttt{gshow} function is originally defined by De Vries and L\"{o}h in the \textsf{basic-sop} package~\cite{basic-sop}. Their implementation has the issue that for recursive datatypes whose recursive nodes should be displayed within parentheses, the function acts incorrectly due to the SOP view's limitations on expressing recursion. We improve the implementation from \textsf{basic-sop} with our approach to handling recursion.

The \texttt{gshow} function makes use of metadata provided by the \texttt{generics-sop} library through the class \texttt{HasDatatypeInfo}. That class delivers generic functions for handling meta-information which \textsf{generics-sop} defines separate from the main generic representation of a datatype. The definition of \texttt{gshow} using those functions is as follows:
\texttt{
\vs
\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
\vs
\indent gshow \ann \fa a. GShow a \Ra a \ra String\\
\ind\ind\s\Ra a \ra String\\
\indent gshow t = gshow' @a (constructorInfo \$ datatypeInfo \$ Proxy @a)\\
\indent\ind\ind\ind\ind\ind \$ from t
\vs
}
The functions from \textsf{generics-sop} use a proxy to fix a type value, where we use \textsf{TypeApplications}---a later language extension than those the library relies on. The auxiliary function \texttt{gshow'}, which works on the metadata encoding and generic representation of a datatype, uses \textsf{generics-sop}'s combinators for collapsing and mapping:
\begin{figure}[t]
\centering
\normalsize
\begin{tabular}{l}
\bf Processing constructors
\vs
\tt goConstructor \ann\hs \fa a xs. All (CaseRecShow a) xs\\
\tt\ind\ind\ind\s\s \Ra ConstructorInfo xs \ra NP I xs \ra K String xs\\
\tt goConstructor (Constructor n) args = K \$ unwords (n : args')\\
\tt\s\s \textbf{where}\\
\tt\ind args' \ann [String]\\
\tt\ind args' = hcollapse \$ hcmap (p @a) (K . caseShow @a . unI) args\\
\tt goConstructor (Record n ns) args\\
\tt\ind = K \$ n ++ " \{" ++ intercalate ", " args' ++ "\}"\\
\tt\s\s \textbf{where}\\
\tt\ind args' \ann [String]\\
\tt\ind args' = hcollapse \$ hczipWith (p @a) (goField @a) ns args\\
\tt goConstructor (Infix n \_ \_) (I arg1 :* I arg2 :* Nil)\\
\tt\ind = K \$ caseShow @a arg1 ++ " " ++ n ++ " " ++ caseShow @a arg2
\vs
\tt p \ann Proxy (CaseRecShow a)\\
\tt p = Proxy
\vspace{0.3cm}\\
\bf Processing record fields
\vs
\tt goField \ann\hs \fa a x. (CaseRecShow a) x\\
\tt\ind\ind \Ra FieldInfo x \ra I x \ra K String x\\
\tt goField (FieldInfo n) (I y)\\
\tt\ind = K \$ n ++ " = " ++ caseShow @a y
\end{tabular}
\caption{Auxiliary functions for implementing \texttt{gshow}.}
\label{fig:cons-fields}
\end{figure}
\texttt{
\vs
\indent gshow' \ann\hs \fa a xss. All2 (CaseRecShow a) xss\\
\indent\ind\s\s\s \Ra NP ConstructorInfo xss \ra SOP I xss \ra String\\
\indent gshow' cs (SOP sop)\\
\indent\ind = hcollapse \$ hczipWith allp (goConstructor @a) cs sop\\
\indent\s\s \textbf{where}\\
\indent\ind allp = Proxy @(All (CaseRecShow a))
\vs
}
When a sum structure is actually homogeneous (i.~e., has type \texttt{NS~(K~a)~xs}), it can be collapsed to a single component of type \texttt{a}. The respective instantiation of the function \texttt{hcollapse} that generalizes collapsing homogeneous structures is
\texttt{
\vs
\indent hcollapse  \ann NS  (K a) xs  \ra  a
\vs
}
The \texttt{hczipWith} function is generalized \texttt{zipWith} that operates with a constrained function on heterogeneous structures, where the proxy fixes the constraint:
\texttt{
\vs
\indent hczipWith \ann\hs All c xs\\
\indent\ind\ind\s\s \Ra proxy c \ra (\fa a. c a \Ra f a \ra g a \ra h a)\\
\indent\ind\ind\s\s \ra NP f xs \ra NS g xs \ra NS h xs
\vs
}
In Figure~\ref{fig:cons-fields}, we show functions that process constructors and record fields obtaining their names from proper constructors of metadata and make use of \texttt{caseShow}. The function \texttt{goConstructor} repeats in general the shape of \texttt{gshow'}, where the \texttt{hcmap} function generalizes \texttt{map} for \texttt{NP}, the return type of \texttt{hczipWith} on two products becomes \texttt{NP}, and on two cases, \texttt{hcollapse} collapses the result \texttt{NP} to a list of strings.

\section*{Appendix 2}

Here, we give the full definition of the \texttt{ToFirstCxt'} constraint using definitions from Section~\ref{sec:generic-zipper}.
\texttt{
\vs
\indent type FstCtx n fam xs = Proxy n ': FstRecToHole fam xs
\vs
\indent class ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
\indent \s\s\Ra CtxConsNInj fam a xs n\\
\indent instance ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
\indent \s\s\Ra CtxConsNInj fam a xs n
\vs
\indent class All (ProofF f (FstCtx n fam xs)) yss\\
\indent \s\s\Ra CtxAllProofF f fam yss xs n\\
\indent instance All (ProofF f (FstCtx n fam xs)) yss\\
\indent \s\s\Ra CtxAllProofF f fam yss xs n
\vs
\indent class FromFstRec (FstRecToHole fam xs) xs\\
\indent \s\s\Ra FromFstRec' fam xs\\
\indent instance FromFstRec (FstRecToHole fam xs) xs\\
\indent \s\s\Ra FromFstRec' fam xs
\vs
\indent type family AllNum (c \ann k \ra ConsN \ra Constraint)\\
\indent \ind\ind\ind\ind\s\s\s (xs \ann [k]) (n \ann ConsN) \ann Constraint where\\
\indent \s\s AllNum c '[]\ind\s\s\s n = ()\\
\indent \s\s AllNum c (x ': xs) n = (c x n, AllNum c xs ('N n))
\vs
\indent type ToFirstCtx' fam a n xss\\
\indent \ind = (SListI (CtxCode fam a),\\
\indent \ind\s\s\s All (FromFstRec' fam) xss,\\
\indent \ind\s\s\s AllNum (CtxConsNInj fam a) xss n,\\
\indent \ind\s\s\s AllNum (CtxAllProofF\\
\indent \ind\ind\ind\ind\s (Injection (CtxCode fam a))\\
\indent \ind\ind\ind\ind\s fam (CtxCode fam a)) xss n)
}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{lncs-biblio}
%
\end{document}

### 2018-04-25 lncs-paper.tex
Related work and Conclusion

@@
 \vs
-\indent injections \ann \fa xs f. SListI xs \Ra NP (Injection f xs) xs
-\vs
-\indent newtype Injection f xs a = Inj \{apInj \ann f a \ra K (NS f xs) a\}
+\indent injections \ann \fa xs f. SListI xs \Ra NP (Inj f xs) xs
+\vs
+\indent newtype Inj f xs a = Inj \{apInj \ann f a \ra K (NS f xs) a\}
 \vs
@@
 For \texttt{f} instantiated to \texttt{NP I} and \texttt{xs} to be the sum of products reflecting a representation code, \texttt{injections} creates a product of all constructor choices that inject appropriate constructor arguments into each sum.
+We can use injections to choose the one that matches the obtained index.
 
 We now can witness the choice using the proof of type equality we have defined in Section~\ref{sec:handling-recursion}.
-In the following definition, \texttt{f} is supposed to be instantiated to \texttt{Injection yss} where \texttt{yss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \texttt{ConsN} instead of defining a type class function.
+In the following definition, \texttt{f} is supposed to be instantiated to \texttt{Inj (NP I) xss} where \texttt{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \texttt{ConsN} instead of defining a type class function.
 However, when using a singleton, the recursive definition of \texttt{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
@@
 \vs
-\indent class ConsNInj (n \ann ConsN) (xs \ann [*]) where\\
-\indent\s\s consNInj \ann All (ProofF f xs) yss \Ra NP f yss \ra f xs
+\indent class ConsNInj (n \ann ConsN) (ys \ann [*]) where\\
+\indent\s\s consNInj \ann All (ProofF f ys) xss \Ra NP f xss \ra f ys
 \vs
 \indent instance ConsNInj n xs \Ra ConsNInj ('N n) xs where\\
-\indent\s\s consNInj (\_ :* yss) = consNInj @n yss\\
+\indent\s\s consNInj (\_ :* xss) = consNInj @n xss\\
 \indent\s\s consNInj Nil\ind\s\s\s\hs \hs\ths\tths= impossible\\
 \indent instance ConsNInj 'F xs where\\
-\indent\s\s consNInj ((ys \ann f ys) :* \_)\\
-\indent\ind = fromMaybe impossible \$ witnessEq @(Equal xs ys) ys\\
+\indent\s\s consNInj ((xs \ann f xs) :* \_)\\
+\indent\ind = fromMaybe impossible \$ witnessEq @(Equal xs ys) xs\\
 \indent\s\s consNInj Nil = impossible\\
@@
 }
-As long as the constructor index and product are computed correctly, the proof should never fail (\texttt{impossible}).
-
-Finally, we define a function that applies the injection and returns the constructed context for the given product, representing its constructor arguments, and index of that constructor:
-\texttt{
-\vs
-\indent type AppInj n xs code = (ConsNInj n xs, SListI code,\\
-\indent\ind\ind\ind\ind\ind\ind\s All (ProofF (Injection code) xs) code)
-\vs
-\indent appInjCtx \ann \fa a n xs fam. AppInj n xs (CtxCode fam a)\\
+As long as the constructor index and product type are computed correctly, the proof should never fail (\texttt{impossible}).
+And when it passes, this ensures by type check that the constructor choice for the given product type is faithful.
+
+Finally, we define a function that applies the injection and returns the constructed context for the given product, representing its constructor arguments, and index of that constructor.
+The type of \texttt{injections} below will be inferred from the return type of the context.
+\texttt{
+\vs
+\indent type AppInj n xs ctx =  (ConsNInj n xs, SListI ctx,\\
+\indent\ind\ind\ind\ind\ind\ind All (ProofF (Inj (NP I) ctx) xs) ctx)
+\vs
+\indent appInjCtx \ann \fa n xs fam a. AppInj n xs (CtxCode fam a)\\
 \indent\ind\ind\s\s \Ra NP I xs \ra Context fam a\\
@@
 
-The definition of \texttt{toFirstCtx} can now be given using \texttt{appInjCtx}.
-The following uses a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurence in the given datatype's product list with type \texttt{Hole}.
+The definition of \texttt{toFirstCtx} can now be given using the defined functions \texttt{appInjCtx} and \texttt{fromFstRec}.
+The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurence in the given datatype's product list with type \texttt{Hole}.
 Its definition is straightforward, and is omitted here.
@@
 \indent toFirstCtxNS (Z (np \ann NP I xs))\\
-\indent\s\s = appInjCtx @a @(CtxConsN (CtxCode fam a) n) \$\\
+\indent\s\s = appInjCtx @(CtxConsN (CtxCode fam a) n) \$\\
 \indent\ind\ind I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
@@
 
-We have established the mechanism of translation, which is cumbersome but verifying, via the proof, that the constructed context will be correct for any given datatype and family.
+We have established the mechanism of translation, which is cumbersome but verifying, via the proof, that the constructed context will have the correct type for any given datatype and family.
 The other functions in the zipper interface can be implemented according to this technique.
@@
 
-Related work
+Over the years, there has been a lot of works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated.
+Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approachs and provide their detailed comparison in different aspects.
+There are several generic views that use forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}.
+And there are a number of approachs that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation.
+The SOP view~\cite{VriLoeh2014} which we use to demonstrate our technique is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype.
+This approach uses heterogeneous lists of types to encode sums and products in the generic representation.
+
+The idea similar to SOP has been proposed by Kiselyov~et~al.~\cite{Kiselyov2004} in their \textsf{HList} library for strongly typed heterogeneous collections.
+In their paper, they also discuss problems concerened with overlap which they use for access operations.
+They restrict overlap by introducing a type class using a functional dependency, which provides functionality of type equality that is similar to our solution.
+Morris and Jones~\cite{Morris2010} introduce the type-class system \textsf{ilab}, based on the Haskell 98 class system, with a new feature, called ``instance chains'', that allows to control overlap by using an explicit syntax in instance declarations, which resembles if-else chains.
+But use of instance chains and local use of overlap leave code error-prone as a consequence of type class oppennes.
+Closed type families~\cite{Eisenberg2014} were recently introduced in Haskell to solve the overlap problem.
+
+Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001}, mutually recursive~\cite{MuRec2009}, and heterogeneous~\cite{Adams2010} types using fixed-point generic views.
 
@@
 \label{sec:conclusion}
+
+It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approachs to generic programming was considered as unsuitable for this task.
+There are some approachs that address this problem by means of local overlaps, where it needs to access the points of recursion.
+We have developed the technique that allows us to define generic functions that treat recursion without its explicit encoding and overlap.
+This makes use of closed type families that are the recent Haskell extension.
+
+We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface, and supports families of mutually recursive datatypes.
+Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approachs is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.
 
@@
 
-The \texttt{gshow} function makes use of metadata provided by the \texttt{generics-sop} library through the class \texttt{HasDatatypeInfo}.
+The \texttt{gshow} function makes use of metadata provided by the \textsf{generics-sop} library through the class \texttt{HasDatatypeInfo}.
 That class delivers generic functions for handling meta-information which \textsf{generics-sop} defines separate from the main generic representation of a datatype.
@@
 \indent gshow \ann \fa a. GShow a \Ra a \ra String\\
-\ind\ind\s\Ra a \ra String\\
 \indent gshow t = gshow' @a (constructorInfo \$ datatypeInfo \$ Proxy @a)\\
@@
 \indent \ind\s\s\s AllNum (CtxAllProofF\\
-\indent \ind\ind\ind\ind\s (Injection (CtxCode fam a))\\
+\indent \ind\ind\ind\ind\s (Inj (NP I) (CtxCode fam a))\\
 \indent \ind\ind\ind\ind\s fam (CtxCode fam a)) xss n)

### 2018-04-25 lncs-paper.tex
Fix an error in RW

@@
 
-Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001}, mutually recursive~\cite{MuRec2009}, and heterogeneous~\cite{Adams2010} types using fixed-point generic views.
+Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001} and mutually recursive~\cite{MuRec2009} types using fixed-point generic views.
+Adams~\cite{Adams2010} defines a generic zipper for heterogeneous types.
 

### 2018-04-25 lncs-paper.tex
Intro

@@
 %
-\author{First Author\inst{1} \and
-Second Author\inst{2}}
+\author{Anonymous%First Author\inst{1} \and
+%Second Author\inst{2}
+}
 %
@@
 %
-\institute{Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany\\
-\email{lncs@springer.com} \and
-ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-\email{abc@uni-heidelberg.de}}
+%~ \institute{Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany\\
+%~ \email{lncs@springer.com} \and
+%~ ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
+%~ \email{abc@uni-heidelberg.de}}
 %
@@
 \begin{abstract}
-Many of the extensively used today libraries for datatype-generic programming offer a fixed-point view on datatypes to express the recursive structure.
-Some approachs based on sums of products, however, do not use a fixed point.
-Those views therefore allow for generic functions that do not require to look at the recursive knots in the datatype representation, but raise issues when it needs to deal with recursion.
-A known unwelcome solution is use of overlapping instances.
-We present a technique that uses closed type families to allow us to eliminate overlap for handling recursion.
+Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express a recursive structure.
+Some approaches based on sums of products, however, do not use a fixed point.
+Those views therefore allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues when it needs to deal with recursion.
+A known and unwelcome solution is the use of overlapping instances.
+We present an approach that uses closed type families to eliminate the need of overlap for handling recursion.
 Moreover, we show that our idiom allows for families of mutually recursive datatypes.
 
-The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are expressed by separate binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view.
+%The \textsf{generics-sop} library is an approach to representing data using $n$-ary sums and products that both are list-like structures, which is different from the classical view where datatypes are expressed by separate binary sums and products, and it does not encode recursive knots explicitly as with the fixed-point view.
 We choose this approach as a case study to demonstrate our solution.
 
-\keywords{Datatype-generic programming \and Sums of products \and Recursion \and Overlapping instances \and Closed type families \and Zippers \and Mutually recursive datatypes \and Haskell.}
+\keywords{Datatype-generic programming \and Sums of products \and Recursion \and Overlapping instances \and Closed type families \and Zipper \and Mutually recursive datatypes \and Haskell.}
 \end{abstract}
@@
 
-A classical way to generically view data is to represent constructors as chains of nested binary sums, and constructor arguments as chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
+A classical way to generically describe a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
 De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types.
-They call their view SOP that stands for a ``sum of products''---this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds} and \emph{kind polymorphism}~\cite{Yorgey2012}, \emph{constraint kinds}, and \emph{GADTs}~\cite{Schrijvers2009}.
+They call their view SOP which stands for a ``sum of products''~--- this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
+%, and \emph{GADTs}~\cite{Schrijvers2009}. -- very old extension, 2003 approx.
 Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.
@@
 Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}, more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}.
-For handling recursion, several generic programming approachs express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}\nocite{MagLoeh2012}.
-The SOP view allows to easily define functions that do not need to specially process recursive occurences.
-But for those that do need it, it normally does not allow.
-
-One possible solution is to construct a more specific universe modifying the SOP core by explicitly encoding recursive positions using a fixed point.
-The fixed-point approach can be combined with SOP~\cite{VriLoeh2014}, but doing so makes that more complicated.
-Besides, such a decision may lead to need for additional conversions between the views.
-Our method makes it possible to define functions that need access to recursive occurences within the SOP view without modifying it anyhow.
-However, the definitions are still quite verbose.
-
-Another known solution is use of overlapping instances.
-This is unwelcome, as involving overlapping instances complicates reasoning about the code semantics, makes that unstable because it can affect the program behavior, if one adds a module with more specific instances.
-Morris and Jones~\cite{Morris2010} extensively discuss problems that may arise from using overlapping instances, and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap, while type classes in \textsf{ilab} are open.
-But openness leaves a possibility for a rise of troubles---thus, defining instances in different modules still can cause issues.
-The overlap problem is also crucial in the security setting when code is compiled as safe, as GHC does not capture unsafe overlaps by detecting the presence of overlapping declarations and marking the module unsafe~\cite{sh-overlapping}.
-Hence, the undesirable effect of overlapping instances becomes a problem in the setting of Safe Haskell.
-Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
+For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}.
+The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise it yields.
+
+One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed point approach.
+This may complicate the whole framework significantly.
+Besides, such a decision may lead to extra conversions between the generic views.
+
+Another known solution uses overlapping instances.
+This usually unwelcome Haskell extension complicates reasoning about the semantics of code.
+In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances.
+Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances, and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap.
+The overlap problem also strikes in the security setting when a code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
+%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
 Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
-
-We believe that our idea is suitable for a number of different sum-of-products approachs that do not exploit a fixed point---so there the described problem appears.
-We choose the \textsf{generics-sop} approach as a case study because we think it is a widely applicable library based on the interesting ideas for generic programming, and it relies on GHC's recent extensions.
-
-To demonstrate our solution, we give several examples including the generic zipper for families of mutually recursive datatypes.
-The approach we introduce has also advantages that contribute to \textsf{generics-sop} in addition to the main result of our work.
-There are examples of helpful generic functions defined in \textsf{basic-sop}~\cite{basic-sop} using the SOP view.
-Those definitions, however, suffer from that they require manual (boilerplate) instance declarations---as a consequence of that they do not take recursion points in a data structure into account, as we will show.
-We surmount this inconvenience in an interface that we provide for generic functions: the type class instances with them do not need to be manually declared.
-The zipper definition we offer is also very flexible for use with different families of datatypes in comparison with known generic solutions for a zipper, as it just takes a family in the form of a list of its constituent types when starts traversing a structure, and needs no boilerplate to be written for a family in addition.
-
-\paragraph{Contributions} This paper makes the following contributions:
+% -- this should go to Realted Work (if stays in the paper at all)
+
+We feel both existing approaches unsatisfactory and make the following contributions:
 \begin{itemize}
-\item We introduce an idiom that allows to exclude use of overlapping instances for handling recursion in a number of sum-of-products approachs to datatype-generic programming that do not express recursive positions through a fixed point.
-\item We present our idea by giving several examples of generic functions, particularly the generic zipper for mutually recursive datatypes, using the \textsf{generics-sop} view.
-\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
-\item We develop the generic zipper that can be simply and flexibly used with different families of mutually recursive datatypes.
+\item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
+\item We introduce an idiom that overcomes the problem.
+The approach avoids both, the use of overlapping instances and changing a generic representation.
+The presentation goes through several examples.
+(Section~\ref{sec:handling-recursion})
+% -- 
+%\item We provide a convenient interface for generic functions: there is no need to manually declare type class instances for using them.
+% -- this doesn't sound like a contribution
+\item We evaluate our approach through development of a larger-scale use case — the generic zipper.
+The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
+\item We note, that our the approach can contribute to \textsf{generics-sop} eliminating boilerplate instance declarations which come as a consequence of absence of information about recursion points~(Section~\ref{sec:gshow}).
 \end{itemize}
 
-\paragraph{Organization} The paper is organized as follows.
+We believe that our idea is suitable for any sum-of-products approach that do not exploit a fixed point view and thus subject to the problem.
+We choose the \textsf{generics-sop} approach as a case study because it appears to be a widely applicable library and builds on powerful language extensions implemented in GHC.
+
+%\paragraph{Organization} The paper is organized as follows.
 In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and demonstrate the problem by a short example.
@@
 In Section~\ref{sec:related-work}, we review related work, and we conclude in Section~\ref{sec:conclusion}.
+%
+% Organization part is not used these days (cf. any of the later papers by SPJ)
+% Instead, the Sec.-references are integrated into the Contributions part.
 

### 2018-04-25 lncs-paper.tex

@@
 In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances.
-Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances, and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap.
+Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap.
 The overlap problem also strikes in the security setting when a code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
@@
 % -- this doesn't sound like a contribution
-\item We evaluate our approach through development of a larger-scale use case — the generic zipper.
+\item We evaluate our approach through development of a larger-scale use case~--- the generic zipper.
 The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).

### 2018-04-25 lncs-paper.tex
polishing contributions

--- 2018-04-25 lncs-paper.tex
@@
 In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances.
-Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances, and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap.
+Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
+% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
 The overlap problem also strikes in the security setting when a code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
@@
 \item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
-\item We introduce an idiom that overcomes the problem.
-The approach avoids both, the use of overlapping instances and changing a generic representation.
-The presentation goes through several examples.
-(Section~\ref{sec:handling-recursion})
+\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation. (Section~\ref{sec:handling-recursion})
 % -- 
@@
 % -- this doesn't sound like a contribution
-\item We evaluate our approach through development of a larger-scale use case — the generic zipper.
+\item We evaluate our approach through the development of a larger-scale use case---the generic zipper.
 The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
-\item We note, that our the approach can contribute to \textsf{generics-sop} eliminating boilerplate instance declarations which come as a consequence of absence of information about recursion points~(Section~\ref{sec:gshow}).
+\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations which necessarily  arise in practice as a consequence of absence of information about recursion points.
+An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{sec:gshow}.
 \end{itemize}

### 2018-04-26 lncs-paper.tex
Remove appendices

@@
 %
-\author{Anonymous%First Author\inst{1} \and
-%Second Author\inst{2}
+\author{Anna Bolotina\inst{1} \and
+Artem Pelenitsyn\inst{2}
 }
@@
 %
-%~ \institute{Springer Heidelberg, Tiergartenstr. 17, 69121 Heidelberg, Germany\\
-%~ \email{lncs@springer.com} \and
-%~ ABC Institute, Rupert-Karls-University Heidelberg, Heidelberg, Germany\\
-%~ \email{abc@uni-heidelberg.de}}
+\institute{Southern Federal University, Rostov-on-Don, Russia\\
+\email{bolotina@sfedu.ru} \and
+Czech Technical University in Prague, Prague, Czech Republic\\
+\email{pelenart@fit.cvut.cz}}
 %
@@
 
-A classical way to generically describe a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
+A classical way to generically express a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
 De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types.
-They call their view SOP which stands for a ``sum of products''~--- this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
+They call their view SOP which stands for a ``sum of products''---this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
 %, and \emph{GADTs}~\cite{Schrijvers2009}. -- very old extension, 2003 approx.
@@
 
-One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed point approach.
+One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed-point approach.
 This may complicate the whole framework significantly.
@@
 % and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
-The overlap problem also strikes in the security setting when a code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
+The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
 %Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
@@
 The function is constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types.
-The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in Appendix~1.
+The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
 
@@
 The definition of \texttt{toFirstCtx} can now be given using the defined functions \texttt{appInjCtx} and \texttt{fromFstRec}.
-The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurence in the given datatype's product list with type \texttt{Hole}.
+The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurrence in the given datatype's product list with type \texttt{Hole}.
 Its definition is straightforward, and is omitted here.
@@
 The \texttt{ToFirstCtx'} constraint here is a complex constraint for use of \texttt{appInjCtx} and \texttt{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints.
-Its definition is shown in Appendix~2.
+Its definition is given in the extended version of the paper in Technical Report.
 We omit this here for reasons of space.
@@
 Over the years, there has been a lot of works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated.
-Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approachs and provide their detailed comparison in different aspects.
+Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in different aspects.
 There are several generic views that use forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}.
-And there are a number of approachs that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation.
+And there are a number of approaches that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation.
 The SOP view~\cite{VriLoeh2014} which we use to demonstrate our technique is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype.
@@
 
-It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approachs to generic programming was considered as unsuitable for this task.
-There are some approachs that address this problem by means of local overlaps, where it needs to access the points of recursion.
+It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approaches to generic programming was considered as unsuitable for this task.
+There are some approaches that address this problem by means of local overlaps, where it needs to access the points of recursion.
 We have developed the technique that allows us to define generic functions that treat recursion without its explicit encoding and overlap.
@@
 We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface, and supports families of mutually recursive datatypes.
-Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approachs is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.
-
-\section*{Appendix 1}
-
-In this appendix, we show the implementation of the generic function \texttt{gshow}, which uses \texttt{caseShow}, described in Section~\ref{sec:gshow}.
-The \texttt{gshow} function is originally defined by De Vries and L\"{o}h in the \textsf{basic-sop} package~\cite{basic-sop}.
-Their implementation has the issue that for recursive datatypes whose recursive nodes should be displayed within parentheses, the function acts incorrectly due to the SOP view's limitations on expressing recursion.
-We improve the implementation from \textsf{basic-sop} with our approach to handling recursion.
-
-The \texttt{gshow} function makes use of metadata provided by the \textsf{generics-sop} library through the class \texttt{HasDatatypeInfo}.
-That class delivers generic functions for handling meta-information which \textsf{generics-sop} defines separate from the main generic representation of a datatype.
-The definition of \texttt{gshow} using those functions is as follows:
-\texttt{
-\vs
-\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
-\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
-\vs
-\indent gshow \ann \fa a. GShow a \Ra a \ra String\\
-\indent gshow t = gshow' @a (constructorInfo \$ datatypeInfo \$ Proxy @a)\\
-\indent\ind\ind\ind\ind\ind \$ from t
-\vs
-}
-The functions from \textsf{generics-sop} use a proxy to fix a type value, where we use \textsf{TypeApplications}---a later language extension than those the library relies on.
-The auxiliary function \texttt{gshow'}, which works on the metadata encoding and generic representation of a datatype, uses \textsf{generics-sop}'s combinators for collapsing and mapping:
-\begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\bf Processing constructors
-\vs
-\tt goConstructor \ann\hs \fa a xs. All (CaseRecShow a) xs\\
-\tt\ind\ind\ind\s\s \Ra ConstructorInfo xs \ra NP I xs \ra K String xs\\
-\tt goConstructor (Constructor n) args = K \$ unwords (n : args')\\
-\tt\s\s \textbf{where}\\
-\tt\ind args' \ann [String]\\
-\tt\ind args' = hcollapse \$ hcmap (p @a) (K . caseShow @a . unI) args\\
-\tt goConstructor (Record n ns) args\\
-\tt\ind = K \$ n ++ " \{" ++ intercalate ", " args' ++ "\}"\\
-\tt\s\s \textbf{where}\\
-\tt\ind args' \ann [String]\\
-\tt\ind args' = hcollapse \$ hczipWith (p @a) (goField @a) ns args\\
-\tt goConstructor (Infix n \_ \_) (I arg1 :* I arg2 :* Nil)\\
-\tt\ind = K \$ caseShow @a arg1 ++ " " ++ n ++ " " ++ caseShow @a arg2
-\vs
-\tt p \ann Proxy (CaseRecShow a)\\
-\tt p = Proxy
-\vspace{0.3cm}\\
-\bf Processing record fields
-\vs
-\tt goField \ann\hs \fa a x. (CaseRecShow a) x\\
-\tt\ind\ind \Ra FieldInfo x \ra I x \ra K String x\\
-\tt goField (FieldInfo n) (I y)\\
-\tt\ind = K \$ n ++ " = " ++ caseShow @a y
-\end{tabular}
-\caption{Auxiliary functions for implementing \texttt{gshow}.}
-\label{fig:cons-fields}
-\end{figure}
-\texttt{
-\vs
-\indent gshow' \ann\hs \fa a xss. All2 (CaseRecShow a) xss\\
-\indent\ind\s\s\s \Ra NP ConstructorInfo xss \ra SOP I xss \ra String\\
-\indent gshow' cs (SOP sop)\\
-\indent\ind = hcollapse \$ hczipWith allp (goConstructor @a) cs sop\\
-\indent\s\s \textbf{where}\\
-\indent\ind allp = Proxy @(All (CaseRecShow a))
-\vs
-}
-When a sum structure is actually homogeneous (i.~e., has type \texttt{NS~(K~a)~xs}), it can be collapsed to a single component of type \texttt{a}.
-The respective instantiation of the function \texttt{hcollapse} that generalizes collapsing homogeneous structures is
-\texttt{
-\vs
-\indent hcollapse  \ann NS  (K a) xs  \ra  a
-\vs
-}
-The \texttt{hczipWith} function is generalized \texttt{zipWith} that operates with a constrained function on heterogeneous structures, where the proxy fixes the constraint:
-\texttt{
-\vs
-\indent hczipWith \ann\hs All c xs\\
-\indent\ind\ind\s\s \Ra proxy c \ra (\fa a. c a \Ra f a \ra g a \ra h a)\\
-\indent\ind\ind\s\s \ra NP f xs \ra NS g xs \ra NS h xs
-\vs
-}
-In Figure~\ref{fig:cons-fields}, we show functions that process constructors and record fields obtaining their names from proper constructors of metadata and make use of \texttt{caseShow}.
-The function \texttt{goConstructor} repeats in general the shape of \texttt{gshow'}, where the \texttt{hcmap} function generalizes \texttt{map} for \texttt{NP}, the return type of \texttt{hczipWith} on two products becomes \texttt{NP}, and on two cases, \texttt{hcollapse} collapses the result \texttt{NP} to a list of strings.
-
-\section*{Appendix 2}
-
-Here, we give the full definition of the \texttt{ToFirstCxt'} constraint using definitions from Section~\ref{sec:generic-zipper}.
-\texttt{
-\vs
-\indent type FstCtx n fam xs = Proxy n ': FstRecToHole fam xs
-\vs
-\indent class ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
-\indent \s\s\Ra CtxConsNInj fam a xs n\\
-\indent instance ConsNInj (CtxConsN (CtxCode fam a) n) (FstCtx n fam xs)\\
-\indent \s\s\Ra CtxConsNInj fam a xs n
-\vs
-\indent class All (ProofF f (FstCtx n fam xs)) yss\\
-\indent \s\s\Ra CtxAllProofF f fam yss xs n\\
-\indent instance All (ProofF f (FstCtx n fam xs)) yss\\
-\indent \s\s\Ra CtxAllProofF f fam yss xs n
-\vs
-\indent class FromFstRec (FstRecToHole fam xs) xs\\
-\indent \s\s\Ra FromFstRec' fam xs\\
-\indent instance FromFstRec (FstRecToHole fam xs) xs\\
-\indent \s\s\Ra FromFstRec' fam xs
-\vs
-\indent type family AllNum (c \ann k \ra ConsN \ra Constraint)\\
-\indent \ind\ind\ind\ind\s\s\s (xs \ann [k]) (n \ann ConsN) \ann Constraint where\\
-\indent \s\s AllNum c '[]\ind\s\s\s n = ()\\
-\indent \s\s AllNum c (x ': xs) n = (c x n, AllNum c xs ('N n))
-\vs
-\indent type ToFirstCtx' fam a n xss\\
-\indent \ind = (SListI (CtxCode fam a),\\
-\indent \ind\s\s\s All (FromFstRec' fam) xss,\\
-\indent \ind\s\s\s AllNum (CtxConsNInj fam a) xss n,\\
-\indent \ind\s\s\s AllNum (CtxAllProofF\\
-\indent \ind\ind\ind\ind\s (Inj (NP I) (CtxCode fam a))\\
-\indent \ind\ind\ind\ind\s fam (CtxCode fam a)) xss n)
-}
+Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approaches is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.
 

### 2018-04-26 lncs-paper.tex
Remove appendices

@@
 We first explain the terminology we adopt from SOP~\cite{VriLoeh2014,Loeh2015} and use throughout the paper.
-The main idea of the SOP view is to use $n$-ary sums and products to view a datatype as its isomorphic code whose kind is a list of lists of types.
-Using the \textsf{DataKinds} extension that enables datatype definitions to be promoted to kinds, SOP expresses this code with a type family:
+The main idea of the SOP view is to use $n$-ary sums and products to represent a datatype as an isomorphic code whose kind is a list of lists of types.
+The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
 \texttt{
@@
 }
-An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments depending on the chosen constructor; the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
+An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
 
-For example, consider a datatype of binary trees:
+Consider, for instance, a datatype of binary trees:
 \texttt{
@@
 
-A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms.
+%A type of promoted lists has no inhabitants, so the universe provides datatypes to operate on $n$-ary sums and products as on terms.
+% -- this sentence falls out of the flow of the paragraph
+
 As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types.
-The universe defines these with embedded functor application, where the functor is some type constructor.
-Thus, each component of \texttt{NS} and \texttt{NP} corresponds to one type (its index) in the type-level list and has type \texttt{f x}, for an index type \texttt{x}.
-The definitions of \texttt{NS} and \texttt{NP} both are kind polymorphic.
+The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the code.
+Therefore, \texttt{NP} is a tiny abstraction of a usual definition of a heterogeneous list The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic.
 The index list allows to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.
 
-As basic instantiation types of \texttt{f}, SOP defines an identity functor \texttt{I} that is type-level equivalent for \texttt{id}, and a constant functor \texttt{K}, for \texttt{const}:
+Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analog of \texttt{const}:
 \texttt{
@@
 }
-Whereas instantiated with \texttt{I}, \texttt{NP} is a direct heterogeneous list of types, with \texttt{K a}, it is essentially homogeneous.
-An example value of an \texttt{NP} for \texttt{I} looks thus:
+
+If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list of types, while \texttt{K a} turns it into homogeneous one. Here is an example value of \texttt{NP I}:
 \texttt{
@@
 
-A clear idea underlies the sum definition.
-The constructor \texttt{S} of a sum skips the first element of an $n$-element index list giving an index in a list that has $n+1$ elements, while \texttt{Z} stores the payload of type \texttt{f x}.
+We turn to the sum definition now.
+The constructor \texttt{S} of \texttt{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \texttt{Z} stores the payload of type \texttt{f x}.
 For example, the following chooses the third element of a sum:
@@
 }
-The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and gives the product representing arguments of that constructor.
+The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.
 
-With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
+With the \texttt{NS} and \texttt{NP} machinery at hand, SOP defines the \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} witnessing the isomorphism between a datatype and its generic representation:
 \texttt{
@@
 }
-The sum of products \texttt{SOP f} is a newtype for \texttt{NS (NP f)}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code.
+The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code.
 The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.
 
-The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be satisfied, and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the constraint \texttt{All} specifying that a particular constraint holds for each member of a list of types\footnote{The definition of \texttt{All} is also kind polymorphic.
-Here, it applies the constraint to each inner list of types in the the outer list of lists.}.
-Using constraints as types of a special kind \texttt{Constraint}, hence they can appear as type parameters, is allowed by the \textsf{ConstraintKinds} language extension.
+We are leaving out discussion of the \texttt{All SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work.
+Although, we do use \texttt{All} constraint combinator in the following.
+Therefore, it is worth noting that \texttt{All} applies particular constraint (e.g.
+\texttt{SListI} above) to each member of a list of types.
+The usage of constraints as type arguments allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.
 
-The \texttt{Generic} class definition in the \textsf{generics-sop} library allows to create the instance for a particular datatype automatically using internal \textsf{GHC.Generics}.
+We have introduced generic representation employed by the SOP library and are ready to describe the problem of handling recursion points stemming from the representation.
 
@@
 
-In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are now going to improve the solution and remove overlap.
+In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are going to improve the solution and remove overlap now.
 
 Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}.
-Equations for the \emph{closed type family} are disallowed to de defined outside its declaration.
-Under this extension, we can give the following definition of type-level equality:
+The main idea of the extension being that he equations for a \emph{closed type family} are disallowed outside its declaration.
+Under the extension, we can give the following definition of a type-level equality:
 \texttt{
@@
 }
-The equations in a closed type family are matched in a top-to-bottom order.
-Since the order is always strictly defined, overlapping equations can be used to define sound type-level functions like this type equality.
+The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.
 
-We now return to our running example from Section~\ref{sec:recursion-problem}. With type equality, we can witness the coersion between the equal types by defining a type class:
+\subsection{Solution to \texttt{subterms} revised}
+
+We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between the equal types by defining a type class:
 \texttt{
@@
 }
-We then need to provide the proof for all elements in \texttt{NP}.
-This can be done by means of the \texttt{All} combinator.
-For using that, we need the constraint can be partially applied, and so use a synonym for \texttt{Proof (Equal a b) a b} that we define as a class with a superclass constraint, rather than as a type synonym:
+
+For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
+This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq} which abbreviates heavy-weighted interface of \texttt{Proof}:
 \texttt{
@@
 }
-The following implementation of \texttt{subtermsNP} using this proof resembles our first definition given in the previous section:
+Resulting implementation of \texttt{subtermsNP} resembles our first definition given in the previous section:
 \texttt{
@@
 }
-We also make use of \textsf{ScopedTypeVariables} in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
-A proper \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
 
-All generic functions accessing recursive knots in the underlying datatype structure can be defined in this way.
-We give another detailed example of one of such functions to show how to adapt our idiom to different scenarios.
+As a side note, we make use of \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
+
+To complete the solution of the problem, the \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
+
+In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task.
+We give another example showing how to adapt our idiom to different scenarios in the following section.
 

### 2018-04-26 lncs-paper.tex
first half of sec. 3

--- 2018-04-25 lncs-paper.tex
@@
 }
-The equations in a closed type family are matched in a top-to-bottom order.
-Since the order is always strictly defined, overlapping equations can be used to define sound type-level functions like this type equality.
-
-We now return to our running example from Section~\ref{sec:recursion-problem}. With type equality, we can witness the coersion between the equal types by defining a type class:
+The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.
+
+\subsection{Solution to \texttt{subterms} revised}
+
+We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between the equal types by defining a type class:
 \texttt{
@@
 }
-We also make use of \textsf{ScopedTypeVariables} in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
-A proper \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
-
-All generic functions accessing recursive knots in the underlying datatype structure can be defined in this way.
-We give another detailed example of one of such functions to show how to adapt our idiom to different scenarios.
+
+As a side note, we make use of \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
+
+To complete the solution of the problem, the \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
+
+In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task.
+We give another example showing how to adapt our idiom to different scenarios in the following section.
 

### 2018-04-26 lncs-paper.tex
Polish Sec. 3

@@
 Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}.
-The main idea of the extension being that he equations for a \emph{closed type family} are disallowed outside its declaration.
-Under the extension, we can give the following definition of a type-level equality:
+The main idea of the extension being that the equations for a \emph{closed type family} are disallowed outside its declaration.
+Under the extension, we can give the following definition of type-level equality:
 \texttt{
@@
 In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task.
-We give another example showing how to adapt our idiom to different scenarios in the following section.
+We give another example showing how to adapt our idiom to different scenarios in the following subsection.
 

### 2018-04-26 lncs-paper.tex
2.1

--- 2018-04-26 lncs-paper.tex
@@
 }
-An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments depending on the chosen constructor; the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
-
-For example, consider a datatype of binary trees:
+An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
+
+Consider, for instance, a datatype of binary trees:
 \texttt{
@@
 }
-The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and gives the product representing arguments of that constructor.
-
-With the representation datatypes \texttt{NS} and \texttt{NP}, SOP defines a \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} to witness the isomorphism between a datatype and its generic representation:
+The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.
+
+With the \texttt{NS} and \texttt{NP} machinery at hand, SOP defines the \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} witnessing the isomorphism between a datatype and its generic representation:
 \texttt{
@@
 }
-The sum of products \texttt{SOP f} is a newtype for \texttt{NS (NP f)}, indexed by a type-level list of lists, and the structural representation \texttt{Rep} of a datatype is a type synonym for a \texttt{SOP I} of its code.
+The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code.
 The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.
 
-The \texttt{All SListI} constraint in the \texttt{Generic} class definition is supposed to be satisfied, and an understanding of this detail is not necessary to understand the paper, but we will next use an understanding of the constraint \texttt{All} specifying that a particular constraint holds for each member of a list of types\footnote{The definition of \texttt{All} is also kind polymorphic.
-Here, it applies the constraint to each inner list of types in the the outer list of lists.}.
-Using constraints as types of a special kind \texttt{Constraint}, hence they can appear as type parameters, is allowed by the \textsf{ConstraintKinds} language extension.
-
-The \texttt{Generic} class definition in the \textsf{generics-sop} library allows to create the instance for a particular datatype automatically using internal \textsf{GHC.Generics}.
+We are leaving out discussion of the \texttt{All SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work.
+Although, we do use \texttt{All} constraint combinator in the following.
+Therefore, it is worth noting that \texttt{All} applies particular constraint (e.g.
+\texttt{SListI} above) to each member of a list of types.
+The usage of constraints as type arguments allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.
+
+We have introduced generic representation employed by the SOP library and are ready to describe the problem of handling recursion points stemming from the representation.
 

### 2018-04-26 lncs-paper.tex
Polish Sec. 2.1

@@
 As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types.
-The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the code.
+The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the index list.
 Therefore, \texttt{NP} is a tiny abstraction of a usual definition of a heterogeneous list The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic.
@@
 
-Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analog of \texttt{const}:
+Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analogue of \texttt{const}:
 \texttt{
@@
 \texttt{SListI} above) to each member of a list of types.
-The usage of constraints as type arguments allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.
+The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.
 

### 2018-04-26 lncs-paper.tex

@@
 
-Let us illustrate the problem by giving a short example.
-The \textsf{QuickCheck}~\cite{Claessen2011}, a library for automatic testing of Haskell code, using the \textsf{GHC.Generics} view, defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
-To implement such a function using the SOP view, we have to say something like this:
+We illustrate the problem through a short example.
+The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
+We reimplement this function using the SOP view:
 \texttt{
@@
 
-The definition of \texttt{subtermsNP} shows the clear idea: it traverses the product adding the element to the result list if its type is the same as a given term's, otherwise skipping that.
-We use GHC's \textsf{TypeAppications} extension, which allows \emph{visible type applications}, here to fix that type.
-Note that using this with the type variables appearing in the function signature requires explicit universal quantification.
+The algorithm of \texttt{subtermsNP} is straightforward---it traverses the product appending current element to the result list if its type is the same as the given term's, otherwise skipping the element.
+We use GHC's \textsf{TypeAppications} extension to pass that type.
 
-Now, we need a way to check type equality, and in the case of equal types, to witness that the element is of the appropriate type that admits adding that to the result list of subterms.
-There is a solution: we can follow \textsf{QuickCheck}'s example and implement this using overlapping instances as it does with \textsf{GHC.Generics}.
-
-The definition of \texttt{subtermsNP} we have shown serves as a clear illustration of the idea and as a template for a more satisfactory solution that we present in Section~\ref{sec:handling-recursion}.
-But right now, the simplest way to implement this is to rewrite \texttt{subtermsNP} as follows:
+Now, we need a way to check type equality and, in the case of equal types, to witness that the element is of the desired type.
+There is no clear path to this at the moment.
+Therefore, we step back and follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.} using overlapping instances of a dedicated class instead.
 \texttt{
@@
 }
-This uses a type class with the overloaded function.
-It needs also to declare that all the products in the datatype code are instances of \texttt{Subterms a} by adding a constraint to the \texttt{subterms} and \texttt{subtermsNS} signatures:
+
+To make the whole solution work we need to propagate the constraints all the way through \texttt{subtermsNS} and \texttt{subterms} signatures:
 \texttt{
@@
 
-Though this will work, and there are a number of the Hackage packages that do it in such a manner, we want to write programs free of overlap because of complexity it introduces into code as we have discussed in Section~\ref{sec:introduction}.
-Fortunately, we can do without overlap, and we will demonstrate the solution in the next section.
+Although, the approach works as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
+That would avoid the complexity ovehead introduced by the approach, as we have noted in the introduction.
 

### 2018-04-26 lncs-paper.tex

@@
 
-The function \texttt{show} is one of common examples of useful functions that traverse a datatype's recursive structure.
+The function \texttt{show} is a common example of useful functions that traverse a datatype's recursive structure.
 It is known that this function can be defined in a generic way for an arbitrary datatype.
-De Vries and L\"{o}h define the generic function \texttt{gshow} in the \textsf{basic-sop}~\cite{basic-sop} package based on the SOP view.
+De Vries and L\"{o}h define generic function \texttt{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view.
 We follow their implementation of \texttt{gshow} for the most part, but improve it in respect of handling recursion.
-The original function \texttt{gshow} from \textsf{basic-sop} is not a substitute for the version of \texttt{show} that can be generated for a particular datatype through \texttt{\textbf{deriving} Show}, because it does not consider recursion to place parentheses at points of recursive calls---but we remove this lack of expressiveness.
+The original \texttt{gshow} yields to the standard \texttt{show} generated through \texttt{\textbf{deriving} Show}, because it does not consult with recursion points to place parentheses.
+We eliminate this drawback.
 
@@
 As before, we consider two cases.
-In the first case when the position we are matching is not recursive, we only require it to be the \texttt{Show} instance, and invoke its \texttt{show}, whereas in the case of the recursion point, we surround it with parentheses and apply our generic function \texttt{gshow}.
-We again use the defined type equality to model a form of pattern matching on types:
+In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function.
+Whereas in the case of recursion point, we surround it with parentheses and apply our generic function \texttt{gshow}.
+Thus, the type family for equality model a form of pattern matching on the types again:
 \texttt{
@@
 }
-We provide the synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, in the same way as before, as well as the synonym for the matching function:
+We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise the synonym for the matching function:
 \texttt{
@@
 
-The function \texttt{gshow} involves meta-information provided by \textsf{generics-sop}'s class \texttt{HasDatatypeInfo} to show a datatype constructor's and its record fields' names.
-The \textsf{generics-sop} library has features for deriving this meta-information automatically.
+The resulting function \texttt{gshow} is a subject of a number of constraints abbreviated by \texttt{GShow} synonym:
 \texttt{
@@
 }
+The function \texttt{gshow} employs metadata provided by \textsf{generics-sop}'s class \texttt{Has\-Datatype\-Info} to show the names of a datatype constructor and its record fields.
+The \textsf{generics-sop} library is able to derive this metadata automatically.
+The function is also constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types.
+
 We define \texttt{gshow} mutually recursive with \texttt{caseShow}.
-The function is constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types.
-The full implemenentation of the function \texttt{gshow} using \texttt{caseShow} is given in the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
+The full implemenentation of the function \texttt{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
 
-The function \texttt{gshow} can now be used to generically show data, for example, a value of type \texttt{Tree~Bool}, where \texttt{Tree~a} (Section~\ref{sec:sop-view}) is now assumed to be an instance of \texttt{Generic}, and of \texttt{HasDatatypeInfo}:
+The function \texttt{gshow} can now be used to generically show data---for example, a value of type \texttt{Tree~Bool}; note that \texttt{Tree~a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \texttt{Generic} and \texttt{HasDatatypeInfo}.
 \texttt{
@@
 }
-And here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
+Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
 \texttt{
@@
 }
-This is a consequence of that in \textsf{basic-sop}, the \texttt{gshow} function does not treat recursive positions separately, and therefore requires the \texttt{Show} constraint for all knots in the datatype structure.
+This is a consequence of \texttt{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \texttt{Show} constraint for all knots in the datatype structure.
 

### 2018-04-26 lncs-paper.tex
Polish

@@
 In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function.
-Whereas in the case of recursion point, we surround it with parentheses and apply our generic function \texttt{gshow}.
-Thus, the type family for equality model a form of pattern matching on the types again:
+Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \texttt{gshow}.
+Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
 \texttt{
@@
 We do not give its definition here, as it is straightforward.
-\begin{figure}[t]
+\begin{figure}[!b]
 \centering
@@
 
-We finally define \texttt{Context} as a newtype that wraps the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper interface functions.
+We finally define \texttt{Context} as a newtype wrapping the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper operations.
 \texttt{

### 2018-04-26 lncs-paper.tex

@@
 \begin{abstract}
-Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express a recursive structure.
-Some approaches based on sums of products, however, do not use a fixed point.
-Those views therefore allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues when it needs to deal with recursion.
+Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express their recursive structure.
+However, some approaches, based on sums of products, do not use a fixed point.
+They allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues otherwise.
 A known and unwelcome solution is the use of overlapping instances.
-We present an approach that uses closed type families to eliminate the need of overlap for handling recursion.
+We present an alternative approach that uses closed type families to eliminate the need of overlap for handling recursion in datatypes.
 Moreover, we show that our idiom allows for families of mutually recursive datatypes.
@@
 
-A classical way to generically express a datatype is to represent its constructors as the chains of nested binary sums, and turn constructor arguments into the chains of nested binary products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
-De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that both are lists of types, a sum of products is thus a list of lists of types.
-They call their view SOP which stands for a ``sum of products''---this is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
+A classical way to generically express a datatype is to represent its constructors as the chains of nested \emph{binary} sums, and turn constructor arguments into the chains of nested \emph{binary} products~\cite{VanNoort2008,Cheney2002,Loeh2004}\nocite{Rodriguez2008}\nocite{Magalhaes2012}.
+De Vries and L\"{o}h~\cite{VriLoeh2014} describe a different sum-of-products approach to representing data using $n$-ary sums and products that are both lists of types; a sum of products is thus a list of lists of types.
+They call their view SOP which stands for a ``sum of products''.
+It is implemented in the \textsf{generics-sop}~\cite{generics-sop} library and is based on several relatively recent extensions to the Haskell type system, such as \emph{data kinds}, \emph{kind polymorphism}~\cite{Yorgey2012} and \emph{constraint kinds}.
 %, and \emph{GADTs}~\cite{Schrijvers2009}. -- very old extension, 2003 approx.
-Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as for constructing sums and products, collapsing to homogeneous structures, and application, which are an expressive instrument for defining generic functions in a more succinct and high-level style compared to the classical binary sum-of-products views.
+Using these Haskell features, the library provides the generic view and equips it with a rich collection of high-level combinators, such as ones for constructing sums and products, collapsing to homogeneous structures, and others.
+They form an expressive instrument for defining generic functions in a more succinct and high-level style as compared to the classical binary sum-of-products views.
 
-There are many generic functions that deal with the recursive knots, when traverse the structure of datatypes.
-Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}, more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}.
+There are many generic functions that deal with the recursive knots when traversing the structure of datatypes.
+Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}; more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}.
 For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}.
-The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise it yields.
+The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise becomes unhandy.
 
-One possible solution to the aforementioned shortcoming of SOP is to modify the SOP core by explicitly encoding recursive positions using the fixed-point approach.
-This may complicate the whole framework significantly.
+One possible solution to the aforementioned shortcoming of SOP is to modify the its core by explicitly encoding recursive positions using the fixed-point approach.
+However, this may complicate the whole framework significantly.
 Besides, such a decision may lead to extra conversions between the generic views.
@@
 Another known solution uses overlapping instances.
-This usually unwelcome Haskell extension complicates reasoning about the semantics of code.
-In particular, the program behavior becomes unstable for it can be affected by any module defining more specific instances.
+This, usually unwelcome, Haskell extension complicates reasoning about the semantics of code.
+In particular, the program behavior becomes unstable, for it can be affected by any module defining more specific instances.
 Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
 % and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
-The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} for GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
+The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
 %Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
@@
 
-We feel both existing approaches unsatisfactory and make the following contributions:
+We feel both existing approaches unsatisfactory and make the following contributions.
 \begin{itemize}
@@
 The zipper is meant to be easily and flexibly used with families of mutually recursive datatypes (Section~\ref{sec:generic-zipper}).
-\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations which necessarily  arise in practice as a consequence of absence of information about recursion points.
+\item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points.
 An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{sec:gshow}.
@@
 
-We believe that our idea is suitable for any sum-of-products approach that do not exploit a fixed point view and thus subject to the problem.
+We believe that our idea is suitable for any sum-of-products approach that do not exploit the fixed point view and thus subject to the problem.
 We choose the \textsf{generics-sop} approach as a case study because it appears to be a widely applicable library and builds on powerful language extensions implemented in GHC.
-
-%\paragraph{Organization} The paper is organized as follows.
-In Section~\ref{sec:sop-problem}, we introduce the SOP view concepts and demonstrate the problem by a short example.
-In Section~\ref{sec:handling-recursion}, we show our solution using closed type families and accompany it with one more example of a generic function adapting the idea.
-In Section~\ref{sec:generic-zipper}, we present an advanced example of the generic zipper for mutually recursive datatypes.
-In Section~\ref{sec:related-work}, we review related work, and we conclude in Section~\ref{sec:conclusion}.
-%
-% Organization part is not used these days (cf. any of the later papers by SPJ)
-% Instead, the Sec.-references are integrated into the Contributions part.
 
@@
 
-In this section, we first review the SOP view on data describing its basic concepts to introduce the terminology we are using.
+In this section, we first review the SOP view on data, describing its basic concepts to introduce the terminology we are using.
 Then we discuss the problem with handling recursion by generic functions and illustrate it with a short example.
@@
 The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the index list.
-Therefore, \texttt{NP} is a tiny abstraction of a usual definition of a heterogeneous list The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic.
-The index list allows to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.
+Therefore, \texttt{NP} is a modest abstraction over a heterogeneous list.
 
+The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic.
+The index list is allowed to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.
 Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analogue of \texttt{const}:
@@
 }
-
-If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list of types, while \texttt{K a} turns it into homogeneous one. Here is an example value of \texttt{NP I}:
+If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list, while \texttt{K a} turns it into a homogeneous one, isomorphic to \texttt{[a]}.
+Here is an example value of type \texttt{NP I}:
 \texttt{
@@
 \vs
-\indent S (S (Z (I 3))) \ann NS I '[Char, Bool, Int, Bool]
+\indent S (S (Z (I 5))) \ann NS I '[Char, Bool, Int, Bool]
 \vs

### 2018-04-26 lncs-paper.tex
Polish

@@
 
-One possible solution to the aforementioned shortcoming of SOP is to modify the its core by explicitly encoding recursive positions using the fixed-point approach.
+One possible solution to the aforementioned shortcoming of SOP is to modify its core by explicitly encoding recursive positions using the fixed-point approach.
 However, this may complicate the whole framework significantly.

### 2018-04-26 lncs-paper.tex
Sec 4: Intro

@@
 
-\section{The generic Zipper}
+\section{The generic zipper}
 \label{sec:generic-zipper}
 
-The zipper is a data structure that enables efficient editing and navigation within the tree-like structure of a datatype by representing a current location in that structure.
-The location is a focus, which can be one of recursive nodes in the tree, along with its context that consists of surroundings of the focal subtree.
+The zipper is a data structure that enables efficient navigation and editing within the tree-like structure of a datatype.
+It  represents a current location in that structure, a \emph{focus}, storing a tree node along with its context.
+%that consists of surroundings of the focal subtree.
 
-The classical zipper described by Huet~\cite{Huet1997} can be generically produced for regular datatypes~\cite{HiJeLo2004}, which are a subset of datatypes that can be viewed as a least fixed point of some polynomial expression on types.
+The classical zipper described by Huet~\cite{Huet1997} can be generically calculated for regular datatypes~\cite{HiJeLo2004}---a subset of datatypes that can be viewed as a least fixed point of some polynomial expression on types.
 Yakushev~et~al.~\cite{MuRec2009} generalize the definition of the generic zipper for an arbitrary family of mutually recursive datatypes.
-Those known solutions require a datatype to be expressed using forms of a fixed-point operator, since the zipper operates on recursion points.
-Using closed type families, it is possible to define the generic zipper using a representation that does not exploit a fixed point.
+All mentioned solutions require a datatype to be expressed using forms of a fixed-point operator, since the zipper operates on recursion points.
+
+Our approach allows to define the generic zipper using a representation that does not exploit a fixed point.
 In this section, we implement the generic zipper interface that includes functions for manipulating locations for mutually recursive datatypes using the SOP view.
@@
 
-Having the location of the zipper, which holds the current focus on one of recursive knots in the structure---for the zipper for mutually recursive datatypes, this means recursion points of the total structure of a family---and its context, we may produce a new location by moving that focus up, down, left, or right.
-We are first going to show the generic zipper interface and an example of how it can be used, and then proceed to implement that interface.
-The interface we provide for using the generic zipper is displayed in Figure~\ref{fig:zipper-interface}.
-This comprises the functions for \emph{movement}, \emph{starting} and \emph{ending navigation}, and \emph{updating} the focus, which are defined over the location structure.
+Having a zipper focused on one of recursive knots in a structure, we may produce a new location by moving the focus up, down, left, or right.
+%for the zipper for mutually recursive datatypes, this means 
+%recursion points of the total structure of a family
+We start with the generic zipper interface for those operations and an example of how it can be used, then proceeding to the implementation.
+
+The interface we provide for the generic zipper is shown on Figure~\ref{fig:zipper-interface}.
+It comprises the functions for \emph{movement}, \emph{starting} and \emph{ending navigation}, and \emph{updating} the focus, which are defined over the location structure.
 \begin{figure}[t]

### 2018-04-26 lncs-paper.tex
JB: Sec 2.2

@@
 The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code.
-The functions \texttt{from} and \texttt{to} perform a shallow conversion of the datatype topmost layer---it does not recursively translate the constructor arguments.
+The functions, \texttt{from} and \texttt{to}, perform a shallow conversion of the datatype topmost layer---they does not recursively translate the constructor arguments.
 
-We are leaving out discussion of the \texttt{All SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work.
-Although, we do use \texttt{All} constraint combinator in the following.
-Therefore, it is worth noting that \texttt{All} applies particular constraint (e.g.
+We leave out discussion of the \texttt{SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work.
+Although, we do use \texttt{All} constraint combinator (as in \texttt{All SListI}) in the following.
+Therefore, it is worth noting that \texttt{All} applies a particular constraint (e.g.
 \texttt{SListI} above) to each member of a list of types.
@@
 
-We have introduced generic representation employed by the SOP library and are ready to describe the problem of handling recursion points stemming from the representation.
+We have introduced generic representation employed by the SOP library and are ready to describe the problem of handling recursion points, stemming from the representation.
 
@@
 }
-The function \texttt{subterms} translates the term to its representation unwrapping the sum of products from \texttt{SOP} and passes that to the auxiliary function \texttt{subtermsNS} that merely traverses the sum until it reaches the product.
-Once it gets the product, it passes that further to \texttt{subtermsNP}.
+The function \texttt{subterms} translates the term to its representation, unwrapping the sum of products from \texttt{SOP}, and passes that to the auxiliary function \texttt{subtermsNS}.
+The latter merely traverses the sum and, once reaches the product,  passes it further to \texttt{subtermsNP}.
 
-The algorithm of \texttt{subtermsNP} is straightforward---it traverses the product appending current element to the result list if its type is the same as the given term's, otherwise skipping the element.
+The algorithm of \texttt{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \texttt{t}, otherwise skipping the element.
 We use GHC's \textsf{TypeAppications} extension to pass that type.
@@
 There is no clear path to this at the moment.
-Therefore, we step back and follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.} using overlapping instances of a dedicated class instead.
+Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \texttt{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
 \texttt{
@@
 
-To make the whole solution work we need to propagate the constraints all the way through \texttt{subtermsNS} and \texttt{subterms} signatures:
+To make the whole solution work, we need to propagate the constraints all the way through \texttt{subtermsNS} and \texttt{subterms} signatures:
 \texttt{
@@
 
-Although, the approach works as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
-That would avoid the complexity ovehead introduced by the approach, as we have noted in the introduction.
+Although, the approach works, as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
+This would remove the complexity overhead introduced by the approach, as we have mentioned in the introduction.
 
@@
 
-\subsection{Solution to \texttt{subterms} revised}
+\subsection{Solution to \texttt{subtermsNP} revised}
+\label{subsec:solve-subtermnp}
 

### 2018-04-26 lncs-paper.tex

@@
 
-In the previous section, we have shown a solution to the problem of handling recursion, which makes use of overlapping instances. We are going to improve the solution and remove overlap now.
+In the previous section, we have shown a solution to the problem of handling recursion, which uses overlapping instances. We are going to improve the solution and remove overlap now.
 
 Closed type families are the Haskell language extension introduced by Eisenberg et al.~\cite{Eisenberg2014}.
-The main idea of the extension being that the equations for a \emph{closed type family} are disallowed outside its declaration.
+The main idea of the extension is that the equations for a \emph{closed type family} are disallowed outside its declaration.
 Under the extension, we can give the following definition of type-level equality:
@@
 
-We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between the equal types by defining a type class:
+We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
 \texttt{
@@
 
-Over the years, there has been a lot of works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated.
-Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in different aspects.
-There are several generic views that use forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}.
+There are many works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated.
+Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in various aspects.
+There are several generic views that use certain forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}.
 And there are a number of approaches that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation.
-The SOP view~\cite{VriLoeh2014} which we use to demonstrate our technique is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype.
+The SOP view~\cite{VriLoeh2014}, which we use to demonstrate our technique, is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype.
 This approach uses heterogeneous lists of types to encode sums and products in the generic representation.
@@
 The idea similar to SOP has been proposed by Kiselyov~et~al.~\cite{Kiselyov2004} in their \textsf{HList} library for strongly typed heterogeneous collections.
-In their paper, they also discuss problems concerened with overlap which they use for access operations.
-They restrict overlap by introducing a type class using a functional dependency, which provides functionality of type equality that is similar to our solution.
-Morris and Jones~\cite{Morris2010} introduce the type-class system \textsf{ilab}, based on the Haskell 98 class system, with a new feature, called ``instance chains'', that allows to control overlap by using an explicit syntax in instance declarations, which resembles if-else chains.
-But use of instance chains and local use of overlap leave code error-prone as a consequence of type class oppennes.
+In the paper, the authors also discuss problems connected with overlap, which they use for access operations.
+Another Haskell extension, functional dependencies, is applied to supply overlap-free machinery of type equality there, which resembles our solution.
+
+Morris and Jones~\cite{Morris2010} introduce the type-class system \textsf{ilab}, based on the Haskell 98 class system, with a new feature called ``instance chains''.
+This enables one to control overlap by using an explicit syntax in instance declarations.
+The approach resembles if-else chains.
+But the use of instance chains and local use of overlap leave code error-prone as a consequence of type class openness.
 Closed type families~\cite{Eisenberg2014} were recently introduced in Haskell to solve the overlap problem.

### 2018-04-26 lncs-paper.tex
Conclusion

@@
 
-It was believed until today, that defining generic functions that consider recursion points when looking at the generic representation of a datatype structure is only possible within generic views that explicitly express recursion in the datatype representation, whereas the other commonly used approaches to generic programming was considered as unsuitable for this task.
-There are some approaches that address this problem by means of local overlaps, where it needs to access the points of recursion.
-We have developed the technique that allows us to define generic functions that treat recursion without its explicit encoding and overlap.
-This makes use of closed type families that are the recent Haskell extension.
+Defining generic functions, which consider recursion points, is easy within generic views that are explicit about recursion in the datatype representation.
+Not so much otherwise.
+Although, there are some approaches that address the problem by means of global or local overlaps.
+We have developed the technique that allows one to define generic functions that treat recursion without its explicit encoding and without overlap.
 
-We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface, and supports families of mutually recursive datatypes.
-Though it is still easier to treat recursion where the explicit encoding is used, once we have shown that the problem of handling recursion in datatype-generic approaches is not critical, it encourages to invent new generic universes not worrying about the recursion support, but rather focusing on other generic programming tasks.
+We have demonstrated that the method suits for advanced recursive schemes, such as the generic zipper interface. Also, it supports families of mutually recursive datatypes.
+
+Arguably, it is still easier to treat recursion when ``explicit'' encoding is used.
+On the other hand, we believe, once the problem of handling recursion is shown to be manageable, new generic universes shall emerge, not worrying about the recursion support, but rather focusing on other generic programming problems.
 

### 2018-04-27 lncs-paper.tex
Polish

@@
 The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code.
-The functions, \texttt{from} and \texttt{to}, perform a shallow conversion of the datatype topmost layer---they does not recursively translate the constructor arguments.
+The functions, \texttt{from} and \texttt{to}, perform a shallow conversion of the datatype topmost layer---they do not recursively translate the constructor arguments.
 
@@
 The zipper is a data structure that enables efficient navigation and editing within the tree-like structure of a datatype.
-It  represents a current location in that structure, a \emph{focus}, storing a tree node along with its context.
+It represents a current location in that structure, storing a tree node, a \emph{focus}, along with its context.
 %that consists of surroundings of the focal subtree.
@@
 %recursion points of the total structure of a family
-We start with the generic zipper interface for those operations and an example of how it can be used, then proceeding to the implementation.
+We start with the generic zipper interface for those operations and an example of how it can be used, then proceed to the implementation.
 
@@
 Together with the datatype of families, we define a class \texttt{ProofIn} that provides a proof of membership of a family, which generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
-The definition of the \texttt{ProofIn} class and its auxiliary class \texttt{ProofFam} is given in Figure~\ref{fig:proof-in}.
+The definitions of \texttt{ProofIn} and its auxiliary class \texttt{ProofFam} are given in Figure~\ref{fig:proof-in}.
 The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
@@
 We do not give its definition here, as it is straightforward.
-\begin{figure}[!b]
+\begin{figure}[t]
 \centering
@@
 In the paper, the authors also discuss problems connected with overlap, which they use for access operations.
-Another Haskell extension, functional dependencies, is applied to supply overlap-free machinery of type equality there, which resembles our solution.
+Another Haskell extension, functional dependencies, is applied to restrict overlap in type equality there, which resembles our solution.
 

### 2018-05-05 lncs-paper.tex

@@
 Moreover, we show that our idiom allows for families of mutually recursive datatypes.
+
+\textbf{Category:} full research paper.
+
+\textbf{Studentship:} both authors are students.
 
@@
 
-\subsection{Interface and using}
+\subsection{Interface and usage}
 
@@
 The functions \texttt{goUp}, \texttt{goDown}, \texttt{goLeft}, and \texttt{goRight} produce a location with the focus moved \emph{up} to the parent of the focal subtree, \emph{down} to its leftmost child, \emph{left} and \emph{right} to the left and right sibling, respectively, if it is possible.
-Movement may fail, as specified by the \texttt{Maybe} monad, if we cannot go further in a chosen direction.
+A movement may fail, as specified by the \texttt{Maybe} monad, if we cannot go further in a chosen direction.
+Navigation in a tree starts at the root, and the type variable \texttt{a} refers to the root type that remains the same during the navigation, while the type in the focus of the location may vary and is one of the types in a type list \texttt{fam}.
 
-The function signature of \texttt{enter} specifies the constraints necessary for starting navigation in a structure.
-A datatype of the structure needs to have the generic representation; the \texttt{In} constraint is defined over a generalization of \texttt{Equal} from Section~\ref{sec:handling-recursion}, which checks a type for membership of a family; the \texttt{Zipper} constraint collects specific constraints that refer to the implementation of movement operations.
-The universal quantifier here sets the instantiation order of the type variables for type applications that will be a part of our usage pattern for the zipper.
-Navigation in a tree starts at the root, and the type variable \texttt{a} refers to the root type that keeps fixed during the navigation, since it is necessary for leaving a tree and returning the root, while type in focus of the location may vary and is one of types in the type list \texttt{fam} (of kind \texttt{[*]}) specifying a family.
+The function signature of \texttt{enter} specifies the constraints necessary to begin navigation in a structure.
+Firstly, a datatype of the structure needs to have the \texttt{Generic} representation.
+Secondly, the \texttt{In} constraint checks if type \texttt{a} is a member of a type family \texttt{fam}.
+Thirdly, the \texttt{Zipper} constraint collects specific constraints that refer to the implementation of movement operations.
+Note that the universal quantifier here sets the instantiation order of the type variables for type applications that will be a part of our usage pattern for the zipper.
+
 The \texttt{leave} function ends navigation moving up to the root and returns its modified value.
 
-The function \texttt{update} modifies the focal subtree with a constrained function, where the type in focus is existential in \texttt{Loc}, and the datatype of locations guarantees that the constraint \texttt{c} (actually, it has kind \texttt{* \ra Constraint}) holds for all datatypes in the family \texttt{fam} and therefore for all recursive nodes that can be in focus, whence the function can be applied.
+The \texttt{update} function modifies the focal subtree with a given constrained function.
+The type in focus is existentially quantified inside \texttt{Loc} and should satisfy constraint \texttt{c}.
+The structure of \texttt{Loc} (shown in Section~\ref{ss:locs}) guarantees that the constraint holds for all types in the family \texttt{fam} and, therefore, for all recursive nodes that can be in focus, hence \texttt{update} can always be applied.
 
-Let us show how to use this with an example.
-Consider the following pair of mutually recursive datatypes for a rose tree and a forest, where the forest is a list of trees, and the tree is defined as a value in the node and a forest of its children:
+Consider the following example of usage of the interface.
+Define a pair of mutually recursive datatypes for a rose tree and a forest, where the forest is a list of trees, and the tree is defined as a value in a node and a forest of its children:
 \texttt{
@@
 
-We can define a class that provides a function for updating the trees, for example:
+Updating the trees can be done through a class:
 \texttt{
@@
 }
-This replaces the tree node with a given tree, and for the forests, this leaves the nodes without change. This is also scalable if we want to extend the family.
+This replaces a tree node with a given tree, and, for the forests, this leaves the nodes untouched.
 
-For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and exploit the flipped function composition \comp~and Kleisli composition \kleisli, using the monad interface of \texttt{Maybe} that wraps the result type of the movement functions.
+For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \comp~and Kleisli composition \kleisli.
+The latter one is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
 \texttt{
@@
 
-We now define the family of our datatypes simply as follows:
+The type family we need to run the example is defined as follows:
 \texttt{
@@
 }
-And the example below shows how we can use the zipper operations with our updating function to change a mutually recursive expression:
+
+Finally, we can use zipper operations with our updating function to traverse and replace a part of a forest:
 \texttt{
@@
 
-We also may use the zipper merely for regular datatypes, when we want to move only to recursive knots of a single datatype, just assigning a one-element type-level list to the family at the \texttt{enter} call.
-Or we may enumerate all types, whose nodes we want to visit, in that list.
-This is flexible, as this requires no boilerplate declarations.
-However, we demand an updating operation to be a type class function to distinguish types of the nodes, and enforce a constraint, which may be single or complex, on all types in the family, because the type of the focus is existential, as already said.
+Our zipper applies to regular datatypes as well.
+In that case, \texttt{fam} list shall contain a single element.
+Generally, the interface is flexible enough to allow us to check in any collection of types we are interested in during traversal.
+However, we demand an updating operation to be a type class function to distinguish the types of the nodes.
 
 \subsection{Locations}
+\label{ss:locs}
 

### 2018-05-10 lncs-paper.tex
wibble

@@
 
-The location structure consists of the focal subtree, which is one of the mutually recursive nodes of total structure of the family of datatypes, and its surrounding context.
-We give the following definition of the datatype \texttt{Loc} for locations:
+The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
 \texttt{

### 2018-05-10 lncs-paper.tex
upd intro to 4; fix intro to 4.2; some wibble

@@
 For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
-This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq} which abbreviates heavy-weighted interface of \texttt{Proof}:
+This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq}, which abbreviates the heavy-weighted interface of \texttt{Proof}:
 \texttt{
@@
 It represents a current location in that structure, storing a tree node, a \emph{focus}, along with its context.
-%that consists of surroundings of the focal subtree.
+Having a zipper focused on a recursive knot in a structure, we may produce a new location by moving the focus up, down, left, or right.
+On the way, we can update the nodes.
+Entering and laving the navigation usually need a special care.
 
@@
 
-Our approach allows to define the generic zipper using a representation that does not exploit a fixed point.
-In this section, we implement the generic zipper interface that includes functions for manipulating locations for mutually recursive datatypes using the SOP view.
+In this section, we describe our approach allowing one to define the generic zipper out of a representation that does not exploit a fixed point.
+We start with the generic zipper interface and an example of how it can be used (Section~\ref{ss:zip-interface}).
+Then, we turn to the type-level machinery employed to define locations inside mutually recursive datatypes using the SOP view  (Section~\ref{ss:locs}).
+Finally, we discuss the implementation of the generic zipper interface~--- the functions for manipulating locations (Section~\ref{ss:zip-impl}).
 
 \subsection{Interface and usage}
-
-Having a zipper focused on one of recursive knots in a structure, we may produce a new location by moving the focus up, down, left, or right.
-%for the zipper for mutually recursive datatypes, this means 
-%recursion points of the total structure of a family
-We start with the generic zipper interface for those operations and an example of how it can be used, then proceed to the implementation.
+\label{ss:zip-interface}
 
@@
 }
-The subtree in focus is wrapped by the \texttt{Family} datatype that we use to prove its required properties, while we do not know the current type of the node in focus.
-For being able to move up in the tree, we keep the type that will be obtained by uniting the current focus and context (essentially, the type of the focal node's parent in the tree): that type is unknown outside the location, and is reflected in the type parameters of \texttt{Family} and the \texttt{Contexts} datatype of the context by the type variable \texttt{a} that is existential in the \texttt{Loc} datatype; the type of the focal subtree itself is existential in \texttt{Family}.
-The other type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} mean the root type of the tree, the list of types constituting the family, and the constraint constructor imposing restrictions on the family's types, respectively.
+The type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively.
+Also, the single constructor is existentially quantified over one more type variable, \texttt{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure.
+We discuss both term parameters of the constructor of \texttt{Loc} in detail below.
 
@@
 
-We define the \texttt{Family} datatype as a GADT, with embedded constraints on its constructor:
+The subtree in a focus is wrapped by the \texttt{Family} datatype. 
+The wrapper incapsulates the proofs about a number of important properties of a focus.
+We define the \texttt{Family} datatype as a GADT, with embedded 
+constraints on its constructor:
 \texttt{
@@
 \subsection{Implementing the zipper interface}
+\label{ss:zip-impl}
 
@@
 
-There are many works that contribute to datatype-generic programming, and a lot of knowledge has been accumulated.
+There are many works that contribute to datatype-generic programming.
 Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in various aspects.

### 2018-05-10 lncs-paper.tex

@@
 The wrapper incapsulates the proofs about a number of important properties of a focus.
-We define the \texttt{Family} datatype as a GADT, with embedded 
-constraints on its constructor:
 \texttt{
@@
 }
-Though we do not know the type of the focus that is the existential type \texttt{b} in the family, we can obtain enough information about that type to provide the zipper functionality, which is captured by the restrictions above.
-We guarantee that the type \texttt{b} is a member of the family, and satisfies a set of constraints of the whole zipper interface, collected in the \texttt{ZipperI} constraint: in particular, it ensures that \texttt{a} is the type of the parent node for the node of focus of type \texttt{b} in the tree.
+Existential type variable \texttt{b} represents the type of a focus.
+We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus.
+Firstly, the type of a focus should have the \texttt{Generic} representation.
+Secondly, it should live \texttt{In} the list of types we are going to visit.
+Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate.
+In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.
 
-We define the \texttt{In} constraint by means of a type family \texttt{InFam} that is a generalization of \texttt{Equal} defined in Section~\ref{sec:handling-recursion}.
-The following checks a type for membership of a type-level list that we use to express a family:
+We implement the \texttt{In} constraint by means of a type family \texttt{InFam} exactly along the lines of \texttt{Equal} family defined in the beginning of Section~\ref{sec:handling-recursion}. 
+%~ The following checks a type for membership of a type-level list that we use to express a family:
+%~ \texttt{
+%~ \vs
+%~ \indent type family InFam (a \ann k) (fam \ann [k]) \ann Bool where\\
+%~ \indent\s\s InFam a (a ': fam) = 'True\\
+%~ \indent\s\s InFam a (x ': fam) = InFam a fam\\
+%~ \indent\s\s InFam a '[]\ind\ind = 'False
+%~ \vs
+%~ }
+%~ The \texttt{In} constraint is defined as
 \texttt{
 \vs
-\indent type family InFam (a \ann k) (fam \ann [k]) \ann Bool where\\
-\indent\s\s InFam a (a ': fam) = 'True\\
-\indent\s\s InFam a (x ': fam) = InFam a fam\\
-\indent\s\s InFam a '[]\ind\ind = 'False
+\indent type In a fam = InFam a fam \tteq 'True\\
 \vs
 }
-The \texttt{In} constraint is defined as
-\texttt{
-\vs
-\indent type In a fam = InFam a fam \tteq 'True
-\vspace{0.2cm}
-}
+The definition of \texttt{InFam} is omitted as a boring one.
 
-Together with the datatype of families, we define a class \texttt{ProofIn} that provides a proof of membership of a family, which generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
-The definitions of \texttt{ProofIn} and its auxiliary class \texttt{ProofFam} are given in Figure~\ref{fig:proof-in}.
-The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
+One last missing piece for managing focuses is the class \texttt{ProofIn}.
+it provides a proof of membership of a focus type to a family.
+Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
+The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFam} are given in Figure~\ref{fig:proof-in}.
+%The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
+% this is not clear and, I think, can be omitted
+
 \begin{figure}[t]
@@
 
+\subsubsection{Type-level Differentiation}
+
 McBride~\cite{McBride2001} studies a relation between the one-hole context definition and \emph{partial differentiation} from calculus: he shows that the type of the context for an arbitrary (regular) type can be derived mechanically from that type by means of a list of differentiation \emph{rules} that serve as formulaic instructions for computing the type in type-level programming.
-Yakushev~et~al.~\cite{MuRec2009} then demonstrate that his method can be generalized for mutually recursive datatypes.
+Yakushev~et~al.~\cite{MuRec2009} then demonstrate that the method can be generalized for mutually recursive datatypes.
 We adapt that technique to \textsf{generics-sop}, and now need a few auxiliary type-level functions to implement the computation of the context type.

### 2018-05-10 lncs-paper.tex
Typos and some wibble

@@
 }
-We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise the synonym for the matching function:
+We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise a synonym for the matching function:
 \texttt{
@@
 
-The resulting function \texttt{gshow} is a subject of a number of constraints abbreviated by \texttt{GShow} synonym:
+The resulting function \texttt{gshow} is a subject of a number of constraints abbreviated by a \texttt{GShow} synonym:
 \texttt{
@@
 On the way, we can update the nodes.
-Entering and laving the navigation usually need a special care.
+Entering and leaving the navigation usually need a special care.
 
@@
 The \texttt{update} function modifies the focal subtree with a given constrained function.
-The type in focus is existentially quantified inside \texttt{Loc} and should satisfy constraint \texttt{c}.
+The type in focus is existentially quantified inside \texttt{Loc} and should satisfy the constraint \texttt{c}.
 The structure of \texttt{Loc} (shown in Section~\ref{ss:locs}) guarantees that the constraint holds for all types in the family \texttt{fam} and, therefore, for all recursive nodes that can be in focus, hence \texttt{update} can always be applied.
@@
 
-We implement the \texttt{In} constraint by means of a type family \texttt{InFam} exactly along the lines of \texttt{Equal} family defined in the beginning of Section~\ref{sec:handling-recursion}. 
+We implement the \texttt{In} constraint by means of a type family \texttt{InFam} exactly along the lines of the \texttt{Equal} type family defined in the beginning of Section~\ref{sec:handling-recursion}.
 %~ The following checks a type for membership of a type-level list that we use to express a family:
@@
 One last missing piece for managing focuses is the class \texttt{ProofIn}.
-it provides a proof of membership of a focus type to a family.
+It provides a proof of membership of a focus type to a family.
 Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
@@
 In the paper, the authors also discuss problems connected with overlap, which they use for access operations.
-Another Haskell extension, functional dependencies, is applied to restrict overlap in type equality there, which resembles our solution.
+Another Haskell extension, functional dependencies, is applied to restrict overlap by introducing a class for type equality there, which resembles our solution.
 

### 2018-05-10 lncs-paper.tex
Replace  with

@@
 For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \comp~and Kleisli composition \kleisli.
-The latter one is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
+The latter is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
 \texttt{
@@
 \indent data Loc (r \ann *) (fam \ann [*]) (c \ann * \ra Constraint) where\\
-\indent\s\s Loc \ann\hs Family r a fam c \ra Contexts r a fam c\\
+\indent\s\s Loc \ann\hs Focus r a fam c \ra Contexts r a fam c\\
 \indent\ind\s\s\ra Loc r fam c
@@
 
-\subsubsection{Families}
+\subsubsection{Focus}
 
-The subtree in a focus is wrapped by the \texttt{Family} datatype. 
+The subtree in a focus is wrapped by the \texttt{Focus} datatype. 
 The wrapper incapsulates the proofs about a number of important properties of a focus.
@@
 \vs
-\indent data Family (r \ann *) (a \ann *) (fam \ann [*])\\
+\indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
 \indent\ind\ind\ind (c \ann * \ra Constraint) where\\
-\indent\s\s Family \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
-\indent\ind\ind\s\Ra b \ra Family r a fam c
+\indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
+\indent\ind\ind\s\Ra b \ra Focus r a fam c
 \vs
@@
 Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
-The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFam} are given in Figure~\ref{fig:proof-in}.
+The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFocus} are given in Figure~\ref{fig:proof-in}.
 %The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
@@
 \begin{tabular}{l}
-\tt class ProofFam (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
+\tt class ProofFocus (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
 \tt\ind\ind\ind\s\s\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
-\tt\s\s witness \ann b \ra Maybe (Family r a fam c)
+\tt\s\s witness \ann b \ra Maybe (Focus r a fam c)
 \vs
-\tt instance ProofFam 'False r a b fam c where\\
+\tt instance ProofFocus 'False r a b fam c where\\
 \tt\s\s witness \_ \hs\nths= Nothing\\
 \tt instance (Generic b, In b fam, ZipperI r a b fam c)\\
-\tt\ind\Ra ProofFam 'True r a b fam c where\\
-\tt\s\s witness\s\s\s = Just . Family
+\tt\ind\Ra ProofFocus 'True r a b fam c where\\
+\tt\s\s witness\s\s\s = Just . Focus
 \vs
-\tt class ProofFam (InFam b fam) r a b fam c\\
+\tt class ProofFocus (InFam b fam) r a b fam c\\
 \tt\ind\Ra ProofIn r a b fam c\\
-\tt instance ProofFam (InFam b fam) r a b fam c\\
+\tt instance ProofFocus (InFam b fam) r a b fam c\\
 \tt\ind\Ra ProofIn r a b fam c
@@
 }
-This has the shape similar to the previously defined \texttt{Loc} and \texttt{Family} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Family} in the \texttt{Loc} constructor.
+This has the shape similar to the previously defined \texttt{Loc} and \texttt{Focus} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Focus} in the \texttt{Loc} constructor.
 The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame here indicates that the constraint for the zipper holds after plugging the hole, so all the properties, conversely, can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack---and the \texttt{CNil} constructor for an empty context with the \texttt{r} and \texttt{a} types equal is the inductive basis in that proof.
@@
 \indent goDown \ann Loc a fam c \ra Maybe (Loc a fam c)\\
-\indent goDown (Loc (Family t) cs)\\
+\indent goDown (Loc (Focus t) cs)\\
 \indent\s\s = case toFirst t of\\
@@
 \tt toFirst \ann\hs\fa fam c r a. (Generic a, ToFirst r a fam c)\\
-\tt\ind\ind\Ra a \ra Maybe (Family r a fam c)\\
+\tt\ind\ind\Ra a \ra Maybe (Focus r a fam c)\\
 \tt toFirst t = appToNP @AllProof toFirstNP \$ unSOP \$ from t
@@
 \tt toFirstNP \ann\hs\fa fam c r a xs. All (ProofIn r a fam c) xs\\
-\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Family r a fam c)\\
+\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Focus r a fam c)\\
 \tt toFirstNP (I (x \ann b) :* xs)\\

### 2018-05-11 lncs-paper.tex

@@
 
-When we focus on a particular node in a data structure, having the surrounding context of that node is enough to reconstruct the entire structure.
-The context for the tree's location has the shape of the original structure of that tree but with one hole at the place of its focus, whence it is sometimes called a \emph{one-hole context}.
-The context can be expressed as a stack (\texttt{Contexts} below), where each frame \texttt{Context} corresponds to the particular node with a hole: it ascends from the focal node keeping its siblings, the siblings of its parent, and until it reaches the root node, through all levels in the tree.
-So the stack of contexts essentially reflects the track of the movement in the structure.
+A focus on a particular node, augmented with surrounding context of that node, is enough to reconstruct the entire structure.
+Therefore, the context of a location has the shape of the original structure but with one hole at the place of its focus.
+This is sometimes called a \emph{one-hole context}.
+
+The context can be expressed as a stack, called \texttt{Contexts}, and each frame, \texttt{Context}, corresponds to the particular node with a hole.
+The stack ascends from the focal node keeping its siblings, the siblings of its parent, etc., until it reaches the root node.
+So the stack of contexts, essentially, reflects the track of the movement inside the structure.
 \texttt{
@@
 }
-This has the shape similar to the previously defined \texttt{Loc} and \texttt{Focus} datatypes: it keeps the types of the root and of the immediate parent of the focal node (of the union with the focal subtree when that plugs the hole)---as ensured by sharing that type with \texttt{Focus} in the \texttt{Loc} constructor.
-The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame here indicates that the constraint for the zipper holds after plugging the hole, so all the properties, conversely, can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack---and the \texttt{CNil} constructor for an empty context with the \texttt{r} and \texttt{a} types equal is the inductive basis in that proof.
-Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype, as we do not need that.
+The type parameters have the same meaning as for the \texttt{Loc} datatype.
+The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame indicates that the constraint for the zipper holds after plugging the focus in the hole.
+Therefore, all the properties can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack.
+The \texttt{CNil} constructor for an empty context, with the \texttt{r} and \texttt{a} types being equal, forms the inductive basis in that kind of proof.
+
+Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype.
 
@@
 
-Using the defined operations, we can implement differentiation of a product of types.
+\subsubsection{Context Frame}
+
+At this point, we can implement differentiation of a product of types and, therefore, the computation of a context frame type.
+
 The definition of differentiation shown in Figure~\ref{fig:diff-prod} resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
@@
 
-We finally define \texttt{Context} as a newtype wrapping the result of the computation, as it allows GHC to perform type inference for the context type, where it is possible, to avoid extra type applications for defining the zipper operations.
+Finally, a context frame has two representations: as a type synonym and as a newtype wrapper
 \texttt{
@@
 }
-The \texttt{CtxCode} type is the computed type of the context for the given code of a datatype.
-We will use this type synonym further when defining constraints for functions providing the generic zipper interface.
+The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym.
+On the other hand, the \texttt{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.
 
@@
 
-There are many works that contribute to datatype-generic programming.
+There are many works that contribute to the datatype-generic programming.
 Rodriguez et al.~\cite{Rodriguez2008} and Magalh{\~{a}}es and L{\"{o}}h~\cite{MagLoeh2012} review a number of existing approaches and provide their detailed comparison in various aspects.

### 2018-05-11 lncs-paper.tex
Wibble

@@
 
-Although, the approach works, as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
+Although the approach works, as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
 This would remove the complexity overhead introduced by the approach, as we have mentioned in the introduction.
@@
 \indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
-\indent\ind\ind\ind (c \ann * \ra Constraint) where\\
+\indent\ind\ind\s\s\s (c \ann * \ra Constraint) where\\
 \indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
-\indent\ind\ind\s\Ra b \ra Focus r a fam c
+\indent\ind\ind\Ra b \ra Focus r a fam c
 \vs
@@
 \tt class ProofFocus (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
-\tt\ind\ind\ind\s\s\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
+\tt\ind\ind\ind\ind\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
 \tt\s\s witness \ann b \ra Maybe (Focus r a fam c)
@@
 
-A focus on a particular node, augmented with surrounding context of that node, is enough to reconstruct the entire structure.
+A focus on a particular node, augmented with a surrounding context of that node, is enough to reconstruct the entire structure.
 Therefore, the context of a location has the shape of the original structure but with one hole at the place of its focus.
@@
 
-The definition of differentiation shown in Figure~\ref{fig:diff-prod} resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
+The definition of differentiation, shown in Figure~\ref{fig:diff-prod}, resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
 When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \texttt{Hole}.

### 2018-05-14 lncs-paper.tex
+ acknowledgements

@@
 
+\section*{Acknowledgments}
+\label{sec:acks}
+
+We are grateful to Julia Belyakova who helped to proof-read certain parts of the paper, 
+and to the participants of the Seminar on Programming Languages and Compilers at 
+I.I.~Vorovich Institute of Mathematics, Mechanics and Computer Science (Southern Federal
+University, Russia) where we presented partial results of the work and got valuable feedback.
+
+This project has received funding from the European Research Council (ERC) under the
+European Union's Horizon 2020 research and innovation programme (grant agreement No 695412).
+
 %

### 2018-07-18 lncs-paper.tex
Polymorphic recursion

--- 2018-05-11 lncs-paper.tex
@@
 \newcommand{\ann}{:\nhs:\s}
+\newcommand{\ds}{.\nhs.\nhs.}
 
@@
 \item We describe the problem with the current approach of SOP in detail (Section~\ref{sec:sop-problem}).
-\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation. (Section~\ref{sec:handling-recursion})
+\item We introduce an idiom that overcomes the problem. The approach avoids both, the use of overlapping instances and changing a generic representation (Section~\ref{sec:handling-recursion}).
 % -- 
@@
 \item We note, that our approach can contribute to the \textsf{generics-sop}'s one eliminating some boilerplate instance declarations, which necessarily  arise in practice as a consequence of absence of information about recursion points.
-An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{sec:gshow}.
+An example of that, taken from the \textsf{basic-sop}~\cite{basic-sop} package, is discussed in Section~\ref{subsec:gshow}.
 \end{itemize}
@@
 \indent subterms \ann Generic a \Ra a \ra [a]\\
-\indent subterms t = subtermsNS (unSOP \$ from t)
+\indent subterms t = subtermsNS \$ unSOP \$ from t
 \vs
@@
 
-In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms} task.
+In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms}' task.
 We give another example showing how to adapt our idiom to different scenarios in the following subsection.
@@
 \subsection{Generic show}
-\label{sec:gshow}
+\label{subsec:gshow}
 
@@
 This is a consequence of \texttt{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \texttt{Show} constraint for all knots in the datatype structure.
+
+\section{Polymorphic recursion}
+\label{sec:polymorphic-recursion}
+
+The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}.
+Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g.
+\texttt{Tree~a} from Section~\ref{sec:sop-view}).
+We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent.
+It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.
+
+Assume we have a polymorphically recursive datatype \texttt{PolyRec~a} defined in terms of a type family \texttt{Poly}:
+\texttt{
+\vs
+\indent data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
+\vs
+\indent type family Poly a where\\
+\indent \ind  Poly Bool = Char\\
+\indent \ind  Poly Char = Bool\\
+\indent \ind  Poly a\ind = a
+\vspace{0.2cm}
+}
+
+For managing polymorphic recursion in this datatype, we can write an analogue of the \texttt{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types.
+Since any datatype with type parameters \texttt{f a b c \ds} has kind \texttt{* \ra (* \ra (* \ra \ds))}, a \texttt{PolyEq} type family can be defined thus:
+\texttt{
+\vs
+\indent type family PolyEq (a \ann k) (x \ann k) \ann Bool where\\
+\indent \ind  PolyEq (f a) (g b) = PolyEq f g\\
+\indent \ind  PolyEq a\ind\s a\ind\s = 'True\\
+\indent \ind  PolyEq a\ind\s x\ind\s = 'False
+\vspace{0.2cm}
+}
+
+The function \texttt{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \texttt{Equal}.
+The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
+\texttt{
+\vs
+\indent instance Show \s b \Ra CaseShow 'False a b where\\
+\indent \ind    caseShow' \s\s = show\\
+\indent instance GShow b \Ra CaseShow 'True \s a b where\\
+\indent \ind    caseShow' t = "(" ++ gshow t ++ ")"
+\vspace{0.2cm}
+}
+
+Unfortunately, this approach, albeit working well for datatypes defined like \texttt{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
+\texttt{
+\vs
+\indent data Nested a = Epsilon | Nest a (Nested [a])
+\vs
+}
+The culprit is the constraint \texttt{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \texttt{GShow b} instance context above.
 
@@
 
-The subtree in a focus is wrapped by the \texttt{Focus} datatype. 
+The subtree in focus is wrapped by the \texttt{Focus} datatype. 
 The wrapper incapsulates the proofs about a number of important properties of a focus.
@@
 
-Finally, a context frame has two representations: as a type synonym and as a newtype wrapper
+Finally, a context frame has two representations: as a type synonym and as a newtype wrapper:
 \texttt{
@@
 
+\section*{Acknowledgments}
+\label{sec:acks}
+
+We are grateful to Julia Belyakova, who helped to proof-read certain parts of the paper, 
+and to the participants of the Seminar on Programming Languages and Compilers at 
+I.I.~Vorovich Institute of Mathematics, Mechanics, and Computer Science (Southern Federal
+University, Russia), where we presented partial results of the work and got valuable feedback.
+
+This project has received funding from the European Research Council (ERC) under the
+European Union's Horizon 2020 research and innovation programme (grant agreement No 695412).
+
 %

### 2018-07-29 lncs-paper.tex
listings

@@
 \usepackage{amssymb}
+\usepackage{listings}
+\usepackage{fancybox}
+
+\lstset{
+basicstyle=\ttfamily,
+xleftmargin=\parindent,
+aboveskip=0.2cm,
+belowskip=0.2cm,
+basewidth=0.5em,
+numberbychapter=false,
+escapeinside={(/}{/)},
+literate={::}{{\ann}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
+{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\ds}}2
+{>>>}{{\comp}}2 {>=>}{{\kleisli}}2
+}
 
@@
 
-\newcommand{\s}{$\hphantom{~}$}
-\newcommand{\ind}{\s\s\s\s}
-\newcommand{\hs}{\hspace{0.06cm}}
-\newcommand{\ths}{\hspace{0.01cm}}
-\newcommand{\tths}{\hspace{0.003cm}}
-\newcommand{\nths}{\hspace{-0.01cm}}
 \newcommand{\nhs}{\hspace{-0.06cm}}
-\newcommand{\vs}{\vspace{0.2cm}\\}
-
-\newcommand{\Ra}{$\Rightarrow$\s}
-\newcommand{\ra}{$\rightarrow$\s}
+
+\newcommand{\Ra}{$\Rightarrow$}
+\newcommand{\ra}{$\rightarrow$}
 \newcommand{\fa}{$\forall$}
-\newcommand{\tteq}{{\scriptsize$\thicksim$}\s}
-\newcommand{\comp}{{\tt>\nhs>\nhs>}}
-\newcommand{\kleisli}{{\tt>\nhs=\nhs>}}
-\newcommand{\comps}{\comp\s}
-\newcommand{\kleislis}{\kleisli\s}
-\newcommand{\ann}{:\nhs:\s}
+\newcommand{\tteq}{{\scriptsize$\thicksim$}}
+\newcommand{\comp}{>\nhs>\nhs>}
+\newcommand{\kleisli}{>\nhs=\nhs>}
+\newcommand{\ann}{:\texttt{\nhs}:}
 \newcommand{\ds}{.\nhs.\nhs.}
+
+\newcommand{\K}[1]{\lstinline{#1}}
+
+\newcommand{\lset}{\lstset{basicstyle=\normalsize\ttfamily}}
+
+\makeatletter
+\lst@AddToHook{OnEmptyLine}{\vspace{-0.2255cm}}
+\makeatother
+
+\makeatletter
+\newenvironment{CenteredBox}{
+\begin{Sbox}}{
+\end{Sbox}\centerline{\parbox{\wd\@Sbox}{\TheSbox}}}
+\makeatother
 
@@
 % and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
-The overlap problem also strikes in the security setting when code is compiled as \texttt{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
+The overlap problem also strikes in the security setting when code is compiled as \K{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
 %Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
@@
 The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
-\texttt{
-\vs
-\indent\textbf{type family} Code (a \ann *) \ann [[*]]
-\vs
-}
+\begin{lstlisting}
+type family Code (a :: *) :: [[*]]
+\end{lstlisting}
 An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
@@
 Consider, for instance, a datatype of binary trees:
-\texttt{
-\vs
-\indent\textbf{data} Tree a = Leaf a | Node (Tree a) (Tree a)
-\vs
-}
+\begin{lstlisting}
+data Tree a = Leaf a | Node (Tree a) (Tree a)
+\end{lstlisting}
 This datatype is isomorphic to the following code:
-\texttt{
-\vs
-\indent\textbf{type instance} Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
-\vspace{0.2cm}
-}
+\begin{lstlisting}
+type family Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
+\end{lstlisting}
 
 \begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\tt \textbf{data} NP (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
-\tt\s\s Nil\s\s \ann NP f '[]\\
-\tt\s\s (:*) \ann f x \ra NP f xs \ra NP f (x ': xs)
-\vs
-\tt \textbf{data} NS (f \ann k \ra *) (xs \ann [k]) \textbf{where}\\
-\tt\s\s Z \ann f x \ra NS f (x ': xs)\\
-\tt\s\s S \ann NS f xs \ra NS f (x ': xs)
-\end{tabular}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+data NP (f :: k -> *) (xs :: [k]) where
+  Nil  :: NP f '[]
+  (:*) :: f x -> NP f xs -> NP f (x ': xs)
+
+data NS (f :: k -> *) (xs :: [k]) where
+  Z :: f x -> NS f (x ': xs)
+  S :: NS f xs -> NS f (x ': xs)
+\end{lstlisting}}
+\end{CenteredBox}
 \caption{Datatypes for $n$-ary sums and products.}
@@
 
-As shown in Figure~\ref{fig:ns-np}, the datatypes \texttt{NS} for an $n$-ary sum and \texttt{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types.
-The encoding also holds an auxiliary type constructor \texttt{f} (typically, a functor) which is meant to be applied to every element of the index list.
-Therefore, \texttt{NP} is a modest abstraction over a heterogeneous list.
-
-The definitions of \texttt{NS} and \texttt{NP} are kind polymorphic.
-The index list is allowed to contain types of arbitrary kind \texttt{k}, since \texttt{k} turns to \texttt{*} by applying the type constructor \texttt{f}.
-Basic instantiations of type parameter \texttt{f} found in SOP are identity functor \texttt{I}, that is, a type-level equivalent for \texttt{id} function, and a constant functor \texttt{K}, an analogue of \texttt{const}:
-\texttt{
-\vs
-\indent\textbf{newtype} I (a \ann *)\ind\ind\s\s\nhs = I \{unI \ann a\}\\
-\indent\textbf{newtype} K (a \ann *) (b \ann k) = K \{unK \ann a\}
-\vs
-}
-If instantiated with \texttt{I}, \texttt{NP} is a plain heterogeneous list, while \texttt{K a} turns it into a homogeneous one, isomorphic to \texttt{[a]}.
-Here is an example value of type \texttt{NP I}:
-\texttt{
-\vs
-\indent I 5 :* I True :* I 'x' :* Nil \ann NP I '[Int, Bool, Char]
-\vspace{0.2cm}
-}
+As shown in Figure~\ref{fig:ns-np}, the datatypes \K{NS} for an $n$-ary sum and \K{NP} for an $n$-ary product are defined as GADTs and are \emph{indexed}~\cite{HiJeLo2004} by a promoted list of types.
+The encoding also holds an auxiliary type constructor \K{f} (typically, a functor) which is meant to be applied to every element of the index list.
+Therefore, \K{NP} is a modest abstraction over a heterogeneous list.
+
+The definitions of \K{NS} and \K{NP} are kind polymorphic. The index list is allowed to contain types of arbitrary kind \K{k}, since \K{k} turns to \K{*} by applying the type constructor \K{f}.
+Basic instantiations of type parameter \K{f} found in SOP are identity functor \K{I}, that is, a type-level equivalent for \K{id} function, and a constant functor \K{K}, an analogue of \K{const}:
+\begin{lstlisting}
+newtype I (a :: *)          = I {unI :: a}
+newtype K (a :: *) (b :: k) = K {unK :: a}
+\end{lstlisting}
+If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[a]}. Here is an example value of type \K{NP I}:
+\begin{lstlisting}
+I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
+\end{lstlisting}
 
 We turn to the sum definition now.
-The constructor \texttt{S} of \texttt{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \texttt{Z} stores the payload of type \texttt{f x}.
+The constructor \K{S} of \K{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \K{Z} stores the payload of type \K{f x}.
 For example, the following chooses the third element of a sum:
-\texttt{
-\vs
-\indent S (S (Z (I 5))) \ann NS I '[Char, Bool, Int, Bool]
-\vs
-}
+\begin{lstlisting}
+S (S (Z (I 5))) :: NS I '[Char, Bool, Int, Bool]
+\end{lstlisting}
 The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.
 
-With the \texttt{NS} and \texttt{NP} machinery at hand, SOP defines the \texttt{Generic} class with conversion functions \texttt{from} and \texttt{to} witnessing the isomorphism between a datatype and its generic representation:
-\texttt{
-\vs
-\indent\textbf{type} Rep a = SOP I (Code a)
-\vs
-\indent\textbf{class} All SListI (Code a) \Ra Generic (a \ann *) \textbf{where}\\
-\indent\s\s \textbf{type} Code a \ann [[*]]\\
-\indent\s\s from \ann a \ra Rep a\\
-\indent\s\s to\s\s\s \ann Rep a \ra a
-\vs
-}
-The sum of products type, \texttt{SOP f}, is a newtype-wrapper for \texttt{NS (NP f)}, and the structural representation \texttt{Rep} of a datatype \texttt{a} is a type synonym for a \texttt{SOP I} of \texttt{a}'s code.
-The functions, \texttt{from} and \texttt{to}, perform a shallow conversion of the datatype topmost layer---they do not recursively translate the constructor arguments.
-
-We leave out discussion of the \texttt{SListI} constraint in the \texttt{Generic} class definition as irrelevant to our work.
-Although, we do use \texttt{All} constraint combinator (as in \texttt{All SListI}) in the following.
-Therefore, it is worth noting that \texttt{All} applies a particular constraint (e.g.
-\texttt{SListI} above) to each member of a list of types.
-The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \texttt{Constraint}.
+With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to} witnessing the isomorphism between a datatype and its generic representation:
+\begin{lstlisting}
+type Rep a = SOP I (Code a)
+
+class All SListI (Code a) => Generic (a :: *) where
+  type Code a :: [[*]]
+  from :: a -> Rep a
+  to   :: Rep a -> a
+\end{lstlisting}
+The sum of products type, \K{SOP f}, is a newtype-wrapper for \K{NS (NP f)}, and the structural representation \K{Rep} of a datatype \K{a} is a type synonym for a \K{SOP I} of \K{a}'s code.
+The functions, \K{from} and \K{to}, perform a shallow conversion of the datatype topmost layer---they do not recursively translate the constructor arguments.
+
+We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work.
+Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following.
+Therefore, it is worth noting that \K{All} applies a particular constraint (e.g.
+\K{SListI} above) to each member of a list of types.
+The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.
 
@@
 We illustrate the problem through a short example.
-The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \texttt{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
+The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \K{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
 We reimplement this function using the SOP view:
-\texttt{
-\vs
-\indent subterms \ann Generic a \Ra a \ra [a]\\
-\indent subterms t = subtermsNS \$ unSOP \$ from t
-\vs
-\indent subtermsNS \ann NS (NP I) xss \ra [a]\\
-\indent subtermsNS (S ns) = subtermsNS ns\\
-\indent subtermsNS (Z np) = subtermsNP np
-\vs
-\indent subtermsNP \ann \fa a xs. NP I xs \ra [a]\\
-\indent subtermsNP p (I y :* ys)\\
-\indent\s\s | typeEq @a y \hs\hs\ths= witnessEq y : subtermsNP ys\\
-\indent\s\s | otherwise\s\s\s\hs\hs\ths = subtermsNP ys\\
-\indent subtermsNP \_ Nil = []
-\vs
-}
-The function \texttt{subterms} translates the term to its representation, unwrapping the sum of products from \texttt{SOP}, and passes that to the auxiliary function \texttt{subtermsNS}.
-The latter merely traverses the sum and, once reaches the product,  passes it further to \texttt{subtermsNP}.
-
-The algorithm of \texttt{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \texttt{t}, otherwise skipping the element.
+\begin{lstlisting}
+subterms :: Generic a => a -> [a]
+subterms t = subtermsNS $ unSOP $ from t
+
+subtermsNS :: NS (NP I) xss -> [a]
+subtermsNS (S ns) = subtermsNS ns
+subtermsNS (Z np) = subtermsNP np
+
+subtermsNP :: ∀a xs. NP I xs -> [a]
+subtermsNP p (I y :* ys)
+  | typeEq @a y = witnessEq y : subtermsNP ys
+  | otherwise   = subtermsNP ys
+subtermsNP _ Nil = []
+\end{lstlisting}
+The function \K{subterms} translates the term to its representation, unwrapping the sum of products from \K{SOP}, and passes that to the auxiliary function \K{subtermsNS}.
+The latter merely traverses the sum and, once reaches the product,  passes it further to \K{subtermsNP}.
+
+The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \K{t}, otherwise skipping the element.
 We use GHC's \textsf{TypeAppications} extension to pass that type.
@@
 There is no clear path to this at the moment.
-Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \texttt{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
-\texttt{
-\vs
-\indent\textbf{class} Subterms a (xs \ann [*]) \textbf{where}\\
-\indent\s\s subtermsNP \ann NP I xs \ra [a]
-\vs
-\indent\textbf{instance} Subterms a xs \Ra Subterms a (x ': xs) \textbf{where}\\
-\indent\s\s subtermsNP (\_\s\s\s :* xs) \hs\ths= subtermsNP xs\\
-\indent\textbf{instance} \{-\# OVERLAPS \#-\}\\
-\indent\ind\ind\s Subterms a xs \Ra Subterms a (a ': xs) \textbf{where}\\
-\indent\s\s subtermsNP (I x :* xs) = x : subtermsNP xs\\
-\indent\textbf{instance} Subterms a '[] \textbf{where}\\
-\indent\s\s subtermsNP \_ = []
-\vs
-}
-
-To make the whole solution work, we need to propagate the constraints all the way through \texttt{subtermsNS} and \texttt{subterms} signatures:
-\texttt{
-\vs
-\indent subterms\s\s\s \ann\hs (Generic a, All (Subterms a) (Code a))\\
-\indent\ind\ind\s\s\s \Ra a \ra [a]\\
-\indent subtermsNS \ann\hs All (Subterms a) xss\\
-\indent\ind\ind\s\s\s \Ra NS (NP I) xss \ra [a]
-\vspace{0.2cm}
-}
+Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
+\begin{lstlisting}
+class Subterms a (xs :: [*]) where
+  subtermsNP :: NP I xs -> [a]
+
+instance Subterms a xs => Subterms a (x ': xs) where
+  subtermsNP (_   :* xs) = subtermsNP xs
+instance {-# OVERLAPS #-}
+         Subterms a xs => Subterms a (a ': xs) where
+  subtermsNP (I x :* xs) = x : subtermsNP xs
+instance Subterms a '[] where
+  subtermsNP _ = []
+\end{lstlisting}
+
+To make the whole solution work, we need to propagate the constraints all the way through \K{subtermsNS} and \K{subterms} signatures:
+\begin{lstlisting}
+subterms   :: (Generic a, All (Subterms a) (Code a))
+           => a -> [a]
+subtermsNS :: All (Subterms a) xss
+           => NS (NP I) xss -> [a]
+\end{lstlisting}
 
@@
 Under the extension, we can give the following definition of type-level equality:
-\texttt{
-\vs
-\indent\textbf{type family} Equal a x \ann Bool \textbf{where}\\
-\indent\s\s Equal a a = 'True\\
-\indent\s\s Equal a x = 'False
-\vs
-}
+\begin{lstlisting}
+type family Equal a x :: Bool where
+  Equal a a = 'True
+  Equal a x = 'False
+\end{lstlisting}
 The equations in a closed type family are matched in a top-to-bottom order. Since the order is fixed, the overlapping equations here cannot be used to define unsound type-level equations.
 
-\subsection{Solution to \texttt{subtermsNP} revised}
+\subsection{Solution to \K{subtermsNP} revised}
 \label{subsec:solve-subtermnp}
@@
 We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
-\texttt{
-\vs
-\indent\textbf{class} Proof (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
-\indent\s\s witnessEq \ann b \ra Maybe a
-\vs
-\indent\textbf{instance} Proof 'False a b \textbf{where}\\
-\indent\s\s witnessEq \_ \hs\nths= Nothing\\
-\indent\textbf{instance} Proof 'True\s\s a a \textbf{where}\\
-\indent\s\s witnessEq\s\s\s = Just
-\vs
-}
+\begin{lstlisting}
+class Proof (eq :: Bool) (a :: *) (b :: *) where
+  witnessEq :: b -> Maybe a
+
+instance Proof 'False a b where
+  witnessEq _ = Nothing
+instance Proof 'True  a a where
+  witnessEq   = Just
+\end{lstlisting}
 
 For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
-This can be done by means of the \texttt{All} combinator and partially applied auxiliary type class \texttt{ProofEq}, which abbreviates the heavy-weighted interface of \texttt{Proof}:
-\texttt{
-\vs
-\indent\textbf{class}\ind Proof (Equal a b) a b \Ra ProofEq a b\\
-\indent\textbf{instance} Proof (Equal a b) a b \Ra ProofEq a b
-\vs
-}
-Resulting implementation of \texttt{subtermsNP} resembles our first definition given in the previous section:
-\texttt{
-\vs
-\indent subtermsNP \ann \fa a xs. All (ProofEq a) xs \Ra NP I xs \ra [a]\\
-\indent subtermsNP (I (y \ann x) :* ys)\\
-\indent\s\s = case witnessEq @(Equal a x) y of\\
-\indent\ind\s\s Just t\s\s \ra t : subtermsNP ys\\
-\indent\ind\s\s  Nothing \ra subtermsNP  ys\\
-\indent subtermsNP \_ Nil = []
-\vs
-}
+This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{Proof}:
+\begin{lstlisting}
+class    Proof (Equal a b) a b => ProofEq a b
+instance Proof (Equal a b) a b => ProofEq a b
+\end{lstlisting}
+Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
+\begin{lstlisting}
+subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
+subtermsNP (I (y :: x) :* ys)
+  = case witnessEq @(Equal a x) y of
+      Just t  -> t : subtermsNP ys
+      Nothing -> subtermsNP  ys
+subtermsNP _ Nil = []
+\end{lstlisting}
 
@@
 
-To complete the solution of the problem, the \texttt{ProofEq} constraint must be added to the \texttt{subterms} and \texttt{subtermsNS} declarations as well.
-
-In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \texttt{subterms}' task.
+To complete the solution of the problem, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.
+
+In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \K{subterms}' task.
 We give another example showing how to adapt our idiom to different scenarios in the following subsection.
@@
 
-The function \texttt{show} is a common example of useful functions that traverse a datatype's recursive structure.
+The function \K{show} is a common example of useful functions that traverse a datatype's recursive structure.
 It is known that this function can be defined in a generic way for an arbitrary datatype.
-De Vries and L\"{o}h define generic function \texttt{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view.
-We follow their implementation of \texttt{gshow} for the most part, but improve it in respect of handling recursion.
-The original \texttt{gshow} yields to the standard \texttt{show} generated through \texttt{\textbf{deriving} Show}, because it does not consult with recursion points to place parentheses.
+De Vries and L\"{o}h define generic function \K{gshow} in the \textsf{basic-sop} package~\cite{basic-sop} based on the SOP view.
+We follow their implementation of \K{gshow} for the most part, but improve it in respect of handling recursion.
+The original \K{gshow} yields to the standard \K{show} generated through \K{deriving Show}, because it does not consult with recursion points to place parentheses.
 We eliminate this drawback.
@@
 As before, we consider two cases.
-In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \texttt{Show}, and invoke its \texttt{show} function.
-Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \texttt{gshow}.
+In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \K{Show}, and invoke its \K{show} function.
+Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \K{gshow}.
 Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
-\texttt{
-\vs
-\indent\textbf{class} CaseShow (eq \ann Bool) (a \ann *) (b \ann *) \textbf{where}\\
-\indent\s\s caseShow' \ann b \ra String
-\vs
-\indent\textbf{instance} Show\s\s b \Ra CaseShow 'False a b \textbf{where}\\
-\indent\s\s caseShow'\s\s\s = show\\
-\indent\textbf{instance} GShow a \Ra CaseShow 'True\s\s a a \textbf{where}\\
-\indent\s\s caseShow' t = "(" ++ gshow t ++ ")"
-\vs
-}
-We provide a synonym for the \texttt{CaseShow (Equal a b) a b} instance, which we call \texttt{CaseRecShow}, as before with \texttt{ProofEq}; likewise a synonym for the matching function:
-\texttt{
-\vs
-\indent caseShow \ann \fa a b. CaseRecShow a b \Ra b \ra String\\
-\indent caseShow t = caseShow' @(Equal a b) @a t
-\vspace{0.2cm}
-}
-
-The resulting function \texttt{gshow} is a subject of a number of constraints abbreviated by a \texttt{GShow} synonym:
-\texttt{
-\vs
-\indent\textbf{type} GShow a = (Generic a, HasDatatypeInfo a,\\
-\indent\ind\ind\ind\ind All2 (CaseRecShow a) (Code a))
-\vs
-\indent gshow \ann \fa a. GShow a \Ra a \ra String
-\vs
-}
-The function \texttt{gshow} employs metadata provided by \textsf{generics-sop}'s class \texttt{Has\-Datatype\-Info} to show the names of a datatype constructor and its record fields.
+\begin{lstlisting}
+class CaseShow (eq :: Bool) (a :: *) (b :: *) where
+  caseShow' :: b -> String
+
+instance Show  b => CaseShow 'False a b where
+  caseShow'   = show
+instance GShow a => CaseShow 'True  a a where
+  caseShow' t = "(" ++ gshow t ++ ")"
+\end{lstlisting}
+We provide a synonym for the \K{CaseShow (Equal a b) a b} instance, which we call \K{CaseRecShow}, as before with \K{ProofEq}; likewise a synonym for the matching function:
+\begin{lstlisting}
+caseShow :: ∀a b. CaseRecShow a b => b -> String
+caseShow t = caseShow' @(Equal a b) @a t
+\end{lstlisting}
+
+The resulting function \K{gshow} is a subject of a number of constraints abbreviated by a \K{GShow} synonym:
+\begin{lstlisting}
+type GShow a = (Generic a, HasDatatypeInfo a,
+                All2 (CaseRecShow a) (Code a))
+
+gshow :: ∀a. GShow a => a -> String
+\end{lstlisting}
+The function \K{gshow} employs meta-information provided by \textsf{generics-sop}'s class \K{HasDatatypeInfo} to show the names of a datatype constructor and its record fields.
 The \textsf{generics-sop} library is able to derive this metadata automatically.
-The function is also constrained by \texttt{CaseRecShow} with the \texttt{All2} combinator that is an analogue of \texttt{All} for a list of lists of types.
-
-We define \texttt{gshow} mutually recursive with \texttt{caseShow}.
-The full implemenentation of the function \texttt{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
-
-The function \texttt{gshow} can now be used to generically show data---for example, a value of type \texttt{Tree~Bool}; note that \texttt{Tree~a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \texttt{Generic} and \texttt{HasDatatypeInfo}.
-\texttt{
-\vs
-\indent *Main> \textbf{let} tree = Node (Leaf True) (Leaf False)\\
-\indent *Main> gshow tree
-\vs
-\indent "Node (Leaf True) (Leaf False)"
-\vs
-}
-Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \texttt{gshow} and some datatype \texttt{T}:
-\texttt{
-\vs
-\indent\textbf{instance} Show T \textbf{where}\\
-\indent\s\s show = gshow
-\vs
-}
-This is a consequence of \texttt{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \texttt{Show} constraint for all knots in the datatype structure.
+The function is also constrained by \K{CaseRecShow} with the \K{All2} combinator that is an analogue of \K{All} for a list of lists of types.
+
+We define \K{gshow} mutually recursive with \K{caseShow}.
+The full implemenentation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
+
+The function \K{gshow} can now be used to generically show data---for example, a value of type \K{Tree Bool}; note that \K{Tree a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \K{Generic} and \K{HasDatatypeInfo}.
+\begin{lstlisting}
+*Main> let tree = Node (Leaf True) (Leaf False)
+*Main> gshow tree
+
+"Node (Leaf True) (Leaf False)"
+\end{lstlisting}
+Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}:
+\begin{lstlisting}
+instance Show T where
+  show = gshow
+\end{lstlisting}
+This is a consequence of \K{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \K{Show} constraint for all knots in the datatype structure.
 
@@
 Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g.
-\texttt{Tree~a} from Section~\ref{sec:sop-view}).
+\K{Tree a} from Section~\ref{sec:sop-view}).
 We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent.
@@
 
-Assume we have a polymorphically recursive datatype \texttt{PolyRec~a} defined in terms of a type family \texttt{Poly}:
-\texttt{
-\vs
-\indent data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
-\vs
-\indent type family Poly a where\\
-\indent \ind  Poly Bool = Char\\
-\indent \ind  Poly Char = Bool\\
-\indent \ind  Poly a\ind = a
-\vspace{0.2cm}
-}
-
-For managing polymorphic recursion in this datatype, we can write an analogue of the \texttt{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types.
-Since any datatype with type parameters \texttt{f a b c \ds} has kind \texttt{* \ra (* \ra (* \ra \ds))}, a \texttt{PolyEq} type family can be defined thus:
-\texttt{
-\vs
-\indent type family PolyEq (a \ann k) (x \ann k) \ann Bool where\\
-\indent \ind  PolyEq (f a) (g b) = PolyEq f g\\
-\indent \ind  PolyEq a\ind\s a\ind\s = 'True\\
-\indent \ind  PolyEq a\ind\s x\ind\s = 'False
-\vspace{0.2cm}
-}
-
-The function \texttt{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \texttt{Equal}.
+Assume we have a polymorphically recursive datatype \K{PolyRec a} defined in terms of a type family \K{Poly}:
+\begin{lstlisting}
+data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
+
+type family Poly a where
+  Poly Bool = Char
+  Poly Char = Bool
+  Poly a    = a
+\end{lstlisting}
+
+For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types.
+Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
+\begin{lstlisting}
+type family PolyEq (a :: k) (x :: k) :: Bool where
+  PolyEq (f a) (g b) = PolyEq f g
+  PolyEq a     a     = 'True
+  PolyEq a     x     = 'False
+\end{lstlisting}
+
+The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}.
 The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
-\texttt{
-\vs
-\indent instance Show \s b \Ra CaseShow 'False a b where\\
-\indent \ind    caseShow' \s\s = show\\
-\indent instance GShow b \Ra CaseShow 'True \s a b where\\
-\indent \ind    caseShow' t = "(" ++ gshow t ++ ")"
-\vspace{0.2cm}
-}
-
-Unfortunately, this approach, albeit working well for datatypes defined like \texttt{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
-\texttt{
-\vs
-\indent data Nested a = Epsilon | Nest a (Nested [a])
-\vs
-}
-The culprit is the constraint \texttt{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \texttt{GShow b} instance context above.
+\begin{lstlisting}
+instance Show  b => CaseShow 'False a b where
+  caseShow'   = show
+instance GShow b => CaseShow 'True  a b where
+  caseShow' t = "(" ++ gshow t ++ ")"
+\end{lstlisting}
+
+Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
+\begin{lstlisting}
+data Nested a = Epsilon | Nest a (Nested [a])
+\end{lstlisting}
+The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow b} instance context above.
 
@@
 \begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\textbf{Movement functions}
-\vs
-\tt goUp\ind \ann Loc a fam c \ra Maybe (Loc a fam c)\\
-\tt goDown\s\s \ann Loc a fam c \ra Maybe (Loc a fam c)\\
-\tt goLeft\s\s \ann Loc a fam c \ra Maybe (Loc a fam c)\\
-\tt goRight \ann Loc a fam c \ra Maybe (Loc a fam c)
-\vspace{0.1cm}
-\vs
-\textbf{Starting navigation}
-\vs
-\tt enter\s\s\s \ann\hs \fa fam c a. (Generic a, In a fam, Zipper a fam c)\\
-\tt\ind\ind \Ra a \ra Loc a fam c
-\vspace{0.1cm}
-\vs
-\textbf{Ending navigation}
-\vs
-\tt leave\s\s\s \ann Loc a fam c \ra a
-\vspace{0.1cm}
-\vs
-\textbf{Updating}
-\vs
-\tt update\s\s \ann (\fa b. c b \Ra b \ra b) \ra Loc a fam c \ra Loc a fam c
-\end{tabular}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+(/\textbf{\textrm{Movement functions}}/)
+
+goUp    :: Loc a fam c -> Maybe (Loc a fam c)
+goDown  :: Loc a fam c -> Maybe (Loc a fam c)
+goLeft  :: Loc a fam c -> Maybe (Loc a fam c)
+goRight :: Loc a fam c -> Maybe (Loc a fam c)
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{Starting navigation}}/)
+
+enter   :: ∀fam c a. (Generic a, In a fam, Zipper a fam c)
+        => a -> Loc a fam c
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{Ending navigation}}/)
+
+leave   :: Loc a fam c -> a
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{Updating}}/)
+
+update  :: (∀b. c b => b -> b) -> Loc a fam c -> Loc a fam c
+\end{lstlisting}}
+\end{CenteredBox}
 \caption{Generic zipper interface.}
@@
 
-The functions \texttt{goUp}, \texttt{goDown}, \texttt{goLeft}, and \texttt{goRight} produce a location with the focus moved \emph{up} to the parent of the focal subtree, \emph{down} to its leftmost child, \emph{left} and \emph{right} to the left and right sibling, respectively, if it is possible.
-A movement may fail, as specified by the \texttt{Maybe} monad, if we cannot go further in a chosen direction.
-Navigation in a tree starts at the root, and the type variable \texttt{a} refers to the root type that remains the same during the navigation, while the type in the focus of the location may vary and is one of the types in a type list \texttt{fam}.
-
-The function signature of \texttt{enter} specifies the constraints necessary to begin navigation in a structure.
-Firstly, a datatype of the structure needs to have the \texttt{Generic} representation.
-Secondly, the \texttt{In} constraint checks if type \texttt{a} is a member of a type family \texttt{fam}.
-Thirdly, the \texttt{Zipper} constraint collects specific constraints that refer to the implementation of movement operations.
+The functions \K{goUp}, \K{goDown}, \K{goLeft}, and \K{goRight} produce a location with the focus moved \emph{up} to the parent of the focal subtree, \emph{down} to its leftmost child, \emph{left} and \emph{right} to the left and right sibling, respectively, if it is possible.
+A movement may fail, as specified by the \K{Maybe} monad, if we cannot go further in a chosen direction.
+Navigation in a tree starts at the root, and the type variable \K{a} refers to the root type that remains the same during the navigation, while the type in the focus of the location may vary and is one of the types in a type list \K{fam}.
+
+The function signature of \K{enter} specifies the constraints necessary to begin navigation in a structure.
+Firstly, a datatype of the structure needs to have the \K{Generic} representation.
+Secondly, the \K{In} constraint checks if type \K{a} is a member of a type family \K{fam}.
+Thirdly, the \K{Zipper} constraint collects specific constraints that refer to the implementation of movement operations.
 Note that the universal quantifier here sets the instantiation order of the type variables for type applications that will be a part of our usage pattern for the zipper.
 
-The \texttt{leave} function ends navigation moving up to the root and returns its modified value.
-
-The \texttt{update} function modifies the focal subtree with a given constrained function.
-The type in focus is existentially quantified inside \texttt{Loc} and should satisfy the constraint \texttt{c}.
-The structure of \texttt{Loc} (shown in Section~\ref{ss:locs}) guarantees that the constraint holds for all types in the family \texttt{fam} and, therefore, for all recursive nodes that can be in focus, hence \texttt{update} can always be applied.
+The \K{leave} function ends navigation moving up to the root and returns its modified value.
+
+The \K{update} function modifies the focal subtree with a given constrained function.
+The type in focus is existentially quantified inside \K{Loc} and should satisfy the constraint \K{c}.
+The structure of \K{Loc} (shown in Section~\ref{ss:locs}) guarantees that the constraint holds for all types in the family \K{fam} and, therefore, for all recursive nodes that can be in focus, hence \K{update} can always be applied.
 
@@
 Define a pair of mutually recursive datatypes for a rose tree and a forest, where the forest is a list of trees, and the tree is defined as a value in a node and a forest of its children:
-\texttt{
-\vs
-\indent data RoseTree a = RTree a (Forest a)
-\vs
-\indent data Forest\s\s\s a = Empty | Forest (RoseTree a) (Forest a)
-\vspace{0.2cm}
-}
+\begin{lstlisting}
+data RoseTree a = RTree a (Forest a)
+
+data Forest   a = Empty | Forest (RoseTree a) (Forest a)
+\end{lstlisting}
 
 Updating the trees can be done through a class:
-\texttt{
-\vs
-\indent class UpdateTree a b where\\
-\indent\s\s replaceBy \ann RoseTree a \ra b \ra b\\
-\indent\s\s replaceBy \_\s\s\s = id\\
-\indent instance UpdateTree a (RoseTree a) where\\
-\indent\s\s replaceBy t \_ = t\\
-\indent instance UpdateTree a (Forest a)
-\vs
-}
+\begin{lstlisting}
+class UpdateTree a b where
+  replaceBy :: RoseTree a -> b -> b
+  replaceBy _   = id
+instance UpdateTree a (RoseTree a) where
+  replaceBy t _ = t
+instance UpdateTree a (Forest a)
+\end{lstlisting}
 This replaces a tree node with a given tree, and, for the forests, this leaves the nodes untouched.
 
-For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \comp~and Kleisli composition \kleisli.
-The latter is instantiated with the \texttt{Maybe} monad that wraps the result type of the movement functions.
-\texttt{
-\vs
-\indent (\comp) \ann (a \ra b) \ra (b \ra c) \ra (a \ra c)\\
-\indent (\kleisli) \ann Monad m \Ra (a \ra m b) \ra (b \ra m c) \ra (a \ra m c)
-\vspace{0.2cm}
-}
+For chaining moves and edits, we can follow Yakushev~et~al.~\cite{MuRec2009} and employ  the flipped function composition \K{>>>}~and Kleisli composition \K{>=>}.
+The latter is instantiated with the \K{Maybe} monad that wraps the result type of the movement functions.
+\begin{lstlisting}
+(>>>) :: (a -> b) -> (b -> c) -> (a -> c)
+(>=>) :: Monad m => (a -> m b) -> (b -> m c) -> (a -> m c)
+\end{lstlisting}
 
 The type family we need to run the example is defined as follows:
-\texttt{
-\vs
-\indent type TreeFam a = '[RoseTree a, Forest a]
-\vs
-}
+\begin{lstlisting}
+type TreeFam a = '[RoseTree a, Forest a]
+\end{lstlisting}
 
 Finally, we can use zipper operations with our updating function to traverse and replace a part of a forest:
-\texttt{
-\vs
-\indent *Main> let forest\\
-\indent\ind\ind\s = Forest (RTree 'a' \$ Forest (RTree 'b' Empty) Empty)\\
-\indent\ind\ind\ind\ind\s\s (Forest (RTree 'x' Empty) Empty)
-\vs
-\indent *Main> let t = RoseTree 'c' Empty
-\vs
-\indent *Main> enter @(TreeFam Char) @(UpdateTree Char)\\
-\indent\ind\ind\s\comps goDown \kleislis goRight \kleislis goDown\\
-\indent\ind\ind\s\kleislis update (replaceBy t)\\
-\indent\ind\ind\s\comps leave \comps return \$ forest
-\vs
-}
+\begin{lstlisting}
+*Main> let forest
+         = Forest (RTree 'a' $ Forest (RTree 'b' Empty) Empty)
+                  (Forest (RTree 'x' Empty) Empty)
+
+*Main> let t = RoseTree 'c' Empty
+
+*Main> enter @(TreeFam Char) @(UpdateTree Char)
+         >>> goDown >=> goRight >=> goDown
+         >=> update (replaceBy t)
+         >>> leave >>> return $ forest
+\end{lstlisting}
 This yields the following result:
-\texttt{
-\vs
-\indent Forest (RTree 'a' \$ Forest (RTree 'b' Empty) Empty)\\
-\indent\ind\s\s\s (Forest (RTree 'c' Empty) Empty)
-\vspace{0.2cm}
-}
+\begin{lstlisting}
+Forest (RTree 'a' $ Forest (RTree 'b' Empty) Empty)
+       (Forest (RTree 'c' Empty) Empty)
+\end{lstlisting}
 
 Our zipper applies to regular datatypes as well.
-In that case, \texttt{fam} list shall contain a single element.
+In that case, \K{fam} list shall contain a single element.
 Generally, the interface is flexible enough to allow us to check in any collection of types we are interested in during traversal.
@@
 The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
-\texttt{
-\vs
-\indent data Loc (r \ann *) (fam \ann [*]) (c \ann * \ra Constraint) where\\
-\indent\s\s Loc \ann\hs Focus r a fam c \ra Contexts r a fam c\\
-\indent\ind\s\s\ra Loc r fam c
-\vs
-}
-The type parameters \texttt{r}, \texttt{fam}, and \texttt{c} in \texttt{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively.
-Also, the single constructor is existentially quantified over one more type variable, \texttt{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure.
-We discuss both term parameters of the constructor of \texttt{Loc} in detail below.
+\begin{lstlisting}
+data Loc (r :: *) (fam :: [*]) (c :: * -> Constraint) where
+  Loc :: Focus r a fam c -> Contexts r a fam c
+      -> Loc r fam c
+\end{lstlisting}
+The type parameters \K{r}, \K{fam}, and \K{c} in \K{Loc} correspond to the root type of the tree, the list of types of nodes to visit, and a constraint imposing restrictions on the types in the list, respectively.
+Also, the single constructor is existentially quantified over one more type variable, \K{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure.
+We discuss both term parameters of the constructor of \K{Loc} in detail below.
 
@@
 
-The subtree in focus is wrapped by the \texttt{Focus} datatype. 
+The subtree in focus is wrapped by the \K{Focus} datatype. 
 The wrapper incapsulates the proofs about a number of important properties of a focus.
-\texttt{
-\vs
-\indent data Focus (r \ann *) (a \ann *) (fam \ann [*])\\
-\indent\ind\ind\s\s\s (c \ann * \ra Constraint) where\\
-\indent\s\s Focus \ann\hs (Generic b, In b fam, ZipperI r a b fam c)\\
-\indent\ind\ind\Ra b \ra Focus r a fam c
-\vs
-}
-Existential type variable \texttt{b} represents the type of a focus.
-We apply a number of predicates to \texttt{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus.
-Firstly, the type of a focus should have the \texttt{Generic} representation.
-Secondly, it should live \texttt{In} the list of types we are going to visit.
-Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \texttt{ZipperI} predicate.
-In particular, the predicate ensures that \texttt{a} is the type of the parent for the focus in the structure under consideration.
-
-We implement the \texttt{In} constraint by means of a type family \texttt{InFam} exactly along the lines of the \texttt{Equal} type family defined in the beginning of Section~\ref{sec:handling-recursion}.
+\begin{lstlisting}
+data Focus (r :: *) (a :: *) (fam :: [*])
+           (c :: * -> Constraint) where
+ Focus :: (Generic b, In b fam, ZipperI r a b fam c)
+       => b -> Focus r a fam c
+\end{lstlisting}
+Existential type variable \K{b} represents the type of a focus.
+We apply a number of predicates to \K{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus.
+Firstly, the type of a focus should have the \K{Generic} representation.
+Secondly, it should live \K{In} the list of types we are going to visit.
+Lastly, it ought to satisfy the set of constraints for the whole zipper interface captured by the \K{ZipperI} predicate.
+In particular, the predicate ensures that \K{a} is the type of the parent for the focus in the structure under consideration.
+
+We implement the \K{In} constraint by means of a type family \K{InFam} exactly along the lines of the \K{Equal} type family defined in the beginning of Section~\ref{sec:handling-recursion}. 
 %~ The following checks a type for membership of a type-level list that we use to express a family:
-%~ \texttt{
+%~ \K{
 %~ \vs
-%~ \indent type family InFam (a \ann k) (fam \ann [k]) \ann Bool where\\
+%~ \indent type family InFam (a :: k) (fam :: [k]) :: Bool where\\
 %~ \indent\s\s InFam a (a ': fam) = 'True\\
@@
 %~ }
-%~ The \texttt{In} constraint is defined as
-\texttt{
-\vs
-\indent type In a fam = InFam a fam \tteq 'True\\
-\vs
-}
-The definition of \texttt{InFam} is omitted as a boring one.
-
-One last missing piece for managing focuses is the class \texttt{ProofIn}.
+%~ The \K{In} constraint is defined as
+\begin{lstlisting}
+type In a fam = InFam a fam == 'True
+\end{lstlisting}
+The definition of \K{InFam} is omitted as a boring one.
+
+One last missing piece for managing focuses is the class \K{ProofIn}.
 It provides a proof of membership of a focus type to a family.
 Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
-The definitions of \texttt{ProofIn} and an auxiliary class \texttt{ProofFocus} are given in Figure~\ref{fig:proof-in}.
-%The constraints on the second \texttt{ProofFam} instance, which repeat the same ones from the \texttt{Family} constructor, constitute the proof.
+The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}.
+%The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
 % this is not clear and, I think, can be omitted
@@
 \begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\tt class ProofFocus (inFam \ann Bool) (r \ann *) (a \ann *) (b \ann *)\\
-\tt\ind\ind\ind\ind\s (fam \ann [*]) (c \ann * \ra Constraint) where\\
-\tt\s\s witness \ann b \ra Maybe (Focus r a fam c)
-\vs
-\tt instance ProofFocus 'False r a b fam c where\\
-\tt\s\s witness \_ \hs\nths= Nothing\\
-\tt instance (Generic b, In b fam, ZipperI r a b fam c)\\
-\tt\ind\Ra ProofFocus 'True r a b fam c where\\
-\tt\s\s witness\s\s\s = Just . Focus
-\vs
-\tt class ProofFocus (InFam b fam) r a b fam c\\
-\tt\ind\Ra ProofIn r a b fam c\\
-\tt instance ProofFocus (InFam b fam) r a b fam c\\
-\tt\ind\Ra ProofIn r a b fam c
-\end{tabular}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+class ProofFocus (inFam :: Bool) (r :: *) (a :: *) (b :: *)
+                 (fam :: [*]) (c :: * -> Constraint) where
+  witness :: b -> Maybe (Focus r a fam c)
+
+instance ProofFocus 'False r a b fam c where
+  witness _ = Nothing
+instance (Generic b, In b fam, ZipperI r a b fam c)
+    => ProofFocus 'True r a b fam c where
+  witness   = Just . Focus
+
+class ProofFocus (InFam b fam) r a b fam c
+    => ProofIn r a b fam c
+instance ProofFocus (InFam b fam) r a b fam c
+    => ProofIn r a b fam c
+\end{lstlisting}}
+\end{CenteredBox}
 \caption{Proof of membership of a family of datatypes.}
@@
 
-The context can be expressed as a stack, called \texttt{Contexts}, and each frame, \texttt{Context}, corresponds to the particular node with a hole.
+The context can be expressed as a stack, called \K{Contexts}, and each frame, \K{Context}, corresponds to the particular node with a hole.
 The stack ascends from the focal node keeping its siblings, the siblings of its parent, etc., until it reaches the root node.
 So the stack of contexts, essentially, reflects the track of the movement inside the structure.
-\texttt{
-\vs
-\indent data Contexts (r \ann *) (a \ann *) (fam \ann [*])\\
-\indent\ind\ind\ind\s\s (c \ann * \ra Constraint) where\\
-\indent\s\s CNil \ann\hs Contexts a a fam c\\
-\indent\s\s Ctxs \ann\hs (Generic a, In a fam, ZipperI r x a fam c)\\
-\indent\ind\s\s\s\Ra Context fam a \ra Contexts r x fam c\\
-\indent\ind\s\s\s\ra Contexts r a fam c
-\vs
-}
-The type parameters have the same meaning as for the \texttt{Loc} datatype.
-The \texttt{ZipperI} constraint with the type \texttt{x} of the previous context frame indicates that the constraint for the zipper holds after plugging the focus in the hole.
+\begin{lstlisting}
+data Contexts (r :: *) (a :: *) (fam :: [*])
+              (c :: * -> Constraint) where
+ CNil :: Contexts a a fam c
+ Ctxs :: (Generic a, In a fam, ZipperI r x a fam c)
+      => Context fam a -> Contexts r x fam c
+      -> Contexts r a fam c
+\end{lstlisting}
+The type parameters have the same meaning as for the \K{Loc} datatype.
+The \K{ZipperI} constraint with the type \K{x} of the previous context frame indicates that the constraint for the zipper holds after plugging the focus in the hole.
 Therefore, all the properties can be proved by induction for the focus type when it moves down in the tree adding new contexts onto the stack.
-The \texttt{CNil} constructor for an empty context, with the \texttt{r} and \texttt{a} types being equal, forms the inductive basis in that kind of proof.
-
-Note that the type of the current focus is not reflected in the \texttt{Contexts} datatype.
+The \K{CNil} constructor for an empty context, with the \K{r} and \K{a} types being equal, forms the inductive basis in that kind of proof.
+
+Note that the type of the current focus is not reflected in the \K{Contexts} datatype.
 
@@
 Those functions, defined recursively via type families, provide algebraic operations for lists of types (which we regard as sums and products of types): addition and multiplication.
-Specifically, we define addition \texttt{.++} of two sums of products (SOP) of types, multiplication \texttt{.*} of a SOP by a single type, and multiplication \texttt{.**} of a SOP by a product of types, as shown in Figure~\ref{fig:type-arithmetic}.
+Specifically, we define addition \K{.++} of two sums of products (SOP) of types, multiplication \K{.*} of a SOP by a single type, and multiplication \K{.**} of a SOP by a product of types, as shown in Figure~\ref{fig:type-arithmetic}.
 \begin{figure}[h]
-\centering
-\normalsize
-\begin{tabular}{l}
-\textbf{SOP addition}
-\vs
-\tt type family (.++) (xs \ann [[*]]) (ys \ann [[*]]) \ann [[*]] where\\
-\tt\s\s (x ': xs) .++ ys = x ': (xs .++ ys)\\
-\tt\s\s '[]\ind\s\s\s .++ ys = ys\\
-\tt infixr 6 .++
-\vspace{0.3cm}\\
-\textbf{SOP-by-type multiplication}
-\vs
-\tt type family (.*) (x \ann *) (ys \ann [[*]]) \ann [[*]] where\\
-\tt\s\s x .* (ys ': yss) = (x ': ys) ': (x .* yss)\\
-\tt\s\s x .* '[]\ind\ind\s = '[]\\
-\tt infixr 7 .*
-\vspace{0.3cm}\\
-\textbf{SOP-by-product multiplication}
-\vs
-\tt type family (.**) (xs \ann [*]) (ys \ann [[*]]) \ann [[*]] where\\
-\tt\s\s (x ': xs) .** yss = x .* (xs .** yss)\\
-\tt\s\s '[]\ind\s\s\s .** yss = yss\\
-\tt infixr 7 .**
-\end{tabular}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+(/\textbf{\textrm{SOP addition}}/)
+
+type family (.++) (xs :: [[*]]) (ys :: [[*]]) :: [[*]] where
+  (x ': xs) .++ ys = x ': (xs .++ ys)
+  '[]       .++ ys = ys
+infixr 6 .++
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{SOP-by-type multiplication}}/)
+
+type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
+  x .* (ys ': yss) = (x ': ys) ': (x .* yss)
+  x .* '[]         = '[]
+infixr 7 .*
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{SOP-by-product multiplication}}/)
+
+type family (.**) (xs :: [*]) (ys :: [[*]]) :: [[*]] where
+  (x ': xs) .** yss = x .* (xs .** yss)
+  '[]       .** yss = yss
+infixr 7 .**
+\end{lstlisting}}
+\end{CenteredBox}
 \caption{Algebraic operations on type-level sums and products.}
@@
 The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum.
-Again, kind \texttt{[*]} here denotes products, and \texttt{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers in these definitions becomes more clear if one realizes that an empty sum \texttt{'[] \ann [[*]]} corresponds to $1$, and an empty product \texttt{'[] \ann [*]} corresponds to $0$.
-We also specify, through the \texttt{infixr} declaration, that multiplication has a higher priority than addition.
+Again, kind \K{[*]} here denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers in these definitions becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $1$, and an empty product \K{'[] :: [*]} corresponds to $0$.
+We also specify, through the \K{infixr} declaration, that multiplication has a higher priority than addition.
 
@@
 
-The definition of differentiation, shown in Figure~\ref{fig:diff-prod}, resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \texttt{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
-When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \texttt{Hole}.
+The definition of differentiation, shown in Figure~\ref{fig:diff-prod}, resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \K{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
+When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \K{Hole}.
 Reflecting this type in the context is helpful when we traverse the context representation to plug the hole.
-We use the empty type \texttt{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice.
-The sum \texttt{'[ '[]]} represents a unit type that is exactly $1$.
-We also use type-level \texttt{If} that returns its second argument for \texttt{'True}, and the third one otherwise.
+We use the empty type \K{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice.
+The sum \K{'[ '[]]} represents a unit type that is exactly $1$.
+We also use type-level \K{If} that returns its second argument for \K{'True}, and the third one otherwise.
 We do not give its definition here, as it is straightforward.
 \begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\tt data Hole = Hole\\
-\tt data End
-\vs
-\tt type family DiffProd (fam \ann [*]) (xs \ann [*]) \ann [[*]] where\\
-\tt\s\s DiffProd fam '[]\ind\s\s\s = '[]\\
-\tt\s\s DiffProd fam '[x]\ind\s\s = If (InFam x fam) '[ '[Hole]] '[]\\
-\tt\s\s DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]\ind\s '[]\\
-\tt\s\s DiffProd fam (x ': xs)\\
-\tt\ind = Hole .* xs .** DiffProd fam '[End, x]  .++\\
-\tt\ind\ind\ind\s\s x\s\s .*\s\s DiffProd fam xs
-\end{tabular}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+data Hole = Hole
+data End
+
+type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
+  DiffProd fam '[]       = '[]
+  DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
+  DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]     '[]
+  DiffProd fam (x ': xs)
+    = Hole .* xs .** DiffProd fam '[End, x] .++
+              x  .*  DiffProd fam xs
+\end{lstlisting}}
+\end{CenteredBox}
 \caption{Differentiation of a product of types.}
@@
 The following completes the computation of the context type:
-\texttt{
-\vs
-\indent data ConsN = F | N ConsN | None\\
-\indent\s\s deriving Eq
-\vs
-\indent type family ToContext (n \ann ConsN) (fam \ann [*])\\
-\indent\ind\ind\ind\ind\ind\s\s (code \ann [[*]]) \ann [[*]] where\\
-\indent\s\s ToContext n fam '[] = '[]\\
-\indent\s\s ToContext n fam (xs ': xss)\\
-\indent\ind = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
-\vs
-}
-The type family \texttt{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code.
+\begin{lstlisting}
+data ConsN = F | N ConsN | None
+  deriving Eq
+
+type family ToContext (n :: ConsN) (fam :: [*])
+                      (code :: [[*]]) :: [[*]] where
+  ToContext n fam '[] = '[]
+  ToContext n fam (xs ': xss)
+    = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
+\end{lstlisting}
+The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code.
 Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype.
-We store the index in the datatype \texttt{ConsN} adding that to the head of each product but wrapped by \texttt{Proxy} because we use \texttt{ConsN} promoted to a kind in the type family, while the products contain types of kind \texttt{*}.
-The constructors \texttt{F} and \texttt{N} denote ``first'' and ``next'', respectively, and the special constructor \texttt{None} will be used further to indicate failure of matching indices.
+We store the index in the datatype \K{ConsN} adding that to the head of each product but wrapped by \K{Proxy} because we use \K{ConsN} promoted to a kind in the type family, while the products contain types of kind \K{*}.
+The constructors \K{F} and \K{N} denote ``first'' and ``next'', respectively, and the special constructor \K{None} will be used further to indicate failure of matching indices.
 
 Finally, a context frame has two representations: as a type synonym and as a newtype wrapper:
-\texttt{
-\vs
-\indent type CtxCode  fam a = ToContext 'F fam (Code a)
-\vs
-\indent newtype Context fam a = Ctx \{ctx \ann SOP I (CtxCode fam a)\}
-\vs
-}
+\begin{lstlisting}
+type    CtxCode fam a = ToContext 'F fam (Code a)
+
+newtype Context fam a = Ctx {ctx :: SOP I (CtxCode fam a)}
+\end{lstlisting}
 The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym.
-On the other hand, the \texttt{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.
+On the other hand, the \K{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.
 
@@
 We now can implement the interface functions of the zipper, which we have previously described.
-We only demonstrate the implementation of the \texttt{goDown} function here.
+We only demonstrate the implementation of the \K{goDown} function here.
 This shows the idea of how we can use our idiom for defining the zipper functions, and the source code with the full implementation of the zipper interface is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.
@@
 To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context.
-The following definition of \texttt{goDown} uses two auxiliary functions: \texttt{toFirst} and \texttt{toFirstCtx}.
-\texttt{
-\vs
-\indent goDown \ann Loc a fam c \ra Maybe (Loc a fam c)\\
-\indent goDown (Loc (Focus t) cs)\\
-\indent\s\s = case toFirst t of\\
-\indent\ind\s\s Just t' \ra Just \$ Loc t' (Ctxs (toFirstCtx t) cs)\\
-\indent\ind\s\s\_\ind\s\s\s\hs\ra Nothing
-\vs
-}
-The function \texttt{toFirst} returns its result in the \texttt{Maybe} monad, and may return \texttt{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down.
-The function \texttt{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.
+The following definition of \K{goDown} uses two auxiliary functions: \K{toFirst} and \K{toFirstCtx}.
+\begin{lstlisting}
+goDown :: Loc a fam c -> Maybe (Loc a fam c)
+goDown (Loc (Focus t) cs)
+  = case toFirst t of
+      Just t' -> Just $ Loc t' (Ctxs (toFirstCtx t) cs)
+      _       -> Nothing
+\end{lstlisting}
+The function \K{toFirst} returns its result in the \K{Maybe} monad, and may return \K{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down.
+The function \K{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.
 
 \begin{figure}[t]
-\centering
-\normalsize
-\begin{tabular}{l}
-\tt toFirst \ann\hs\fa fam c r a. (Generic a, ToFirst r a fam c)\\
-\tt\ind\ind\Ra a \ra Maybe (Focus r a fam c)\\
-\tt toFirst t = appToNP @AllProof toFirstNP \$ unSOP \$ from t
-\vspace{0.3cm}\\
-\bf Proof
-\vs
-\tt class\ind All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs\\
-\tt instance All (ProofIn r a fam c) xs \Ra AllProof r a fam c xs
-\vs
-\tt type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
-\vspace{0.3cm}\\
-\bf Processing products
-\vs
-\tt toFirstNP \ann\hs\fa fam c r a xs. All (ProofIn r a fam c) xs\\
-\tt\ind\ind\s\s\Ra NP I xs \ra Maybe (Focus r a fam c)\\
-\tt toFirstNP (I (x \ann b) :* xs)\\
-\tt\s\s = witness @(InFam b fam) x `mplus` toFirstNP xs\\
-\tt toFirstNP Nil = Nothing
-\end{tabular}
-\caption{Implementation of \texttt{toFirst}.}
+\begin{CenteredBox}
+{\lset
+\begin{lstlisting}
+toFirst :: ∀fam c r a. (Generic a, ToFirst r a fam c)
+        => a -> Maybe (Focus r a fam c)
+toFirst t = appToNP @AllProof toFirstNP $ unSOP $ from t
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{Proof}}/)
+
+class    All (ProofIn r a fam c) xs => AllProof r a fam c xs
+instance All (ProofIn r a fam c) xs => AllProof r a fam c xs
+
+type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
+
+(/\vspace{-0.3cm}/)
+(/\textbf{\textrm{Processing products}}/)
+
+toFirstNP :: ∀fam c r a xs. All (ProofIn r a fam c) xs
+          => NP I xs -> Maybe (Focus r a fam c)
+toFirstNP (I (x :: b) :* xs)
+  = witness @(InFam b fam) x `mplus` toFirstNP xs
+toFirstNP Nil = Nothing
+\end{lstlisting}}
+\end{CenteredBox}
+\caption{Implementation of \K{toFirst}.}
 \label{fig:toFirst}
@@
 
-We first implement the function \texttt{toFirst}.
+We first implement the function \K{toFirst}.
 Its full definition is displayed in Figure~\ref{fig:toFirst}.
-The \texttt{toFirst} function uses the higher-oredered function \texttt{appToNP} that unwraps the product from \texttt{NS} and applies the given function to that product.
-The function \texttt{toFirstNP} is defined recursively using the proof we have defined for families: it traverses the product until it finds the first recursive node by means of \texttt{witness} that for the node \texttt{x} of unknown type \texttt{b}, witnesses its membership of the family, or else returns \texttt{Nothing}.
-To provide the proof for the representation code of a datatype, we define the proof for all products in a sum, and pass this proof through explicit type application to \texttt{appToNP} which takes a constrained function.
-The \texttt{appToNP} function is defined similarly to \texttt{subtermsNS} from Section~\ref{sec:recursion-problem}, and we omit its definition here.
+The \K{toFirst} function uses the higher-oredered function \K{appToNP} that unwraps the product from \K{NS} and applies the given function to that product.
+The function \K{toFirstNP} is defined recursively using the proof we have defined for families: it traverses the product until it finds the first recursive node by means of \K{witness} that for the node \K{x} of unknown type \K{b}, witnesses its membership of the family, or else returns \K{Nothing}.
+To provide the proof for the representation code of a datatype, we define the proof for all products in a sum, and pass this proof through explicit type application to \K{appToNP} which takes a constrained function.
+The \K{appToNP} function is defined similarly to \K{subtermsNS} from Section~\ref{sec:recursion-problem}, and we omit its definition here.
 
@@
 
-The definition of \texttt{toFirstCtx} is more complicated, as it performs the computation of the context.
+The definition of \K{toFirstCtx} is more complicated, as it performs the computation of the context.
 The implementation comprises several steps including type- and term-level programming.
@@
 
-At first, for a datatype's given constructor represented by \texttt{NP}, we build its matching constructor of the context.
+At first, for a datatype's given constructor represented by \K{NP}, we build its matching constructor of the context.
 All constructors of the context have the same shape as the constructors of its respective datatype but with a hole at one of points of recursion---we are now computing the context with the first recursive node deleted.
 Assuming that we have the product type for the context computed, we can compute the product by matching on that type:
-\texttt{
-\vs
-\indent class FromFstRec (ys \ann [*]) (xs \ann [*]) where\\
-\indent\s\s fromFstRec \ann NP I xs \ra NP I ys
-\vs
-\indent instance FromFstRec (Hole ': xs) (x ': xs) where\\
-\indent\s\s fromFstRec (\_ \hs\nths:* xs) = I Hole :* xs\\
-\indent instance FromFstRec ys xs\\
-\indent\ind\Ra FromFstRec (x ': ys) (x ': xs) where\\
-\indent\s\s fromFstRec (x :* xs) = x\ind\s\s :* fromFstRec xs\\
-\indent instance FromFstRec '[] '[] where\\
-\indent\s\s fromFstRec \_\ind\ind\s \hs\nths= Nil
-\vs
-}
-Note that the first type parameter in the \texttt{FromFstRec} class is the index type list of the result \texttt{NP}---this order of type variables remains for type application through which we will supply the computed type.
+\begin{lstlisting}
+class FromFstRec (ys :: [*]) (xs :: [*]) where
+  fromFstRec :: NP I xs -> NP I ys
+
+instance FromFstRec (Hole ': xs) (x ': xs) where
+  fromFstRec (_ :* xs) = I Hole :* xs
+instance FromFstRec ys xs
+      => FromFstRec (x ': ys) (x ': xs) where
+  fromFstRec (x :* xs) = x :* fromFstRec xs
+instance FromFstRec '[] '[] where
+  fromFstRec _ = Nil
+\end{lstlisting}
+Note that the first type parameter in the \K{FromFstRec} class is the index type list of the result \K{NP}---this order of type variables remains for type application through which we will supply the computed type.
 
@@
 If we have the index of the chosen constructor for the datatype, the one that we find for its context can be computed as follows:
-\texttt{
-\vs
-\indent type family CtxConsN (xss \ann [[*]]) (n \ann ConsN) \ann ConsN where\\
-\indent\s\s CtxConsN '[]\ind\ind\ind\ind\ind\s\s\s n = 'None\\
-\indent\s\s CtxConsN ((Proxy n \s': xs) ': xss) n = 'F\\
-\indent\s\s CtxConsN ((Proxy n' ': xs) ': xss) n = 'N (CtxConsN xss n)
-\vs
-}
-The list \texttt{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.
+\begin{lstlisting}
+type family CtxConsN (xss :: [[*]]) (n :: ConsN) :: ConsN where
+  CtxConsN '[]                       n = 'None
+  CtxConsN ((Proxy n  ': xs) ': xss) n = 'F
+  CtxConsN ((Proxy n' ': xs) ': xss) n = 'N (CtxConsN xss n)
+\end{lstlisting}
+The list \K{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.
 
@@
 We adapt that for presentation.}:
-\texttt{
-\vs
-\indent injections \ann \fa xs f. SListI xs \Ra NP (Inj f xs) xs
-\vs
-\indent newtype Inj f xs a = Inj \{apInj \ann f a \ra K (NS f xs) a\}
-\vs
-}
-For \texttt{f} instantiated to \texttt{NP I} and \texttt{xs} to be the sum of products reflecting a representation code, \texttt{injections} creates a product of all constructor choices that inject appropriate constructor arguments into each sum.
+\begin{lstlisting}
+injections :: ∀xs f. SListI xs => NP (Inj f xs) xs
+
+newtype Inj f xs a = Inj {apInj :: f a -> K (NS f xs) a}
+\end{lstlisting}
+For \K{f} instantiated to \K{NP I} and \K{xs} to be the sum of products reflecting a representation code, \K{injections} creates a product of all constructor choices that inject appropriate constructor arguments into each sum.
 We can use injections to choose the one that matches the obtained index.
@@
 We now can witness the choice using the proof of type equality we have defined in Section~\ref{sec:handling-recursion}.
-In the following definition, \texttt{f} is supposed to be instantiated to \texttt{Inj (NP I) xss} where \texttt{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \texttt{ConsN} instead of defining a type class function.
-However, when using a singleton, the recursive definition of \texttt{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
-\texttt{
-\vs
-\indent class\ind Proof (Equal a b) (f a) (f b) \Ra ProofF f a b\\
-\indent instance Proof (Equal a b) (f a) (f b) \Ra ProofF f a b
-\vs
-\indent class ConsNInj (n \ann ConsN) (ys \ann [*]) where\\
-\indent\s\s consNInj \ann All (ProofF f ys) xss \Ra NP f xss \ra f ys
-\vs
-\indent instance ConsNInj n xs \Ra ConsNInj ('N n) xs where\\
-\indent\s\s consNInj (\_ :* xss) = consNInj @n xss\\
-\indent\s\s consNInj Nil\ind\s\s\s\hs \hs\ths\tths= impossible\\
-\indent instance ConsNInj 'F xs where\\
-\indent\s\s consNInj ((xs \ann f xs) :* \_)\\
-\indent\ind = fromMaybe impossible \$ witnessEq @(Equal xs ys) xs\\
-\indent\s\s consNInj Nil = impossible\\
-\indent instance ConsNInj 'None xs where\\
-\indent\s\s consNInj \_\s\s\s \hs\nths= impossible
-\vs
-\indent impossible \ann a\\
-\indent impossible =  error "impossible"
-\vs
-}
-As long as the constructor index and product type are computed correctly, the proof should never fail (\texttt{impossible}).
+In the following definition, \K{f} is supposed to be instantiated to \K{Inj (NP I) xss} where \K{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \K{ConsN} instead of defining a type class function.
+However, when using a singleton, the recursive definition of \K{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
+\begin{lstlisting}
+class    Proof (Equal a b) (f a) (f b) => ProofF f a b
+instance Proof (Equal a b) (f a) (f b) => ProofF f a b
+
+class ConsNInj (n :: ConsN) (ys :: [*]) where
+  consNInj :: All (ProofF f ys) xss => NP f xss -> f ys
+
+instance ConsNInj n xs => ConsNInj ('N n) xs where
+  consNInj (_ :* xss) = consNInj @n xss
+  consNInj Nil        = impossible
+instance ConsNInj 'F xs where
+  consNInj ((xs :: f xs) :* _)
+    = fromMaybe impossible $ witnessEq @(Equal xs ys) xs
+  consNInj Nil = impossible
+instance ConsNInj 'None xs where
+  consNInj _   = impossible
+
+impossible :: a
+impossible = error "impossible"
+\end{lstlisting}
+As long as the constructor index and product type are computed correctly, the proof should never fail (\K{impossible}).
 And when it passes, this ensures by type check that the constructor choice for the given product type is faithful.
@@
 Finally, we define a function that applies the injection and returns the constructed context for the given product, representing its constructor arguments, and index of that constructor.
-The type of \texttt{injections} below will be inferred from the return type of the context.
-\texttt{
-\vs
-\indent type AppInj n xs ctx =  (ConsNInj n xs, SListI ctx,\\
-\indent\ind\ind\ind\ind\ind\ind All (ProofF (Inj (NP I) ctx) xs) ctx)
-\vs
-\indent appInjCtx \ann \fa n xs fam a. AppInj n xs (CtxCode fam a)\\
-\indent\ind\ind\s\s \Ra NP I xs \ra Context fam a\\
-\indent appInjCtx np\\
-\indent\s\s = Ctx \$ SOP \$ unK \$ apInj (consNInj @n injections) np
-\vspace{0.2cm}
-}
-
-The definition of \texttt{toFirstCtx} can now be given using the defined functions \texttt{appInjCtx} and \texttt{fromFstRec}.
-The following makes use of a type-level function \texttt{FstRecToHole} to compute the product type by replacing the first recursive occurrence in the given datatype's product list with type \texttt{Hole}.
+The type of \K{injections} below will be inferred from the return type of the context.
+\begin{lstlisting}
+type AppInj n xs ctx = (ConsNInj n xs, SListI ctx,
+                        All (ProofF (Inj (NP I) ctx) xs) ctx)
+
+appInjCtx :: ∀n xs fam a. AppInj n xs (CtxCode fam a)
+          => NP I xs -> Context fam a
+appInjCtx np
+  = Ctx $ SOP $ unK $ apInj (consNInj @n injections) np
+\end{lstlisting}
+
+The definition of \K{toFirstCtx} can now be given using the defined functions \K{appInjCtx} and \K{fromFstRec}.
+The following makes use of a type-level function \K{FstRecToHole} to compute the product type by replacing the first recursive occurrence in the given datatype's product list with type \K{Hole}.
 Its definition is straightforward, and is omitted here.
-\texttt{
-\vs
-\indent toFirstCtx \ann \fa fam c a. (Generic a, ToFirstCtx fam a)\\
-\indent\ind\ind\s\s\s \Ra a \ra Context fam a\\
-\indent toFirstCtx t = toFirstCtxNS @'F \$ unSOP \$ from t
-\vs
-\indent toFirstCtxNS \ann \fa n fam a xss. ToFirstCtx' fam a n xss\\
-\indent\ind\ind\ind\s \Ra NS (NP I) xss \ra Context fam a\\
-\indent toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns\\
-\indent toFirstCtxNS (Z (np \ann NP I xs))\\
-\indent\s\s = appInjCtx @(CtxConsN (CtxCode fam a) n) \$\\
-\indent\ind\ind I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
-\vs
-\indent type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
-\vs
-}
-The \texttt{ToFirstCtx'} constraint here is a complex constraint for use of \texttt{appInjCtx} and \texttt{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints.
+\begin{lstlisting}
+toFirstCtx :: ∀fam c a. (Generic a, ToFirstCtx fam a)
+           => a -> Context fam a
+toFirstCtx t = toFirstCtxNS @'F $ unSOP $ from t
+
+toFirstCtxNS :: ∀n fam a xss. ToFirstCtx' fam a n xss
+             => NS (NP I) xss -> Context fam a
+toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns
+toFirstCtxNS (Z (np :: NP I xs))
+  = appInjCtx @(CtxConsN (CtxCode fam a) n) $
+        I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
+
+type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
+\end{lstlisting}
+The \K{ToFirstCtx'} constraint here is a complex constraint for use of \K{appInjCtx} and \K{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints.
 Its definition is given in the extended version of the paper in Technical Report.

### 2018-08-04 lncs-paper.tex
Address small-scale remarks, some wibble

@@
 
-We feel both existing approaches unsatisfactory and make the following contributions.
+The problem of using overlapping instances was first addressed by Kiselyov~et~al.~\cite{Kiselyov2004}.
+Their technique for avoiding overlap relies on a Haskell 98 extension for functional dependencies.
+The solution proposes two variants of defining a type-level equality predicate, a type class, and then systematically localizes overlap by circumventing it with that predicate.
+The first version of type equality maps types to unique type representations and compares them.
+Its later implementation with type families~\cite{Kiselyov2012} fully eliminates \textsf{OverlappingInstances}.
+Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC.
+The most generic solution for type equality, the second version, again makes use of overlapping instances, however.
+
+Closed type families, today's Haskell extension, has been proposed primarily to obviate the need for overlapping instances.
+But it does not seem to be well-known that they have useful application in the generic programming area.
+
+We follow up the overlap-avoiding idea from the existing approach and make the following contributions.
 \begin{itemize}
@@
 \begin{lstlisting}
-type family Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
+type instance Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
 \end{lstlisting}
@@
 \K{SListI} above) to each member of a list of types.
-The usage of constraints as type arguments is allowed due to \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.
+The usage of constraints as type arguments is allowed due to the \textsf{ConstraintKinds} language extension introducing a dedicated kind \K{Constraint}.
 
@@
 The \textsf{QuickCheck} library~\cite{Claessen2011} for automatic testing of Haskell code defines a helper function \K{subterms} that takes a term and obtains a list of all its immediate subterms that are of the same type as the given term, that is, all the recursive positions in the term structure.
-We reimplement this function using the SOP view:
+In the following, we reimplement this function using the SOP view.
+But first we give a sketch of solution to introduce the idea.
+
+\subsubsection{Non-implementation}
+
+The SOP-based \K{subterms} definition is outlined below:
 \begin{lstlisting}
@@
 subtermsNP p (I y :* ys)
-  | typeEq @a y = witnessEq y : subtermsNP ys
+  | typeOf @a y = castEq y : subtermsNP ys
   | otherwise   = subtermsNP ys
@@
 We use GHC's \textsf{TypeAppications} extension to pass that type.
+
+\subsubsection{Overlap-based implementation}
 
@@
   Equal a a = 'True
-  Equal a x = 'False
+  Equal _ _ = 'False
 \end{lstlisting}
@@
 
-We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality, we can witness the coercion between equal types by defining a type class:
-\begin{lstlisting}
-class Proof (eq :: Bool) (a :: *) (b :: *) where
-  witnessEq :: b -> Maybe a
-
-instance Proof 'False a b where
-  witnessEq _ = Nothing
-instance Proof 'True  a a where
-  witnessEq   = Just
-\end{lstlisting}
+We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
+\begin{lstlisting}
+class DecideEq (eq :: Bool) (a :: *) (b :: *) where
+  decideEq :: Maybe (b :==: a)
+
+instance a == b => DecideEq True  a b where
+  decideEq = Just Refl
+instance          DecideEq False a b where
+  decideEq = Nothing
+\end{lstlisting}
+If the types \K{a} and \K{b} are the same, the \K{:==:} type from \K{Data.Type.Equality} witnesses the equality.
 
 For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
-This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{Proof}:
-\begin{lstlisting}
-class    Proof (Equal a b) a b => ProofEq a b
-instance Proof (Equal a b) a b => ProofEq a b
+This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
+\begin{lstlisting}
+class    DecideEq (Equal a b) a b => ProofEq a b
+instance DecideEq (Equal a b) a b => ProofEq a b
+\end{lstlisting}
+The equality proof can then be employed to provide a type-safe cast between two equal types:
+\begin{lstlisting}
+castEq :: ∀a b. ProofEq a b => b -> Maybe a
+castEq t = (\d -> castWith d t) <$> decideEq @(Equal a b)
 \end{lstlisting}
@@
 subtermsNP (I (y :: x) :* ys)
-  = case witnessEq @(Equal a x) y of
+  = case castEq y of
       Just t  -> t : subtermsNP ys
@@
 
-As a side note, we make use of \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
+As a side note, we make use of the \textsf{ScopedTypeVariables} extension in the definition above, as the type of the element being matched does not appear in the function signature, since it may match an empty list.
 
@@
   PolyEq a     a     = 'True
-  PolyEq a     x     = 'False
+  PolyEq _     _     = 'False
 \end{lstlisting}
@@
 The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}.
-The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type:
-\begin{lstlisting}
-instance Show  b => CaseShow 'False a b where
+The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
+\begin{lstlisting}
+instance Show  a => CaseShow 'False a where
   caseShow'   = show
-instance GShow b => CaseShow 'True  a b where
+instance GShow a => CaseShow 'True  a where
   caseShow' t = "(" ++ gshow t ++ ")"
@@
 
-Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec~a}, becomes unsuitable for nested datatypes, such as one below:
+Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
 \begin{lstlisting}
@@
 
-The classical zipper described by Huet~\cite{Huet1997} can be generically calculated for regular datatypes~\cite{HiJeLo2004}---a subset of datatypes that can be viewed as a least fixed point of some polynomial expression on types.
+The classical zipper described by Huet~\cite{Huet1997} can be generically calculated for regular datatypes~\cite{HiJeLo2004}---all datatypes expressible as polynomial expressions on types.
 Yakushev~et~al.~\cite{MuRec2009} generalize the definition of the generic zipper for an arbitrary family of mutually recursive datatypes.
@@
 The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}.
+In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak form of proof, which is more flexible.
 %The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
@@
   x .* (ys ': yss) = (x ': ys) ': (x .* yss)
-  x .* '[]         = '[]
+  _ .* '[]         = '[]
 infixr 7 .*
@@
 type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
-  DiffProd fam '[]       = '[]
+  DiffProd _   '[]       = '[]
   DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
@@
                       (code :: [[*]]) :: [[*]] where
-  ToContext n fam '[] = '[]
+  ToContext _ _ '[] = '[]
   ToContext n fam (xs ': xss)
@@
 type family CtxConsN (xss :: [[*]]) (n :: ConsN) :: ConsN where
-  CtxConsN '[]                       n = 'None
-  CtxConsN ((Proxy n  ': xs) ': xss) n = 'F
-  CtxConsN ((Proxy n' ': xs) ': xss) n = 'N (CtxConsN xss n)
+  CtxConsN '[]                      _ = 'None
+  CtxConsN ((Proxy n  ': _) ': _  ) n = 'F
+  CtxConsN ((Proxy n' ': _) ': xss) n = 'N (CtxConsN xss n)
 \end{lstlisting}
@@
 
-We now can witness the choice using the proof of type equality we have defined in Section~\ref{sec:handling-recursion}.
+We can witness the choice using a variant of the type-safe cast we have defined in Section~\ref{sec:handling-recursion}, which takes the result of the equality predicate:
+\begin{lstlisting}
+typeCastEq :: ∀eq a b. DecideEq eq a b => b -> Maybe a
+\end{lstlisting}
 In the following definition, \K{f} is supposed to be instantiated to \K{Inj (NP I) xss} where \K{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \K{ConsN} instead of defining a type class function.
@@
 \begin{lstlisting}
-class    Proof (Equal a b) (f a) (f b) => ProofF f a b
-instance Proof (Equal a b) (f a) (f b) => ProofF f a b
+class    DecideEq (Equal a b) (f a) (f b) => ProofF f a b
+instance DecideEq (Equal a b) (f a) (f b) => ProofF f a b
 
@@
   consNInj ((xs :: f xs) :* _)
-    = fromMaybe impossible $ witnessEq @(Equal xs ys) xs
+    = fromMaybe impossible $ typeCastEq @(Equal xs ys) xs
   consNInj Nil = impossible
@@
 Several works show how to define the Zipper~\cite{Huet1997} generically for regular~\cite{HiJeLo2004,McBride2001} and mutually recursive~\cite{MuRec2009} types using fixed-point generic views.
-Adams~\cite{Adams2010} defines a generic zipper for heterogeneous types.
+Adams~\cite{Adams2010} defines a generic zipper for heterogeneous types: a different kind of zipper that can traverse knots of various types, so it does not use the recursive structure.
 

### 2018-08-20 lncs-paper.tex
Address remaining remarks and questions, cut the zipper

@@
 {>>>}{{\comp}}2 {>=>}{{\kleisli}}2
+{_1}{{$_\mathtt{1}$}}1 {_2}{{$_\mathtt{2}$}}1
+{_n}{{$_\mathtt{n}$}}1 {_k}{{$_\mathtt{k}$}}1
+{_Aux}{{\large$_\mathtt{\hs Aux}$}}2
+{__Aux}{{\large$_\mathtt{Aux}$}}2
 }
@@
 
+\newcommand{\hs}{\hspace{0.1cm}}
 \newcommand{\nhs}{\hspace{-0.06cm}}
@@
 The first version of type equality maps types to unique type representations and compares them.
-Its later implementation with type families~\cite{Kiselyov2012} fully eliminates \textsf{OverlappingInstances}.
+That variant and its later implementation with type families~\cite{Kiselyov2012} fully eliminate \textsf{OverlappingInstances}.
 Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC.
@@
 \subsection{Problem with handling recursion}
-\label{sec:recursion-problem}
+\label{subsec:recursion-problem}
 
@@
 
+We describe a straghtforward implementation based on plain pattern matching, for reasons of simplicity.
+However, in Section~\ref{subsec:solve-subterms}, we show a better high-level solution using one of the primary advantages of \textsf{generics-sop}---high-level traversal functions.
+
 The SOP-based \K{subterms} definition is outlined below:
@@
 subterms :: Generic a => a -> [a]
-subterms t = subtermsNS $ unSOP $ from t
+subterms = subtermsNS . unSOP . from
 
@@
 There is no clear path to this at the moment.
-Therefore, we step back (until Section~\ref{subsec:solve-subtermnp}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
+Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
 \begin{lstlisting}
@@
 
-\subsection{Solution to \K{subtermsNP} revised}
-\label{subsec:solve-subtermnp}
+\subsection{Solution to \K{subterms} revised}
+\label{subsec:solve-subterms}
 
-We now return to our running example from Section~\ref{sec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
+We now return to our running example from Section~\ref{subsec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
 \begin{lstlisting}
@@
 For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
-This can be done by means of the \K{All} combinator and partially applied auxiliary type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
+This can be done by means of the \K{All} combinator and partially applied dedicated type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
 \begin{lstlisting}
@@
 \end{lstlisting}
-Resulting implementation of \K{subtermsNP} resembles our first definition given in the previous section:
+
+Below we show two variants of completing the \K{subterms} implementation.
+The low-level version performs pattern matching on the structure of \K{NS} and \K{NP}, as we do in Section~\ref{subsec:recursion-problem}.
+For this variant, we reimplement \K{subtermsNP}.
+The different implementation employes powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}.
+Although, those high-level functions become one extra layer of complexity.
+
+\subsubsection{Low-level implementation}
+
+The low-level definition of \K{subtermsNP} using the type cast resembles its outline given in the previous section:
 \begin{lstlisting}
 subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
-subtermsNP (I (y :: x) :* ys)
-  = case castEq y of
+subtermsNP (I (y :: x) :* ys) =
+  case castEq y of
     Just t  -> t : subtermsNP ys
@@
 
-To complete the solution of the problem, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.
+To finish up, the \K{ProofEq} constraint must be added to the \K{subterms} and \K{subtermsNS} declarations as well.
 
-In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for \K{subterms}' task.
+\subsubsection{High-level implementation}
+
+By taking advantage of \textsf{generics-sop}'s powerful functions for collapsing and mapping, one can define the functions \K{subtermsNS} and \K{subtermsNP} as follows:
+\begin{lstlisting}
+subtermsNS :: ∀a xss. All (AllProofEq a) xss
+           => NS (NP I) xss -> [a]
+subtermsNS = collapse_NS
+           . cmap_NS (Proxy @(AllProofEq a)) subtermsNP
+
+subtermsNP :: ∀a xs. AllProofEq a xs => NP I xs -> K [a] xs
+subtermsNP = K . catMaybes
+               . collapse_NP
+               . cmap_NP (Proxy @(ProofEq a)) (mapIK $ castEq)
+\end{lstlisting}
+This implementation requires also defining \K{AllProofEq} and properly changing the \K{subterms} type annotation.
+
+In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for the task of \K{subterms}.
 We give another example showing how to adapt our idiom to different scenarios in the following subsection.
@@
 We follow their implementation of \K{gshow} for the most part, but improve it in respect of handling recursion.
-The original \K{gshow} yields to the standard \K{show} generated through \K{deriving Show}, because it does not consult with recursion points to place parentheses.
-We eliminate this drawback.
+The example shows, how the better implementation, consulting with recursive positions, affects the usability of the function, obviating boilerplate code.
 
@@
 instance GShow a => CaseShow 'True  a a where
-  caseShow' t = "(" ++ gshow t ++ ")"
+  caseShow' = gshow
 \end{lstlisting}
@@
 
-\section{Polymorphic recursion}
-\label{sec:polymorphic-recursion}
+\subsection{Abstract description of the design pattern}
+\label{subsec:abstract-pattern}
 
-The solution, described in the previous section, is applicable to a range of datatypes that are \emph{monomorphically recursive}.
-Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g.
-\K{Tree a} from Section~\ref{sec:sop-view}).
-We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent.
-It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested datatypes.
+We can continue along the lines of the previous examples to consider the approach separately from concrete ones. One can use the technique abstractedly, following the pattern below:
 
-Assume we have a polymorphically recursive datatype \K{PolyRec a} defined in terms of a type family \K{Poly}:
 \begin{lstlisting}
-data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
+class DispatchRec_Aux (p :: Bool)
+                    (a_1 :: k_1) ... (a_n :: k_n) (b :: *) where
+  dispatchRec_Aux :: b -> X
 
-type family Poly a where
-  Poly Bool = Char
-  Poly Char = Bool
-  Poly a    = a
+instance C_1 ... => DispatchRec_Aux 'False a_1 ... a_n b where
+  dispatchRec_Aux = f_1
+instance C_2 ... => DispatchRec_Aux 'True  a_1 ... a_n b where
+  dispatchRec_Aux = f_2
+
+class    DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
+  => DispatchRec a_1 ... a_n b
+instance DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
+  => DispatchRec a_1 ... a_n b
+
+dispatchRec :: ∀a_1 ... a_n b. DispatchRec a_1 ... a_n b => b -> X
+dispatchRec = dispatchRec_Aux @(P a_1 ... a_k b) @a_1 ... @a_k ... @a_n
 \end{lstlisting}
 
-For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types.
-Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
-\begin{lstlisting}
-type family PolyEq (a :: k) (x :: k) :: Bool where
-  PolyEq (f a) (g b) = PolyEq f g
-  PolyEq a     a     = 'True
-  PolyEq _     _     = 'False
-\end{lstlisting}
-
-The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}.
-The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
-\begin{lstlisting}
-instance Show  a => CaseShow 'False a where
-  caseShow'   = show
-instance GShow a => CaseShow 'True  a where
-  caseShow' t = "(" ++ gshow t ++ ")"
-\end{lstlisting}
-
-Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
-\begin{lstlisting}
-data Nested a = Epsilon | Nest a (Nested [a])
-\end{lstlisting}
-The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow b} instance context above.
+This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec__Aux}.
+The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language.
+The introduction of the associated type class \K{DispatchRec} provides a handy interface for dispatch.
+The dispatch function \K{dispatchRec} decides on one of the functions \K{f_1} and \K{f_2}---depending on whether the predicate \K{P} is true for types \K{a_1}, \dots, \K{a_k}, \K{b}.
 
@@
 Then, we turn to the type-level machinery employed to define locations inside mutually recursive datatypes using the SOP view  (Section~\ref{ss:locs}).
-Finally, we discuss the implementation of the generic zipper interface~--- the functions for manipulating locations (Section~\ref{ss:zip-impl}).
+
+In Technical Report, we discuss the implementation of the generic zipper interface---the functions for manipulating locations.
+The source code with the full implementation of the zipper is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.
 
@@
 Generally, the interface is flexible enough to allow us to check in any collection of types we are interested in during traversal.
-However, we demand an updating operation to be a type class function to distinguish the types of the nodes.
+However, we demand an updating operation to rely on a type class function to distinguish the types of the nodes.
 
@@
 Also, the single constructor is existentially quantified over one more type variable, \K{a}, for we need to store a type of the focus' parent to be able to move up successively in a tree-like structure.
-We discuss both term parameters of the constructor of \K{Loc} in detail below.
+We discuss both the term parameters of the constructor of \K{Loc} in detail below.
 
@@
 In particular, the predicate ensures that \K{a} is the type of the parent for the focus in the structure under consideration.
+Also, it guarantees that \K{b} fulfils the constraint \K{c}.
+Again, we make \K{c} embedded in \K{Loc} by reason of existentiality of \K{b}, which precludes the enforcement of this restriction by means of external \K{All}.
 
@@
 
-One last missing piece for managing focuses is the class \K{ProofIn}.
+One last missing piece for managing focuses is the class \K{ProofFocus}.
 It provides a proof of membership of a focus type to a family.
-Again, this generalizes the proof of type equality from Section~\ref{sec:handling-recursion}.
-The definitions of \K{ProofIn} and an auxiliary class \K{ProofFocus} are given in Figure~\ref{fig:proof-in}.
-In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak form of proof, which is more flexible.
+Again, this generalizes (in a weak form) the proof of type equality from Section~\ref{subsec:solve-subterms}.
+The definitions of \K{ProofFocus} and an auxiliary class \K{ProofFocus__Aux} are given in Figure~\ref{fig:proof-in}.
+In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak kind of proof, which is more flexible.
+We can then make it handy by defining a function \K{castFocus} using \K{castFocus__Aux}, as with \K{castEq} before.
 %The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
@@
 \begin{lstlisting}
-class ProofFocus (inFam :: Bool) (r :: *) (a :: *) (b :: *)
+class ProofFocus_Aux (inFam :: Bool) (r :: *) (a :: *) (b :: *)
                    (fam :: [*]) (c :: * -> Constraint) where
-  witness :: b -> Maybe (Focus r a fam c)
+  castFocus_Aux :: b -> Maybe (Focus r a fam c)
 
-instance ProofFocus 'False r a b fam c where
-  witness _ = Nothing
+instance ProofFocus_Aux 'False r a b fam c where
+  castFocus_Aux _ = Nothing
 instance (Generic b, In b fam, ZipperI r a b fam c)
-    => ProofFocus 'True r a b fam c where
-  witness   = Just . Focus
+      => ProofFocus_Aux 'True  r a b fam c where
+  castFocus_Aux   = Just . Focus
 
-class ProofFocus (InFam b fam) r a b fam c
-    => ProofIn r a b fam c
-instance ProofFocus (InFam b fam) r a b fam c
-    => ProofIn r a b fam c
+class    ProofFocus_Aux (InFam b fam) r a b fam c
+    => ProofFocus r a b fam c
+instance ProofFocus_Aux (InFam b fam) r a b fam c
+    => ProofFocus r a b fam c
 \end{lstlisting}}
@@
 Those functions, defined recursively via type families, provide algebraic operations for lists of types (which we regard as sums and products of types): addition and multiplication.
-Specifically, we define addition \K{.++} of two sums of products (SOP) of types, multiplication \K{.*} of a SOP by a single type, and multiplication \K{.**} of a SOP by a product of types, as shown in Figure~\ref{fig:type-arithmetic}.
-\begin{figure}[h]
-\begin{CenteredBox}
-{\lset
+Specifically, we introduce addition \K{.++} of two sums of products (SOP) of types, multiplication \K{.*} of a SOP by a single type, and multiplication \K{.**} of a SOP by a product of types.
+The definition of the latter uses \K{.*}.
+
+The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum.
+We define multiplication by a type below.
+The definitions for the other type-level operations are similar.
 \begin{lstlisting}
-(/\textbf{\textrm{SOP addition}}/)
-
-type family (.++) (xs :: [[*]]) (ys :: [[*]]) :: [[*]] where
-  (x ': xs) .++ ys = x ': (xs .++ ys)
-  '[]       .++ ys = ys
-infixr 6 .++
-
-(/\vspace{-0.3cm}/)
-(/\textbf{\textrm{SOP-by-type multiplication}}/)
-
 type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
@@
   _ .* '[]         = '[]
-infixr 7 .*
+\end{lstlisting}
 
-(/\vspace{-0.3cm}/)
-(/\textbf{\textrm{SOP-by-product multiplication}}/)
-
-type family (.**) (xs :: [*]) (ys :: [[*]]) :: [[*]] where
-  (x ': xs) .** yss = x .* (xs .** yss)
-  '[]       .** yss = yss
-infixr 7 .**
-\end{lstlisting}}
-\end{CenteredBox}
-\caption{Algebraic operations on type-level sums and products.}
-\label{fig:type-arithmetic}
-\end{figure}
-
-The addition operation just appends two type-level lists of lists (sums of products), multiplication by a type adds the type to the head of each inner list of the sum (here we see multiplication of a product and the distributive property of multiplication over addition, just as in arithmetic of numbers), and multiplication by a product appends the list to the head of each inner product of the sum.
-Again, kind \K{[*]} here denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers in these definitions becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $1$, and an empty product \K{'[] :: [*]} corresponds to $0$.
-We also specify, through the \K{infixr} declaration, that multiplication has a higher priority than addition.
+Again, kind \K{[*]} denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $0$, and an empty product \K{'[] :: [*]} corresponds to $1$.
 
@@
 
-At this point, we can implement differentiation of a product of types and, therefore, the computation of a context frame type.
+At this point, we can implement differentiation of a product of types and, therefore, the computation of a context frame type\footnote{The actual definitions slightly differ from ones, presented in the following.
+We omit some implementation details for simplicity.}.
 
-The definition of differentiation, shown in Figure~\ref{fig:diff-prod}, resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes: the differentiation of the single type reflected by a one-element list here results in $0$ (reflected by \K{'[]}) if that is not in the family and hence is regarded as a constant, otherwise it results in $1$.
-When differentiation gives $1$, it is actually the hole, which we express by defining the unit type \K{Hole}.
-Reflecting this type in the context is helpful when we traverse the context representation to plug the hole.
-We use the empty type \K{End} to distinguish the case when the hole is found at the end of the list, in order not to add that into the result twice.
-The sum \K{'[ '[]]} represents a unit type that is exactly $1$.
+The definition of differentiation resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes:
+\begin{lstlisting}
+type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
+  DiffProd _   '[]       = '[]
+  DiffProd fam '[x]      = If (InFam x fam) '[ '[]] '[]
+  DiffProd fam (x ': xs) =
+    xs .** DiffProd fam '[x] .++ '[x] .** DiffProd fam xs
+\end{lstlisting}
+The differentiation of the single type reflected by a one-element list here results in $0$ reflected by \K{'[]}, if that is not in the family and hence is regarded as a constant.
+Otherwise it results in $1$ represented by the sum \K{'[ '[]]}.
+When differentiation gives $1$, it is actually the hole.
 We also use type-level \K{If} that returns its second argument for \K{'True}, and the third one otherwise.
 We do not give its definition here, as it is straightforward.
-\begin{figure}[t]
-\begin{CenteredBox}
-{\lset
-\begin{lstlisting}
-data Hole = Hole
-data End
-
-type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
-  DiffProd _   '[]       = '[]
-  DiffProd fam '[x]      = If (InFam x fam) '[ '[Hole]] '[]
-  DiffProd fam '[End, x] = If (InFam x fam) '[ '[]]     '[]
-  DiffProd fam (x ': xs)
-    = Hole .* xs .** DiffProd fam '[End, x] .++
-              x  .*  DiffProd fam xs
-\end{lstlisting}}
-\end{CenteredBox}
-\caption{Differentiation of a product of types.}
-\label{fig:diff-prod}
-\end{figure}
 
@@
 \begin{lstlisting}
-data ConsN = F | N ConsN | None
-  deriving Eq
-
-type family ToContext (n :: ConsN) (fam :: [*])
+type family ToContext (fam :: [*])
                       (code :: [[*]]) :: [[*]] where
-  ToContext _ _ '[] = '[]
-  ToContext n fam (xs ': xss)
-    = Proxy n .* DiffProd fam xs .++ ToContext ('N n) fam xss
+  ToContext _   '[]         = '[]
+  ToContext fam (xs ': xss) =
+    DiffProd fam xs .++ ToContext fam xss
 \end{lstlisting}
 The type family \K{ToContext} derives the type of the context of a datatype performing differentiation of a sum on its code.
-Since each product from the code matches a sum of multiple products in the context, it is helpful for each product of the context representation, to keep the index of its matching constructor of the datatype.
-We store the index in the datatype \K{ConsN} adding that to the head of each product but wrapped by \K{Proxy} because we use \K{ConsN} promoted to a kind in the type family, while the products contain types of kind \K{*}.
-The constructors \K{F} and \K{N} denote ``first'' and ``next'', respectively, and the special constructor \K{None} will be used further to indicate failure of matching indices.
 
-Finally, a context frame has two representations: as a type synonym and as a newtype wrapper:
+Finally, the type of a context frame representation is a type synonym:
 \begin{lstlisting}
-type    CtxCode fam a = ToContext 'F fam (Code a)
-
-newtype Context fam a = Ctx {ctx :: SOP I (CtxCode fam a)}
-\end{lstlisting}
-The wrapper allows GHC to perform type inference where it is doomed to fail with a plain type synonym.
-On the other hand, the \K{CtxCode} type comes handy in constraints applied to the generic zipper interface functions.
-
-\subsection{Implementing the zipper interface}
-\label{ss:zip-impl}
-
-We now can implement the interface functions of the zipper, which we have previously described.
-We only demonstrate the implementation of the \K{goDown} function here.
-This shows the idea of how we can use our idiom for defining the zipper functions, and the source code with the full implementation of the zipper interface is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.
-
-To move focus down to the leftmost child of the current focal node in the tree, we should analyze the focal subtree's representation to find its first immediate child, and compute its respective context.
-The following definition of \K{goDown} uses two auxiliary functions: \K{toFirst} and \K{toFirstCtx}.
-\begin{lstlisting}
-goDown :: Loc a fam c -> Maybe (Loc a fam c)
-goDown (Loc (Focus t) cs)
-  = case toFirst t of
-      Just t' -> Just $ Loc t' (Ctxs (toFirstCtx t) cs)
-      _       -> Nothing
-\end{lstlisting}
-The function \K{toFirst} returns its result in the \K{Maybe} monad, and may return \K{Nothing}, if the focal node has no children, that is, we currently focus on the leaf node and cannot go down.
-The function \K{toFirstCtx} should not fail: if we can move, it computes the context that matches the leftmost subtree selected from the focus' children.
-
-\begin{figure}[t]
-\begin{CenteredBox}
-{\lset
-\begin{lstlisting}
-toFirst :: ∀fam c r a. (Generic a, ToFirst r a fam c)
-        => a -> Maybe (Focus r a fam c)
-toFirst t = appToNP @AllProof toFirstNP $ unSOP $ from t
-
-(/\vspace{-0.3cm}/)
-(/\textbf{\textrm{Proof}}/)
-
-class    All (ProofIn r a fam c) xs => AllProof r a fam c xs
-instance All (ProofIn r a fam c) xs => AllProof r a fam c xs
-
-type ToFirst r a fam c = All (AllProof r a fam c) (Code a)
-
-(/\vspace{-0.3cm}/)
-(/\textbf{\textrm{Processing products}}/)
-
-toFirstNP :: ∀fam c r a xs. All (ProofIn r a fam c) xs
-          => NP I xs -> Maybe (Focus r a fam c)
-toFirstNP (I (x :: b) :* xs)
-  = witness @(InFam b fam) x `mplus` toFirstNP xs
-toFirstNP Nil = Nothing
-\end{lstlisting}}
-\end{CenteredBox}
-\caption{Implementation of \K{toFirst}.}
-\label{fig:toFirst}
-\end{figure}
-
-\subsubsection{toFirst}
-
-We first implement the function \K{toFirst}.
-Its full definition is displayed in Figure~\ref{fig:toFirst}.
-The \K{toFirst} function uses the higher-oredered function \K{appToNP} that unwraps the product from \K{NS} and applies the given function to that product.
-The function \K{toFirstNP} is defined recursively using the proof we have defined for families: it traverses the product until it finds the first recursive node by means of \K{witness} that for the node \K{x} of unknown type \K{b}, witnesses its membership of the family, or else returns \K{Nothing}.
-To provide the proof for the representation code of a datatype, we define the proof for all products in a sum, and pass this proof through explicit type application to \K{appToNP} which takes a constrained function.
-The \K{appToNP} function is defined similarly to \K{subtermsNS} from Section~\ref{sec:recursion-problem}, and we omit its definition here.
-
-\subsubsection{toFirstCtx}
-
-The definition of \K{toFirstCtx} is more complicated, as it performs the computation of the context.
-The implementation comprises several steps including type- and term-level programming.
-In the following, we systematically construct the context representation from the given generic representation of a datatype.
-
-At first, for a datatype's given constructor represented by \K{NP}, we build its matching constructor of the context.
-All constructors of the context have the same shape as the constructors of its respective datatype but with a hole at one of points of recursion---we are now computing the context with the first recursive node deleted.
-Assuming that we have the product type for the context computed, we can compute the product by matching on that type:
-\begin{lstlisting}
-class FromFstRec (ys :: [*]) (xs :: [*]) where
-  fromFstRec :: NP I xs -> NP I ys
-
-instance FromFstRec (Hole ': xs) (x ': xs) where
-  fromFstRec (_ :* xs) = I Hole :* xs
-instance FromFstRec ys xs
-      => FromFstRec (x ': ys) (x ': xs) where
-  fromFstRec (x :* xs) = x :* fromFstRec xs
-instance FromFstRec '[] '[] where
-  fromFstRec _ = Nil
-\end{lstlisting}
-Note that the first type parameter in the \K{FromFstRec} class is the index type list of the result \K{NP}---this order of type variables remains for type application through which we will supply the computed type.
-
-Once we have constructed the product, we have to build the sum representing the choice of that product.
-If we have the index of the chosen constructor for the datatype, the one that we find for its context can be computed as follows:
-\begin{lstlisting}
-type family CtxConsN (xss :: [[*]]) (n :: ConsN) :: ConsN where
-  CtxConsN '[]                      _ = 'None
-  CtxConsN ((Proxy n  ': _) ': _  ) n = 'F
-  CtxConsN ((Proxy n' ': _) ': xss) n = 'N (CtxConsN xss n)
-\end{lstlisting}
-The list \K{xss} here is expected to be the context code, which this traverses until the first product storing the constructor index equal to the given one, i.~e., the context's first constructor matching with the chosen constructor of the datatype.
-
-To construct the sum with the computed index and product, we again need a proof to witness that the product is type-consistent with the constructor choice.
-To choose the constructor from the context code, we can adopt \textsf{generics-sop}'s \emph{injections}\footnote{The actual definition of the type of injections in \textsf{generics-sop} slightly differs from this.
-We adapt that for presentation.}:
-\begin{lstlisting}
-injections :: ∀xs f. SListI xs => NP (Inj f xs) xs
-
-newtype Inj f xs a = Inj {apInj :: f a -> K (NS f xs) a}
-\end{lstlisting}
-For \K{f} instantiated to \K{NP I} and \K{xs} to be the sum of products reflecting a representation code, \K{injections} creates a product of all constructor choices that inject appropriate constructor arguments into each sum.
-We can use injections to choose the one that matches the obtained index.
-
-We can witness the choice using a variant of the type-safe cast we have defined in Section~\ref{sec:handling-recursion}, which takes the result of the equality predicate:
-\begin{lstlisting}
-typeCastEq :: ∀eq a b. DecideEq eq a b => b -> Maybe a
-\end{lstlisting}
-In the following definition, \K{f} is supposed to be instantiated to \K{Inj (NP I) xss} where \K{xss} reflects the code of the context\footnote{This might be simplified by using a singleton type for \K{ConsN} instead of defining a type class function.
-However, when using a singleton, the recursive definition of \K{toFirstCtx} we give further requires a constraint on its function signature that leads to a nonterminating computation.}:
-\begin{lstlisting}
-class    DecideEq (Equal a b) (f a) (f b) => ProofF f a b
-instance DecideEq (Equal a b) (f a) (f b) => ProofF f a b
-
-class ConsNInj (n :: ConsN) (ys :: [*]) where
-  consNInj :: All (ProofF f ys) xss => NP f xss -> f ys
-
-instance ConsNInj n xs => ConsNInj ('N n) xs where
-  consNInj (_ :* xss) = consNInj @n xss
-  consNInj Nil        = impossible
-instance ConsNInj 'F xs where
-  consNInj ((xs :: f xs) :* _)
-    = fromMaybe impossible $ typeCastEq @(Equal xs ys) xs
-  consNInj Nil = impossible
-instance ConsNInj 'None xs where
-  consNInj _ = impossible
-
-impossible :: a
-impossible = error "impossible"
-\end{lstlisting}
-As long as the constructor index and product type are computed correctly, the proof should never fail (\K{impossible}).
-And when it passes, this ensures by type check that the constructor choice for the given product type is faithful.
-
-Finally, we define a function that applies the injection and returns the constructed context for the given product, representing its constructor arguments, and index of that constructor.
-The type of \K{injections} below will be inferred from the return type of the context.
-\begin{lstlisting}
-type AppInj n xs ctx = (ConsNInj n xs, SListI ctx,
-                        All (ProofF (Inj (NP I) ctx) xs) ctx)
-
-appInjCtx :: ∀n xs fam a. AppInj n xs (CtxCode fam a)
-          => NP I xs -> Context fam a
-appInjCtx np
-  = Ctx $ SOP $ unK $ apInj (consNInj @n injections) np
+type Context fam a = SOP I (ToContext fam (Code a))
 \end{lstlisting}
 
-The definition of \K{toFirstCtx} can now be given using the defined functions \K{appInjCtx} and \K{fromFstRec}.
-The following makes use of a type-level function \K{FstRecToHole} to compute the product type by replacing the first recursive occurrence in the given datatype's product list with type \K{Hole}.
-Its definition is straightforward, and is omitted here.
+\section{Discussion}
+\label{sec:discussion}
+
+In this section, we give a discussion of some concerns relating to the scope and user-friendliness of the introduced no-overlap technique for handling recursion.
+
+\subsubsection{Polymorphic recursion}
+
+The approach, described in this paper, is applicable to a range of datatypes that are \emph{monomorphically recursive}.
+Any of those datatypes has the same type parameters in the left-hand side of its definition and at its recursion points (e.g.
+\K{Tree a} from Section~\ref{sec:sop-view}).
+We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent.
+It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested~\cite{Bird1998} datatypes.
+
+Assume we have a polymorphically recursive datatype \K{PolyRec a} defined in terms of a type family \K{Poly}:
 \begin{lstlisting}
-toFirstCtx :: ∀fam c a. (Generic a, ToFirstCtx fam a)
-           => a -> Context fam a
-toFirstCtx t = toFirstCtxNS @'F $ unSOP $ from t
+data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
 
-toFirstCtxNS :: ∀n fam a xss. ToFirstCtx' fam a n xss
-             => NS (NP I) xss -> Context fam a
-toFirstCtxNS (S ns) = toFirstCtxNS @('N n) ns
-toFirstCtxNS (Z (np :: NP I xs))
-  = appInjCtx @(CtxConsN (CtxCode fam a) n) $
-        I (Proxy @n) :* fromFstRec @(FstRecToHole fam xs) np
+type family Poly a where
+  Poly Bool = Char
+  Poly Char = Bool
+  Poly a    = a
+\end{lstlisting}
 
-type ToFirstCtx fam a = ToFirstCtx' fam a 'F (Code a)
+For managing polymorphic recursion in this datatype, we can write an analogue of the \K{Equal} type family from Section~\ref{sec:handling-recursion}, which ignores type parameters when checking two polymorphic types.
+Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
+\begin{lstlisting}
+type family PolyEq (a :: k) (x :: k) :: Bool where
+  PolyEq (f a) (g b) = PolyEq f g
+  PolyEq a     a     = 'True
+  PolyEq _     _     = 'False
 \end{lstlisting}
-The \K{ToFirstCtx'} constraint here is a complex constraint for use of \K{appInjCtx} and \K{fromFstRec}, which involves extra type classes and a bit of type-level programming to deduce those particular constraints.
-Its definition is given in the extended version of the paper in Technical Report.
-We omit this here for reasons of space.
 
-We have established the mechanism of translation, which is cumbersome but verifying, via the proof, that the constructed context will have the correct type for any given datatype and family.
-The other functions in the zipper interface can be implemented according to this technique.
+The function \K{gshow} from Section~\ref{subsec:gshow} can be reimplemented by using this type family instead of \K{Equal}.
+The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
+\begin{lstlisting}
+instance Show  a => CaseShow 'False a where
+  caseShow' = show
+instance GShow a => CaseShow 'True  a where
+  caseShow' = gshow
+\end{lstlisting}
+
+Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
+\begin{lstlisting}
+data Nested a = Epsilon | Nest a (Nested [a])
+\end{lstlisting}
+The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow a} instance context above.
+
+Nevertheless, there is a useful example of a generic operation that uses the recursive structure of data, but has not recursive calls in itself---and, therefore, can deal with nested datatypes.
+\emph{Generic generic programming}~\cite{MagLoeh2014} is a technique of defining translations between different generic programming universes.
+In our setting, it is possible to define a translation from \textsf{generics-sop} to another library with an explicit encoding of recursive positions, such as \textsf{regular}~\cite{VanNoort2008}.
+This usually needs only to detect recursion points, because the most of generic views provide shallow conversion functions.
+
+\subsubsection{Usability}
+
+The shown no-overlap technique may involve a lot of multiple constraints on generic functions.
+Those complex constraints introduce certain limitations to the approach and make it difficult in use.
+Especially, making a mistake in type-level programming with constraints which can usually be abbreviated in a code, one can get an embarrassing long error message where multiple constraints are displayed.
+Unfortunately, compound error messages on deducing constraints complicate the use of the technique, particularly for non-specialists in datatype-generic programming, in comparison with the existing approaches to treating recursion.
 
@@
 
-We are grateful to Julia Belyakova, who helped to proof-read certain parts of the paper, 
+We are thankful to Andres L\"{o}h for his helpful recommendations and comments on this paper and to the participants of the TFP symposium for their questions and suggestions.
+We are also grateful to Julia Belyakova, who helped to proof-read certain parts of the paper,
 and to the participants of the Seminar on Programming Languages and Compilers at 

### 2018-08-22 lncs-paper.tex
Change to the abstract description

@@
 subtermsNP :: ∀a xs. NP I xs -> [a]
-subtermsNP p (I y :* ys)
+subtermsNP (I y :* ys)
   | typeOf @a y = castEq y : subtermsNP ys
   | otherwise   = subtermsNP ys
-subtermsNP _ Nil = []
+subtermsNP Nil = []
 \end{lstlisting}
@@
     Nothing -> subtermsNP  ys
-subtermsNP _ Nil = []
+subtermsNP Nil = []
 \end{lstlisting}
@@
 caseShow :: ∀a b. CaseRecShow a b => b -> String
-caseShow t = caseShow' @(Equal a b) @a t
+caseShow = caseShow' @(Equal a b) @a
 \end{lstlisting}
@@
 
-We can continue along the lines of the previous examples to consider the approach separately from concrete ones. One can use the technique abstractedly, following the pattern below:
+We can continue along the lines of the previous examples to consider the approach separately from concrete ones. One can use the technique, following the abstracted pattern below:
 
@@
                     (a_1 :: k_1) ... (a_n :: k_n) (b :: *) where
-  dispatchRec_Aux :: b -> X
+  dispatchRec_Aux :: X b
 
@@
 
-dispatchRec :: ∀a_1 ... a_n b. DispatchRec a_1 ... a_n b => b -> X
+dispatchRec :: ∀a_1 ... a_n b. DispatchRec a_1 ... a_n b => X b
 dispatchRec = dispatchRec_Aux @(P a_1 ... a_k b) @a_1 ... @a_k ... @a_n
@@
 The dispatch function \K{dispatchRec} decides on one of the functions \K{f_1} and \K{f_2}---depending on whether the predicate \K{P} is true for types \K{a_1}, \dots, \K{a_k}, \K{b}.
+
+For consumer functions, such as \K{subterms} and \K{gshow}, the type \K{X b} typically is instantiated with \K{b -> X} for some datatype \K{X}.
+The pattern fits equally well for defining producers, such as \K{garbitrary} from the \textsf{basic-sop} package, where \K{X b} would be instantiated with \K{Gen b}.
 

### 2018-08-22 lncs-paper.tex
Wibble

@@
 
-The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending current element to the result list if its type is the same as of the term \K{t}, otherwise skipping the element.
+The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending the current element to the result list if its type is the same as of the original term, otherwise skipping the element.
 We use GHC's \textsf{TypeAppications} extension to pass that type.
@@
 
-For consumer functions, such as \K{subterms} and \K{gshow}, the type \K{X b} typically is instantiated with \K{b -> X} for some datatype \K{X}.
+For consumer functions, such as \K{subterms} and \K{gshow}, the type \K{X b} typically is instantiated with \K{b -> T} for some datatype \K{T}.
 The pattern fits equally well for defining producers, such as \K{garbitrary} from the \textsf{basic-sop} package, where \K{X b} would be instantiated with \K{Gen b}.

### 2018-08-24 lncs-paper.tex

@@
 Many of the extensively used libraries for datatype-generic programming offer a fixed-point view on datatypes to express their recursive structure.
-However, some approaches, based on sums of products, do not use a fixed point.
-They allow for generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues otherwise.
-A known and unwelcome solution is the use of overlapping instances.
-We present an alternative approach that uses closed type families to eliminate the need of overlap for handling recursion in datatypes.
-Moreover, we show that our idiom allows for families of mutually recursive datatypes.
+However, some other approaches, especially the ones based on sums of products, avoid the fixed point encoding.
+They facilitate implementation of generic functions that do not require to look at the recursive knots in a datatype representation, but raise issues otherwise.
+A widely used and unwelcome solution to the problem resorts to overlapping instances.
+In this paper, we present an alternative approach that uses closed type families to eliminate the need of overlap for handling recursion in datatypes.
+Moreover, we show that this idiom allows for generic programming with families of mutually recursive datatypes.
 
@@
 Closed type families, today's Haskell extension, has been proposed primarily to obviate the need for overlapping instances.
-But it does not seem to be well-known that they have useful application in the generic programming area.
-
-We follow up the overlap-avoiding idea from the existing approach and make the following contributions.
+In the area of generic programming, the extension does not seem to be widely leveraged yet.
+Exploiting this observation, we make the following contributions.
 \begin{itemize}
@@
 
-We believe that our idea is suitable for any sum-of-products approach that do not exploit the fixed point view and thus subject to the problem.
-We choose the \textsf{generics-sop} approach as a case study because it appears to be a widely applicable library and builds on powerful language extensions implemented in GHC.
+We believe that the idea presented is suitable for any sum-of-products approach that does not employ the fixed point view and thus subject to the problem.
+We choose the \textsf{generics-sop} library as a case study because it appears to be a widely applicable library and builds on powerful language extensions implemented in GHC.
 
@@
 
-We describe a straghtforward implementation based on plain pattern matching, for reasons of simplicity.
-However, in Section~\ref{subsec:solve-subterms}, we show a better high-level solution using one of the primary advantages of \textsf{generics-sop}---high-level traversal functions.
-
-The SOP-based \K{subterms} definition is outlined below:
+We outline one possible approach to SOP-based implementation of \K{subterms} below.
+It gives a clear idea of our intention, but so far, we miss the necessary toolbox (in particular, \K{typeOf} / \K{castEq}) to finish it.
 \begin{lstlisting}
@@
 We use GHC's \textsf{TypeAppications} extension to pass that type.
+
+Note that our implementation uses plain pattern matching for reasons of simplicity.
+However, in Section~\ref{subsec:solve-subterms}, we show a higher-level solution using one of the primary advantages of \textsf{generics-sop}---high-level traversal functions.
 
@@
 
-\subsubsection{Type-level Differentiation}
+\subsubsection{Type-Level Differentiation}
 
@@
 We can go further and proceed with a solution for generic functions, which covers some datatypes whose type parameters in each recursive knot may differ from those in its parent.
-It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for nested~\cite{Bird1998} datatypes.
+It turns out, as we will show below, that the solution allows for datatypes with a ``simple'' form of \emph{polymorphic recursion}, but fails to work for \emph{nested datatypes}~\cite{Bird1998}.
 
@@
 \end{lstlisting}
-The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the constraint, because of the \K{GShow a} instance context above.
+The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the \K{GShow} constraint for K{[a]}, K{[[a]]}, and so on ad infinitum.
 
-Nevertheless, there is a useful example of a generic operation that uses the recursive structure of data, but has not recursive calls in itself---and, therefore, can deal with nested datatypes.
-\emph{Generic generic programming}~\cite{MagLoeh2014} is a technique of defining translations between different generic programming universes.
+Still, one can think of problems where a generic operation uses the recursive structure of data, but has no recursive calls---and, therefore, can tolerate nested datatypes.
+An example of this kind of problems is \emph{generic generic programming}~\cite{MagLoeh2014}, a technique of defining translations between different generic programming universes.
 In our setting, it is possible to define a translation from \textsf{generics-sop} to another library with an explicit encoding of recursive positions, such as \textsf{regular}~\cite{VanNoort2008}.
@@
 
-\subsubsection{Usability}
+\subsubsection{Usability and error-reporting}
 
 The shown no-overlap technique may involve a lot of multiple constraints on generic functions.
-Those complex constraints introduce certain limitations to the approach and make it difficult in use.
-Especially, making a mistake in type-level programming with constraints which can usually be abbreviated in a code, one can get an embarrassing long error message where multiple constraints are displayed.
-Unfortunately, compound error messages on deducing constraints complicate the use of the technique, particularly for non-specialists in datatype-generic programming, in comparison with the existing approaches to treating recursion.
+We tend to abbreviate them with type aliases.
+Although convenient when reading and writing code, one may encounter a leak in abstraction by making a mistake in client code.
+In this case, GHC sometimes produces an embarrassingly long error message where multiple underlying constraints are displayed.
+We currently investigate the \K{TypeError} recent mechanism of GHC which allows for user-defined error messages, specifically targeted for type level-heavy computations.
 
@@
 
-We are thankful to Andres L\"{o}h for his helpful recommendations and comments on this paper and to the participants of the TFP symposium for their questions and suggestions.
-We are also grateful to Julia Belyakova, who helped to proof-read certain parts of the paper,
-and to the participants of the Seminar on Programming Languages and Compilers at 
+We are thankful to Andres L\"{o}h for his helpful recommendations and comments on the paper.
+We address some insightful questions and suggestions from participants of the TFP symposium, to whom we are deeply grateful.
+We thank Julia Belyakova, who helped to proof-read certain parts of the paper,
+and the participants of the Seminar on Programming Languages and Compilers at 
 I.I.~Vorovich Institute of Mathematics, Mechanics, and Computer Science (Southern Federal

### 2018-08-24 lncs-paper.tex
Typos

@@
 Moreover, we show that this idiom allows for generic programming with families of mutually recursive datatypes.
-
-\textbf{Category:} full research paper.
-
-\textbf{Studentship:} both authors are students.
 
@@
 \end{lstlisting}
-The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the \K{GShow} constraint for K{[a]}, K{[[a]]}, and so on ad infinitum.
+The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the \K{GShow} constraint for \K{[a]}, \K{[[a]]}, and so on ad infinitum.
 

### 2018-11-08 lncs-paper.tex
code style

@@
 \usepackage{fancybox}
+\usepackage{xcolor}
 
-\lstset{
-basicstyle=\ttfamily,
+\lstdefinestyle{fancy}{
+language=Haskell,
+deletekeywords={Int,Bool,String,Functor,Char,True,False,
+  Maybe,Nothing,Just,Show,show,Monad,catMaybes,const,id},
+morekeywords={family},
+basicstyle=\color{cbasic}\rmfamily,
+identifierstyle=\color{cid}\rmfamily\itshape,
+keywordstyle=\color{ckey}\rmfamily\bfseries,
+xleftmargin=\parindent,
+aboveskip=0.2cm,
+belowskip=0.2cm,
+basewidth=0.48em,
+numberbychapter=false,
+showstringspaces=false,
+moredelim=[is][\color{ccom}\itshape]{/}{/},
+escapeinside={(/}{/)},
+literate={:: }{{$::$}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
+{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\scriptsize\bf\dots}}2
+{>>>}{{\scriptsize\comp}}3 {>=>}{{\scriptsize\kleisli}}3
+{.1}{{$_\mathit{1}$}}1 {.2}{{$_\mathit{2}$}}1
+{.n}{{$_\mathit{n}$}}1 {.k}{{$_\mathit{k}$}}1
+{.Aux}{{\large$_\mathit{\hspace{0,05cm} Aux}$}}3
+{..Aux}{{\large$_\mathit{\hspace{-0,03cm} Aux}$}}3
+{@}{{@\hspace{-0,045cm}}}2
+{_}{{\underline{\hspace{0.45em}}}}1
+{+}{{\scriptsize+}}1
+{.-}{{\nhs-}}1
+{-.}{{-\nhs}}1
+}
+
+\lstdefinestyle{ttstyle}{
+basicstyle=\color{ctt}\ttfamily,
 xleftmargin=\parindent,
@@
 basewidth=0.5em,
-numberbychapter=false,
-escapeinside={(/}{/)},
-literate={::}{{\ann}}2 {=>}{{\Ra}}2 {->}{{\ra}}2
-{∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\ds}}2
-{>>>}{{\comp}}2 {>=>}{{\kleisli}}2
-{_1}{{$_\mathtt{1}$}}1 {_2}{{$_\mathtt{2}$}}1
-{_n}{{$_\mathtt{n}$}}1 {_k}{{$_\mathtt{k}$}}1
-{_Aux}{{\large$_\mathtt{\hs Aux}$}}2
-{__Aux}{{\large$_\mathtt{Aux}$}}2
+literate={>>>}{{>\nhs>\nhs>}}2 {>=>}{{>\nhs=\nhs>}}2
 }
@@
 
+\definecolor{cbasic}{HTML}{5C3566}
+\definecolor{cid}{HTML}{002184}
+\definecolor{ckey}{HTML}{BE6A0D}
+\definecolor{ccom}{HTML}{A0821E}
+\definecolor{ctt}{HTML}{223244}
+
 \newcommand{\hs}{\hspace{0.1cm}}
@@
 
+\newcommand{\tg}{\textgreater}
 \newcommand{\Ra}{$\Rightarrow$}
@@
 \newcommand{\tteq}{{\scriptsize$\thicksim$}}
-\newcommand{\comp}{>\nhs>\nhs>}
-\newcommand{\kleisli}{>\nhs=\nhs>}
+\newcommand{\comp}{\tg\nhs\tg\nhs\tg}
+\newcommand{\kleisli}{\tg\nhs=\nhs\tg}
 \newcommand{\ann}{:\texttt{\nhs}:}
@@
 
-\newcommand{\K}[1]{\lstinline{#1}}
+\newcommand{\K}[1]{\lstinline[style=fancy]{#1}}
 
@@
 The SOP approach expresses the code using the \textsf{DataKinds} extension, with a type family:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy][style=fancy]
 type family Code (a :: *) :: [[*]]
@@
 Consider, for instance, a datatype of binary trees:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data Tree a = Leaf a | Node (Tree a) (Tree a)
@@
 This datatype is isomorphic to the following code:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type instance Code (Tree a) = '[ '[a], '[Tree a, Tree a]]
@@
 {\lset
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data NP (f :: k -> *) (xs :: [k]) where
@@
 Basic instantiations of type parameter \K{f} found in SOP are identity functor \K{I}, that is, a type-level equivalent for \K{id} function, and a constant functor \K{K}, an analogue of \K{const}:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 newtype I (a :: *)          = I {unI :: a}
@@
 If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[a]}. Here is an example value of type \K{NP I}:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 I 5 :* I True :* I 'x' :* Nil :: NP I '[Int, Bool, Char]
@@
 For example, the following chooses the third element of a sum:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 S (S (Z (I 5))) :: NS I '[Char, Bool, Int, Bool]
@@
 With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to} witnessing the isomorphism between a datatype and its generic representation:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type Rep a = SOP I (Code a)
@@
 It gives a clear idea of our intention, but so far, we miss the necessary toolbox (in particular, \K{typeOf} / \K{castEq}) to finish it.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 subterms :: Generic a => a -> [a]
@@
 Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 class Subterms a (xs :: [*]) where
@@
   subtermsNP (_   :* xs) = subtermsNP xs
-instance {-# OVERLAPS #-}
+instance /{.-# OVERLAPS #-.}/
          Subterms a xs => Subterms a (a ': xs) where
@@
 To make the whole solution work, we need to propagate the constraints all the way through \K{subtermsNS} and \K{subterms} signatures:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 subterms   :: (Generic a, All (Subterms a) (Code a))
@@
 Under the extension, we can give the following definition of type-level equality:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type family Equal a x :: Bool where
@@
 We now return to our running example from Section~\ref{subsec:recursion-problem}. With the type equality predicate, we can decide if \K{a == b} by defining a type class:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 class DecideEq (eq :: Bool) (a :: *) (b :: *) where
@@
 This can be done by means of the \K{All} combinator and partially applied dedicated type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 class    DecideEq (Equal a b) a b => ProofEq a b
@@
 The equality proof can then be employed to provide a type-safe cast between two equal types:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 castEq :: ∀a b. ProofEq a b => b -> Maybe a
@@
 The low-level definition of \K{subtermsNP} using the type cast resembles its outline given in the previous section:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 subtermsNP :: ∀a xs. All (ProofEq a) xs => NP I xs -> [a]
@@
 By taking advantage of \textsf{generics-sop}'s powerful functions for collapsing and mapping, one can define the functions \K{subtermsNS} and \K{subtermsNP} as follows:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 subtermsNS :: ∀a xss. All (AllProofEq a) xss
@@
 Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 class CaseShow (eq :: Bool) (a :: *) (b :: *) where
@@
 We provide a synonym for the \K{CaseShow (Equal a b) a b} instance, which we call \K{CaseRecShow}, as before with \K{ProofEq}; likewise a synonym for the matching function:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 caseShow :: ∀a b. CaseRecShow a b => b -> String
@@
 The resulting function \K{gshow} is a subject of a number of constraints abbreviated by a \K{GShow} synonym:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type GShow a = (Generic a, HasDatatypeInfo a,
@@
 The function \K{gshow} can now be used to generically show data---for example, a value of type \K{Tree Bool}; note that \K{Tree a} from Section~\ref{sec:sop-view} is now assumed to be an instance of \K{Generic} and \K{HasDatatypeInfo}.
-\begin{lstlisting}
+\begin{lstlisting}[style=ttstyle]
 *Main> let tree = Node (Leaf True) (Leaf False)
@@
 Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 instance Show T where
@@
 
-\begin{lstlisting}
-class DispatchRec_Aux (p :: Bool)
-                    (a_1 :: k_1) ... (a_n :: k_n) (b :: *) where
-  dispatchRec_Aux :: X b
+\begin{lstlisting}[style=fancy]
+class DispatchRec.Aux (p :: Bool)
+                     (a.1 :: k.1) ... (a.n :: k.n) (b :: *) where
+  dispatchRec.Aux :: X b
 
-instance C_1 ... => DispatchRec_Aux 'False a_1 ... a_n b where
-  dispatchRec_Aux = f_1
-instance C_2 ... => DispatchRec_Aux 'True  a_1 ... a_n b where
-  dispatchRec_Aux = f_2
+instance C.1 ... => DispatchRec.Aux 'False a.1 ... a.n b where
+  dispatchRec.Aux = f.1
+instance C.2 ... => DispatchRec.Aux 'True  a.1 ... a.n b where
+  dispatchRec.Aux = f.2
 
-class    DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
-  => DispatchRec a_1 ... a_n b
-instance DispatchRec_Aux (P a_1 ... a_k b) a_1 ... a_k ... a_n b
-  => DispatchRec a_1 ... a_n b
+class    DispatchRec.Aux (P a.1 ... a.k b) a.1 ... a.k ... a.n b
+  => DispatchRec a.1 ... a.n b
+instance DispatchRec.Aux (P a.1 ... a.k b) a.1 ... a.k ... a.n b
+  => DispatchRec a.1 ... a.n b
 
-dispatchRec :: ∀a_1 ... a_n b. DispatchRec a_1 ... a_n b => X b
-dispatchRec = dispatchRec_Aux @(P a_1 ... a_k b) @a_1 ... @a_k ... @a_n
+dispatchRec :: ∀a.1 ... a.n b. DispatchRec a.1 ... a.n b => X b
+dispatchRec = dispatchRec.Aux @(P a.1 ... a.k b) @a.1 ... @a.k ... @a.n
 \end{lstlisting}
 
-This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec__Aux}.
+This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec..Aux}.
 The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language.
 The introduction of the associated type class \K{DispatchRec} provides a handy interface for dispatch.
-The dispatch function \K{dispatchRec} decides on one of the functions \K{f_1} and \K{f_2}---depending on whether the predicate \K{P} is true for types \K{a_1}, \dots, \K{a_k}, \K{b}.
+The dispatch function \K{dispatchRec} decides on one of the functions \K{f.1} and \K{f.2}---depending on whether the predicate \K{P} is true for types \K{a.1}, \dots, \K{a.k}, \K{b}.
 
@@
 {\lset
-\begin{lstlisting}
-(/\textbf{\textrm{Movement functions}}/)
+\begin{lstlisting}[style=fancy]
+(/\textbf{\textrm{\color{black}Movement functions}}/)
 
@@
 (/\vspace{-0.3cm}/)
-(/\textbf{\textrm{Starting navigation}}/)
+(/\textbf{\textrm{\color{black}Starting navigation}}/)
 
@@
 (/\vspace{-0.3cm}/)
-(/\textbf{\textrm{Ending navigation}}/)
+(/\textbf{\textrm{\color{black}Ending navigation}}/)
 
@@
 (/\vspace{-0.3cm}/)
-(/\textbf{\textrm{Updating}}/)
+(/\textbf{\textrm{\color{black}Updating}}/)
 
@@
 Define a pair of mutually recursive datatypes for a rose tree and a forest, where the forest is a list of trees, and the tree is defined as a value in a node and a forest of its children:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data RoseTree a = RTree a (Forest a)
@@
 Updating the trees can be done through a class:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 class UpdateTree a b where
@@
 The latter is instantiated with the \K{Maybe} monad that wraps the result type of the movement functions.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 (>>>) :: (a -> b) -> (b -> c) -> (a -> c)
@@
 The type family we need to run the example is defined as follows:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type TreeFam a = '[RoseTree a, Forest a]
@@
 Finally, we can use zipper operations with our updating function to traverse and replace a part of a forest:
-\begin{lstlisting}
+\begin{lstlisting}[style=ttstyle]
 *Main> let forest
@@
 This yields the following result:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 Forest (RTree 'a' $ Forest (RTree 'b' Empty) Empty)
@@
 The location structure consists of a focal subtree, which is one of the mutually recursive nodes of the whole structure of the family of datatypes, and its surrounding context:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data Loc (r :: *) (fam :: [*]) (c :: * -> Constraint) where
@@
 The wrapper incapsulates the proofs about a number of important properties of a focus.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data Focus (r :: *) (a :: *) (fam :: [*])
@@
 %~ The \K{In} constraint is defined as
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type In a fam = InFam a fam == 'True
@@
 Again, this generalizes (in a weak form) the proof of type equality from Section~\ref{subsec:solve-subterms}.
-The definitions of \K{ProofFocus} and an auxiliary class \K{ProofFocus__Aux} are given in Figure~\ref{fig:proof-in}.
+The definitions of \K{ProofFocus} and an auxiliary class \K{ProofFocus..Aux} are given in Figure~\ref{fig:proof-in}.
 In the setting of families, where there are no analogues of \K{:==:}, we can define only a weak kind of proof, which is more flexible.
-We can then make it handy by defining a function \K{castFocus} using \K{castFocus__Aux}, as with \K{castEq} before.
+We can then make it handy by defining a function \K{castFocus} using \K{castFocus..Aux}, as with \K{castEq} before.
 %The constraints on the second \K{ProofFam} instance, which repeat the same ones from the \K{Family} constructor, constitute the proof.
@@
 {\lset
-\begin{lstlisting}
-class ProofFocus_Aux (inFam :: Bool) (r :: *) (a :: *) (b :: *)
+\begin{lstlisting}[style=fancy]
+class ProofFocus.Aux (inFam :: Bool) (r :: *) (a :: *) (b :: *)
                     (fam :: [*]) (c :: * -> Constraint) where
-  castFocus_Aux :: b -> Maybe (Focus r a fam c)
+  castFocus.Aux :: b -> Maybe (Focus r a fam c)
 
-instance ProofFocus_Aux 'False r a b fam c where
-  castFocus_Aux _ = Nothing
+instance ProofFocus.Aux 'False r a b fam c where
+  castFocus.Aux _ = Nothing
 instance (Generic b, In b fam, ZipperI r a b fam c)
-      => ProofFocus_Aux 'True  r a b fam c where
-  castFocus_Aux   = Just . Focus
+      => ProofFocus.Aux 'True  r a b fam c where
+  castFocus.Aux   = Just . Focus
 
-class    ProofFocus_Aux (InFam b fam) r a b fam c
+class    ProofFocus.Aux (InFam b fam) r a b fam c
     => ProofFocus r a b fam c
-instance ProofFocus_Aux (InFam b fam) r a b fam c
+instance ProofFocus.Aux (InFam b fam) r a b fam c
     => ProofFocus r a b fam c
@@
 So the stack of contexts, essentially, reflects the track of the movement inside the structure.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data Contexts (r :: *) (a :: *) (fam :: [*])
@@
 The definitions for the other type-level operations are similar.
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type family (.*) (x :: *) (ys :: [[*]]) :: [[*]] where
@@
 The definition of differentiation resembles its analogue from calculus, but it is now generalized for the setting of families of datatypes:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type family DiffProd (fam :: [*]) (xs :: [*]) :: [[*]] where
@@
 The following completes the computation of the context type:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type family ToContext (fam :: [*])
@@
 Finally, the type of a context frame representation is a type synonym:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type Context fam a = SOP I (ToContext fam (Code a))
@@
 Assume we have a polymorphically recursive datatype \K{PolyRec a} defined in terms of a type family \K{Poly}:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data PolyRec a = Tail a | Rec a (PolyRec (Poly a))
@@
 Since any datatype with type parameters \K{f a b c ...} has kind \K{* -> (* -> (* -> ...))}, a \K{PolyEq} type family can be defined thus:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 type family PolyEq (a :: k) (x :: k) :: Bool where
@@
 The only piece of its definition must be changed as well in order to recursively invoke the function each time with a new type (with the proper change in the \K{CaseRecShow} definition):
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 instance Show  a => CaseShow 'False a where
@@
 Unfortunately, this approach, albeit working well for datatypes defined like \K{PolyRec a}, becomes unsuitable for nested datatypes, such as one below:
-\begin{lstlisting}
+\begin{lstlisting}[style=fancy]
 data Nested a = Epsilon | Nest a (Nested [a])

### 2018-11-16 lncs-paper.tex
More examples of generic functions, address some small-scale remarks from the TFP reviews

@@
 \usepackage{amssymb}
+\usepackage{amsbsy}
 \usepackage{listings}
@@
 {∀}{{\fa}}1 {==}{{\tteq}}1 {...}{{\scriptsize\bf\dots}}2
-{>>>}{{\scriptsize\comp}}3 {>=>}{{\scriptsize\kleisli}}3
+{>>>}{{\footnotesize\comp}}3 {>=>}{{\footnotesize\kleisli}}3
 {.1}{{$_\mathit{1}$}}1 {.2}{{$_\mathit{2}$}}1
+{.3}{{$_\mathit{3}$}}1
+{.i}{{$_\mathit{i}$}}1 {.j}{{$_\mathit{j}$}}1
 {.n}{{$_\mathit{n}$}}1 {.k}{{$_\mathit{k}$}}1
+{.R}{{\scriptsize$_\mathit{\hspace{0,07cm} \boldsymbol{R}}$}}1
+{..R}{{\scriptsize$_\mathit{\hspace{-0,02cm} \boldsymbol{R}}$}}1
+{,R}{{\scriptsize$_\mathit{\hspace{-0,05cm} \boldsymbol{R}}$}}1
+{,,R}{{\scriptsize$_\mathit{\boldsymbol{R}}$}}1
 {.Aux}{{\large$_\mathit{\hspace{0,05cm} Aux}$}}3
+{.AExpr}{{\large$_\mathit{\hspace{0,05cm} AExpr}$}}5
 {..Aux}{{\large$_\mathit{\hspace{-0,03cm} Aux}$}}3
@@
 {_}{{\underline{\hspace{0.45em}}}}1
+{[.}{{[\hspace{-0,06cm}}}1
+{[..}{{[\hspace{-0,1cm}}}1
+{,,]}{{\hspace{-0,1cm}}]}1
+{,,,]}{{\hspace{-0,16cm}}]}1
+{,]}{{\hspace{-0,06cm}]}}1
 {+}{{\scriptsize+}}1
+{,*.}{{\nhs$*$\nhs}}1
+{+.}{{\footnotesize+}}1
+{.+.}{{\nhs\footnotesize+\nhs}}1
+{*.}{{\footnotesize$\times$}}1
+{.*.}{{\nhs\footnotesize$\times$\nhs}}1
 {.-}{{\nhs-}}1
@@
 \newcommand{\nhs}{\hspace{-0.06cm}}
+\newcommand{\Nhs}{\hspace{-0.1cm}}
 
@@
 \newcommand{\tteq}{{\scriptsize$\thicksim$}}
-\newcommand{\comp}{\tg\nhs\tg\nhs\tg}
-\newcommand{\kleisli}{\tg\nhs=\nhs\tg}
+\newcommand{\comp}{\tg\Nhs\tg\Nhs\tg}
+\newcommand{\kleisli}{\tg\Nhs=\Nhs\tg}
 \newcommand{\ann}{:\texttt{\nhs}:}
@@
 \email{bolotina@sfedu.ru} \and
-Czech Technical University in Prague, Prague, Czech Republic\\
-\email{pelenart@fit.cvut.cz}}
+Northeastern University, Boston, USA\\
+\email{artem@ccs.neu.edu}}
 %
@@
 There are many generic functions that deal with the recursive knots when traversing the structure of datatypes.
-Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}; more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}.
+Some of the most general examples are \emph{maps}~\cite{Magalhaes2010} and \emph{folds}~\cite{Meijer1991}; a more advanced one is a \emph{zipper}~\cite{Huet1997,HiJeLo2004,Adams2010}.
 For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}.
@@
 That variant and its later implementation with type families~\cite{Kiselyov2012} fully eliminate \textsf{OverlappingInstances}.
-Although, each type needs a representation instance to be derived---by means of Template Haskell or GHC.
+Despite this, each type needs a representation instance to be derived---by means of Template Haskell or GHC.
 The most generic solution for type equality, the second version, again makes use of overlapping instances, however.
@@
 \end{lstlisting}
-An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list is an $n$-ary product that represents a sequence of constructor arguments, while the outer list, an $n$-ary sum, corresponds to a choice of a particular constructor.
+An $n$-ary sum and an $n$-ary product are therefore modelled as type-level heterogeneous lists: the inner list represents an $n$-ary product, isomorphic to a sequence of constructor arguments, while the outer list, representing an $n$-ary sum, corresponds to a choice of a particular constructor.
 
@@
 \end{lstlisting}
-If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[a]}. Here is an example value of type \K{NP I}:
+If instantiated with \K{I}, \K{NP} is a plain heterogeneous list, while \K{K a} turns it into a homogeneous one, isomorphic to \K{[.a]}. Here is an example value of type \K{NP I}:
 \begin{lstlisting}[style=fancy]
@@
 We turn to the sum definition now.
-The constructor \K{S} of \K{NS}, given an index in $n$-element list, results in an index in a list with $n+1$ elements, skipping the first one, while \K{Z} stores the payload of type \K{f x}.
-For example, the following chooses the third element of a sum:
+The constructors \K{S} and \K{Z} of \K{NS} allow to represent choices from a sum as indices in the form of Peano numbers.
+Firstly, the application of \K{Z} represents the choice of the first component of the sum.
+Secondly, \K{S} is the successor constructor.
+That is, \K{S .
+Z} chooses the second component, and so on up to the length of the index list \K{xs}.
+Given a sum of type \K{NS f xs}, the $i$-th choice has type \K{f x}, where \K{x} is the $i$-th element in \K{xs}.
+For example, the following represents the third choice from a sum:
 \begin{lstlisting}[style=fancy]
@@
 \end{lstlisting}
-The sum constructors are similar to Peano numbers, so the choice from a sum of products of a datatype matches the index of its particular constructor in the index list and stores the product representing arguments of that constructor.
-
-With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to} witnessing the isomorphism between a datatype and its generic representation:
+Putting this together, each term of a particular datatype can be represented as the choice from the sum of products, representing that datatype.
+The choice matches the index of its particular constructor in the index list, and stores the product, representing arguments of that constructor.
+
+With the \K{NS} and \K{NP} machinery at hand, SOP defines the \K{Generic} class with conversion functions \K{from} and \K{to}, witnessing the isomorphism between a datatype and its generic representation:
 \begin{lstlisting}[style=fancy]
@@
 
-We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition as irrelevant to our work.
+We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition, as it is irrelevant to our work.
 Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following.
@@
 The algorithm of \K{subtermsNP} is straightforward---it traverses the product, appending the current element to the result list if its type is the same as of the original term, otherwise skipping the element.
-We use GHC's \textsf{TypeAppications} extension to pass that type.
+We use GHC's \textsf{TypeApplications} extension to pass that type.
 
@@
 There is no clear path to this at the moment.
-Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
+Therefore, we step back (until Section~\ref{subsec:solve-subterms}) and, to implement \K{subtermsNP}, follow the \textsf{QuickCheck}'s example\footnote{The \textsf{QuickCheck} library applies another approach to generic programming, namely \textsf{GHC.Generics}.}, using overlapping instances of a dedicated class instead.
 \begin{lstlisting}[style=fancy]
@@
 
-Although the approach works, as exemplified by a number of the packages on Hackage, we aim for release of generic programs from overlap.
+Although the approach works, as exemplified by a number of the packages on Hackage, we aim to release generic programs from overlap.
 This would remove the complexity overhead introduced by the approach, as we have mentioned in the introduction.
@@
 For this variant, we reimplement \K{subtermsNP}.
-The different implementation employes powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}.
+The different implementation employs powerful machinery of high-level combinators---one of the major advantages provided by \textsf{generics-sop}.
 Although, those high-level functions become one extra layer of complexity.
@@
 In summary, we claim that any generic function accessing recursive knots in the underlying datatype structure can be defined in the way described above for the task of \K{subterms}.
-We give another example showing how to adapt our idiom to different scenarios in the following subsection.
-
-\subsection{Generic show}
+We give other examples showing how to adapt our idiom to different scenarios in the following subsections.
+
+\subsection{Generic \K{show}}
 \label{subsec:gshow}
@@
 In the first case, when the position we are matching on is not recursive, we only require it to be an  instance of \K{Show}, and invoke its \K{show} function.
-Whereas in the case of the recursive position, we surround it with parentheses and apply our generic function \K{gshow}.
+Whereas in the case of the recursive position, we apply our generic function \K{gshow}.
 Thus, by means of the type family for equality, we model a form of pattern matching on the types again:
@@
 
-The resulting function \K{gshow} is a subject of a number of constraints abbreviated by a \K{GShow} synonym:
+The resulting function \K{gshow} is subject to a number of constraints abbreviated by a \K{GShow} synonym:
 \begin{lstlisting}[style=fancy]
@@
 We define \K{gshow} mutually recursive with \K{caseShow}.
-The full implemenentation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
+The full implementation of the function \K{gshow} is left for the extended version of the paper in Technical Report\footnote{\url{https://users.fit.cvut.cz/~pelenart/2018-generic-zipper-tr.pdf}}.
 
@@
 \end{lstlisting}
-Here is a benefit of our implementation: it can be used directly, without any additional instance declarations, whereas \textsf{basic-sop}~\cite{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}:
+
+Our implementation of \K{gshow} obviates one drawback of its analogue from \textsf{basic-sop}.
+This drawback is that \K{gshow} from \textsf{basic-sop} does not treat recursive positions separately, and therefore requires the \K{Show} constraint for all knots in the datatype structure:
+\begin{lstlisting}[style=fancy]
+gshow :: ∀a. (Generic a, HasDatatypeInfo a, All2 Show (Code a))
+      => a -> String
+\end{lstlisting}
+As a consequence, \textsf{basic-sop} offers the following usage pattern for \K{gshow} and some datatype \K{T}, requiring an additional \K{Show} instance declaration for \K{T}:
 \begin{lstlisting}[style=fancy]
@@
 \end{lstlisting}
-This is a consequence of \K{gshow} from \textsf{basic-sop} not treating recursive positions separately, and therefore requiring the \K{Show} constraint for all knots in the datatype structure.
+As we have shown,  the variant of the function \K{gshow}, distinguishing the recursion cases, can be used directly, without this extra declaration.
+This takes advantage of employing the polytypic instance of \K{CaseRecShow} for all datatypes.
+
+\subsection{Generic recursion schemes}
+\label{subsec:rec-schemes}
+
+Recursion schemes, such as \K{fold}~\cite{Meijer1991,MuRec2009} and \K{compos}~\cite{Bringert2006}, are classical examples of generic functions that are usually treated by a fixed point view.
+In this subsection, we show how they can be defined in the SOP view, adopting our approach.
+
+\subsubsection{Generic \K{compos}}
+
+The function \K{compos} is a traversal combinator for defining compositional functions. It applies a given function to all immediate children of a given term ot type \K{T}:
+\begin{lstlisting}[style=fancy]
+compos :: (T -> T) -> T -> T
+\end{lstlisting}
+This can be used, for example, with a datatype \K{Expr} for defining the \K{renameVar} function, updating all variables in a given expression without affecting its other subexpressions:
+\begin{lstlisting}[style=fancy]
+data Expr = EAbs String Expr
+          | EApp Expr Expr
+          | EVar String
+
+renameVar :: Expr -> Expr
+renameVar (EVar x) = EVar $ x ++ "_"
+renameVar e        = compos renameVar e
+\end{lstlisting}
+
+The implementation of the generic function \K{gcompos} employs a class for resolving recursion cases, which now looks as the following:
+\begin{lstlisting}[style=fancy]
+class CaseCompos.Aux (eq :: Bool) (a :: *) (x :: *) (y :: *) where
+  caseCompos.Aux :: (a -> a) -> I x -> I y
+
+instance CaseCompos.Aux 'False a x x where
+  caseCompos.Aux _ = id
+instance CaseCompos.Aux 'True  a a a where
+  caseCompos.Aux f = I . f . unI
+\end{lstlisting}
+The function \K{caseCompos}, an abbreviation for \K{caseCompos..Aux @(Equal a x)}, is supposed to be applied to each component of the product, representing arguments of a chosen constructor.
+Given a component of type \K{I x}, it results in the component of different type \K{I y}.
+(The implementation below makes use of a higher-order function, whose type does not admit that \K{x} and \K{y} can be unified here.) The first case in this definition corresponds to mapping over a constant, which should not be changed.
+The second case is the application of the given function \K{f} to a recursive position.
+
+The rest of the work is handled by \textsf{generics-sop}'s function \K{trans_SOP}:
+\begin{lstlisting}[style=fancy]
+trans_SOP :: AllZip2 c xss yss
+          => proxy c -> (forall x y. c x y => f x -> g y)
+          -> SOP f xss -> SOP g yss
+\end{lstlisting}
+This function transforms the generic representation of a term, applying a given function to every component of a chosen product.
+\K{AllZip2} here is analogous to \K{All2}, zipping two lists of lists with the binary constraint \K{c}.
+
+Finally, the definition of \K{gcompos}, using \K{trans_SOP} with \K{caseCompos}, is following, with \K{CaseCompos} abbreviating the use of \K{CaseCompos..Aux}:
+\begin{lstlisting}[style=fancy]
+type GCompos a = (Generic a, AllZip2 (CaseCompos a) (Code a) (Code a))
+
+gcompos :: ∀a. GCompos a => (a -> a) -> a -> a
+gcompos f = to . composSOP . from
+  where
+    composSOP = trans_SOP (Proxy @(CaseCompos a)) $ caseCompos f
+\end{lstlisting}
+
+\subsubsection{Generic \K{fold}}
+
+A fold is a higher-order function that recursively deconstructs a term, using a combining operation on its structure to aggregate the result.
+For example, consider a fold for a datatype of arithmetic expressions \K{AExpr}:
+\begin{lstlisting}[style=fancy]
+data AExpr = EConst Int
+           | EAdd AExpr AExpr
+           | EMul AExpr AExpr
+
+fold.AExpr :: (Int -> r, r -> r -> r, r -> r -> r) -> AExpr -> r
+fold.AExpr f@(f.1, f.2, f.3) e = case e of
+  EConst x -> f.1 x
+  EAdd l r -> f.2 (fold.AExpr f l) (fold.AExpr f r)
+  EMul l r -> f.3 (fold.AExpr f l) (fold.AExpr f r)
+\end{lstlisting}
+Given \K{f}---called an \emph{algebra}---it proceeds through the recursive structure of a given term \K{e}, applying its components \K{f.1}, \K{f.2}, and \K{f.3} to elements of \K{e} and combining the results.
+This can be used to evaluate expressions.
+Consider:
+\begin{lstlisting}[style=fancy]
+example = EAdd (EMul (EConst 3) (EConst 2))
+               (EAdd (EMul (EConst 2) (EConst 2))
+                     (EConst 5))
+\end{lstlisting}
+The call
+\begin{lstlisting}[style=fancy]
+fold.AExpr (id, (+), (*)) example
+\end{lstlisting}
+yields 15.
+
+In the example above, \K{(Int -> r, r -> r -> r, r -> r -> r)} is a type of algebras for the datatype \K{AExpr} to a result type \K{r}.
+This type represents the underlying recursive structure of \K{AExpr}.
+In particular, the constructors of \K{AExpr} form an algebra for expressions, which is isomorphic to \K{AExpr}:
+\begin{lstlisting}[style=fancy]
+(EConst, EAdd, EMul) :: (Int -> AExpr,
+                         AExpr -> AExpr -> AExpr,
+                         AExpr -> AExpr -> AExpr)
+\end{lstlisting}
+A type of algebras for a particular datatype \K{a} can be computed generically from its representation code.
+The result type \K{r} should occur in this type whenever \K{a} (a recursive position) occurs in the code.
+In the following, the type families \K{AlgebraS} and \K{AlgebraP} are employed to compute a type of algebras for a given datatype \K{a} and a result type \K{r}.
+\begin{lstlisting}[style=fancy]
+type family AlgebraS (code :: [[*]]) a r :: [*] where
+  AlgebraS (xs ': xss) a r = (AlgebraP xs a r ': AlgebraS xss a r)
+  AlgebraS '[]         _ _ = '[]
+
+type family AlgebraP (xs :: [*]) a r where
+  AlgebraP (a ': xs) a r = r -> AlgebraP xs a r
+  AlgebraP (x ': xs) a r = x -> AlgebraP xs a r
+  AlgebraP '[]       _ r = r
+
+type Algebra a r = NP I (AlgebraS (Code a) a r)
+\end{lstlisting}
+The calculated type of algebras \K{Algebra a r} is meant to be an \K{NP} of its components, which are constructed by \K{AlgebraP} from each product in the code.
+All the components, being applied to the elements of a datatype structure, return a result of type \K{r}, as handled by the last case in the \K{AlgebraP} definition.
+
+We turn to defining a generic fold.
+First, define a class \K{CaseFold..Aux} with a function \K{caseFold..Aux}, which perform the main work and serve for managing the two recursion cases.
+As before, we will use the synonyms \K{CaseFold} and \K{caseFold} to abbreviate their use with the application of \K{Equal a x}:
+\begin{lstlisting}[style=fancy]
+class CaseFold.Aux (eq :: Bool)
+                  (a :: *) (b :: *) (x :: *) (y :: *) where
+  caseFold.Aux :: Algebra a b -> I x -> I y
+
+instance CaseFold.Aux 'False a b x x where
+  caseFold.Aux _ = id
+instance GFold a b => CaseFold.Aux 'True a b a b where
+  caseFold.Aux f = I . applyAlgebra @a f . foldSOP . from . unI
+    where
+      foldSOP = trans_SOP (Proxy @(CaseFold a b)) $
+                          caseFold @a @b f
+\end{lstlisting}
+The function \K{caseFold..Aux} is meant to operate on the elements of a product constituting a part of a datatype's representation.
+For constants, there are no recursive calls, so an operation for that case is trivially the identity function.
+Whereas in the case of the recursion point, this turns the element to its representation, then recursively processes that, and applies the given algebra to the result.
+This employs the function \K{applyAlgebra} that applies the algebra to the representation gained as a result of processing:
+\begin{lstlisting}[style=fancy]
+applyAlgebra :: ∀a b. ApplyAlgebra a b
+             => Algebra a b -> SOP I (AlgCode a b) -> b
+\end{lstlisting}
+That representation has the code, calculated from the origin code of \K{a} by replacing its recursive occurences therein with the algebra's result type \K{b}.
+\K{AlgCode a b} here is a type synonym for the result of this computation.
+We omit the implementation of \K{applyAlgebra}, as well as of the type-level machinery of \K{AlgCode} and \K{ApplyAlgebra}, for the sake of space.
+
+The type applications \K{@a} and \K{@b} appear in the lines above, because they are needed to calculate the \K{Algebra a b} type.
+Also, the constraint \K{GFold a b} on the second instance of \K{CaseFold..Aux} is necessary to use \K{applyAlgebra}, \K{caseFold}, and \K{from}:
+\begin{lstlisting}[style=fancy]
+type GFold a b = (Generic a
+                 , AllZip2 (CaseFold a b) (Code a) (AlgCode a b)
+                 , ApplyAlgebra a b)
+\end{lstlisting}
+
+The generic fold function \K{gfold} is defined using \K{caseFold}:
+\begin{lstlisting}[style=fancy]
+gfold :: ∀a b. GFold a b => Algebra a b -> a -> b
+gfold f = unI . caseFold @a @b f . I
+\end{lstlisting}
+
+This can be conveniently used with various algebras that are now of type \K{NP}, if we define an infix operator \K{&} for constructing products:
+\begin{lstlisting}[style=fancy]
+infixr 1 &
+x & ys = I x :* ys
+\end{lstlisting}
+The call of \K{gfold} with \K{example}, defined above in this subsection,
+\begin{lstlisting}[style=fancy]
+gfold (id & (+) & (*) & Nil) example
+\end{lstlisting}
+again yields 15.
 
@@
 
-This machinery employes the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec..Aux}.
+This machinery employs the abstract multi-place predicate \K{P} on types, supposed to be a closed typed family, to resolve the overlap of two dispatch branches, as captured by \K{DispatchRec..Aux}.
 The predicate \K{P} generalizes \K{Equal} to an arbitrary relation on types, so that one can model, e.g., subtyping for some domain-specific language.
@@
 The dispatch function \K{dispatchRec} decides on one of the functions \K{f.1} and \K{f.2}---depending on whether the predicate \K{P} is true for types \K{a.1}, \dots, \K{a.k}, \K{b}.
-
-For consumer functions, such as \K{subterms} and \K{gshow}, the type \K{X b} typically is instantiated with \K{b -> T} for some datatype \K{T}.
+Here, the type \K{b} is supposed to be an instance of \K{Generic}.
+
+The type \K{X b}, associated with \K{b}, depends of the kind of the generic function.
+All generic functions may be classified into three categories~\cite{Rodriguez2008}: \emph{consumers}, \emph{transformers}, and \emph{producers}.
+Given type \K{b}, an instance of \K{Generic}, they turn \K{b} into a constant type, \K{b} into \K{b} with a changed value, and a constant type into \K{b}, respectively.
+For consumer functions, such as \K{gshow} (Section~\ref{subsec:gshow}), the type \K{X b} typically is instantiated with \K{b -> T} for some datatype \K{T}.
+For transformers, such as \K{gcompos} (Section~\ref{subsec:rec-schemes}), \K{X b} typically takes the form \K{S -> I a.i -> I a.j}.
+Here, \K{a.i}, \K{a.j} for $\textcolor{cbasic}{i} \neq \textcolor{cbasic}{j}$ are from \K{a.1 ...
+a.n}, and are supposed to be equal to \K{b} at recursion points, and \K{S} is some type.
+This suggests that the transformer function changes \K{b}, employing values of type \K{S}.
 The pattern fits equally well for defining producers, such as \K{garbitrary} from the \textsf{basic-sop} package, where \K{X b} would be instantiated with \K{Gen b}.
@@
 
-In Technical Report, we discuss the implementation of the generic zipper interface---the functions for manipulating locations.
+In a technical report, we discuss the implementation of the generic zipper interface---the functions for manipulating locations.
 The source code with the full implementation of the zipper is available at our \textsf{GitHub} repository\footnote{\url{https://github.com/Maryann13/Zipper}}.
@@
 The subtree in focus is wrapped by the \K{Focus} datatype. 
-The wrapper incapsulates the proofs about a number of important properties of a focus.
+The wrapper encapsulates the proofs about a number of important properties of a focus.
 \begin{lstlisting}[style=fancy]
@@
 Existential type variable \K{b} represents the type of a focus.
-We apply a number of predicates to \K{b}, hence we can implement the steps of the navigation not knowing the actual type of a focus.
+We apply a number of predicates to \K{b}, hence we can implement the steps of the navigation without knowing the actual type of a focus.
 Firstly, the type of a focus should have the \K{Generic} representation.
@@
 \end{lstlisting}
-The definition of \K{InFam} is omitted as a boring one.
+The definition of \K{InFam} is omitted, as it is a boring one.
 
@@
 
-Again, kind \K{[*]} denotes products, and \K{[[*]]} denotes sums (of products), so the relation with arithmetic of numbers becomes more clear if one realizes that an empty sum \K{'[] :: [[*]]} corresponds to $0$, and an empty product \K{'[] :: [*]} corresponds to $1$.
+Again, kind \K{[,*.]} denotes products, and \K{[.[,*.],]} denotes sums (of products), so the relation with arithmetic of numbers becomes more clear if one realizes that an empty sum \K{'[,,] :: [.[,*.],]} corresponds to $0$, and an empty product \K{'[,,] :: [,*.]} corresponds to $1$.
 
@@
 \end{lstlisting}
-The differentiation of the single type reflected by a one-element list here results in $0$ reflected by \K{'[]}, if that is not in the family and hence is regarded as a constant.
-Otherwise it results in $1$ represented by the sum \K{'[ '[]]}.
+The differentiation of the single type reflected by a one-element list here results in $0$ reflected by \K{'[,,]}, if that is not in the family and hence is regarded as a constant.
+Otherwise it results in $1$ represented by the sum \K{'[ '[,,],,,]}.
 When differentiation gives $1$, it is actually the hole.
@@
 
+\section{Generic generic programming}
+\label{sec:ggp}
+
+In the previous sections, we have shown that many generic functions that treat recursion can be defined, following one pattern, in the SOP view, which does not employ a fixed point.
+Despite this, some generic functions can be defined in a more elegant way, when using a fixed point view.
+Furthermore, there are such ones, for which a fixed point view is essential.
+A prominent example of this is generic pattern-matching~\cite{VanNoort2008}.
+It involves generic definitions of patterns for datatypes, which extend their nodes with a metavariable environment.
+A pattern type is defined as a sum of a datatype's representation functor and a type of variables, closed under a fixed point.
+This, therefore, cannot be builded out of the SOP generic representation of a datatype.
+
+Still, one can efficiently employ these views together in one program, taking advantages of the both.
+Combining the use of different generic encodings to write programs is a known technique in generic programming.
+Magalh{\~a}es and L{\"o}h~\cite{MagLoeh2014} investigate optimization of interaction between various generic views in their work on \emph{generic generic programming}.
+It proposes to automatically derive generic representations of datatypes for numerous generic views, defining them via conversions from one particular view.
+Specifically, these are translations between the analogues of the \K{Generic} class in different libraries.
+This obviates the need of writing multiple blocks of code for their instances for one datatype, thus optimizing generic programs.
+
+\textsf{regular}~\cite{VanNoort2008} is a generic programming library, designed for generic rewriting, that represents regular datatypes, using a fixed point view.
+Converting to this view requires detecting recursion points in the datatype structure, so it can adopt our approach.
+In this section, we define a conversion from the \textsf{generics-sop} view to the \textsf{regular} one.
+
+\subsubsection{Encoding \textsf{\textmd{regular}}}
+
+The \textsf{regular} library represents datatypes as sums of products by means of the following single combinators (we omit the metadata combinator, for simplicity):
+\begin{lstlisting}[style=fancy]
+newtype K.R a    r = K.R a
+newtype I.R      r = I.R r
+data U.R         r = U.R
+data (f :+.:,R g) r = L.R (f r) | R.R (g r)
+data (f :*.:,R g) r = f r :*.:,R g r
+
+infixr 6 :+.:,R
+infixr 7 :*.:,R
+\end{lstlisting}
+It employs the type \K{:.+.:,R} of binary sums and the type \K{:.*.:,R} of binary products.
+By nesting the sums, it chains constructors, and by nesting the products, it chains fields within one constructor.
+The types \K{K..R}, \K{I..R}, and \K{U..R} represent fields of a constant type, recursive positions, and constructors without fields (called ``units''), respectively.
+
+Using these representation types, \textsf{regular} encodes the underlying recursive structure of datatypes as polynomial functors. For example, recall the type \K{AExpr}:
+\begin{lstlisting}[style=fancy]
+data AExpr = EConst Int | EAdd AExpr AExpr | EMul AExpr AExpr
+\end{lstlisting}
+The corresponding polynomial functor is
+\begin{lstlisting}[style=fancy]
+type PF.AExpr = K.R Int :+.:,R I.R :*.:,R I.R :+.:,R I.R :*.:,R I.R
+\end{lstlisting}
+Datatypes can then be represented as their polynomial functors, closed under a fixed point operator.
+Although, for practical tasks, it is sufficient to convert only one layer of the datatype structure.
+\textsf{regular} provides a \K{Regular} class of representable datatypes, with an associated type \K{PF} for a polynomial functor and shallow conversion functions \K{from..R} and \K{to..R}:
+\begin{lstlisting}[style=fancy]
+class Regular a where
+  type PF a :: * -> *
+  from.R :: a -> PF a a
+  to.R   :: PF a a -> a
+\end{lstlisting}
+The functions perform conversion between the datatype \K{a} and its generic representation \K{PF a a} that stores elements of type \K{a} in the recursive positions.
+
+Now, we turn to defining translation from \textsf{generics-sop} to \textsf{regular}.
+All the conversion work is divided into two steps.
+First, we perform type-level translation of the SOP codes into the \textsf{regular} representation types.
+Then, we convert representations at the level of terms.
+
+\subsubsection{Type-level conversion}
+
+The type-level conversion is done by type families:
+\begin{lstlisting}[style=fancy]
+type family RegS t (xss :: [[*]]) :: * -> * where
+  RegS t '[a]     = RegP t a
+  RegS t (a ': b) = RegP t a :+.:,R RegS t b
+
+type family RegP t (xs :: [*]) :: * -> * where
+  RegP t '[a]     = RegEl t a
+  RegP t (a ': b) = RegEl t a :*.:,R RegP t b
+  RegP _ '[]      = U.R
+
+type family RegEl t a :: * -> * where
+  RegEl t t = I.R
+  RegEl _ a = K.R a
+\end{lstlisting}
+This machinery systematically turns $n$-ary sums of products to corresponding nested binary sums of binary products.
+Empty products are being turned into units \K{U..R}, and recursive positions and constants are being wrapped into the combinators \K{I..R} and \K{K..R}, respectively.
+
+The \textsf{regular} representation type for the datatype \K{a} is then defined as
+\begin{lstlisting}[style=fancy]
+type Reg a = RegS a (Code a) a
+\end{lstlisting}
+
+\subsubsection{Term-level conversion}
+
+The value conversion of representations is handled by the type classes \K{ConvS}, \K{ConvP}, and \K{ConvEl}:
+\begin{lstlisting}[style=fancy]
+class ConvS (s :: Bool) a (xss :: [[*]]) where
+  toRegS :: NS (NP I) xss -> RegS a xss a
+
+class ConvP (s :: Bool) a (xs :: [*]) where
+  toRegP :: NP I xs -> RegP a xs a
+
+class ConvEl (eq :: Bool) a x where
+  toRegEl :: x -> RegEl a x a
+\end{lstlisting}
+The logical type parameter \K{s} in the first two declarations is used to distinguish the cases, when the lists \K{xss} and \K{xs} consist of a single element.
+
+We only show the \K{ConvEl} instances that do work on managing recursion:
+\begin{lstlisting}[style=fancy]
+instance ConvEl 'True t t where
+  toRegEl x  = I.R x
+instance RegEl t a == K.R a => ConvEl 'False t a where
+  toRegEl x  = K.R x
+\end{lstlisting}
+The second case requires the context, witnessing that \K{RegEl t a} results in \K{K..R a}, because \K{RegEl t a} is less specific than \K{RegEl t t}, hence may be overlapped.
+
+The conversion function from SOP to \textsf{regular} is defined as
+\begin{lstlisting}[style=fancy]
+toReg :: ConvS a (Code a) => SOP I (Code a) -> Reg a
+toReg = toRegS . unSOP
+\end{lstlisting}
+
+Finally, we use this function to give an instance of \K{Regular} for all instances of SOP's \K{Generic}:
+\begin{lstlisting}[style=fancy]
+instance (Generic a, ConvS a (Code a)) => Regular a where
+  type PF a = RegS a (Code a)
+  from.R = toReg . from
+\end{lstlisting}
+
 \section{Discussion}
@@
 \end{lstlisting}
-The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the \K{GShow} constraint for \K{[a]}, \K{[[a]]}, and so on ad infinitum.
+The culprit is the constraint \K{All2 (CaseRecShow a) (Code a)} that now turns out to be a root of nonterminating computation of the \K{GShow} constraint for \K{[.a]}, \K{[..[.a],]}, and so on ad infinitum.
+Similarly, the approach fails for functions that are nonrecursive itself, but meant to be used recursively, such as \K{gcompos} from Section~\ref{subsec:rec-schemes}, as it induces an infinite constraint at a call site of the generic function.
 
 Still, one can think of problems where a generic operation uses the recursive structure of data, but has no recursive calls---and, therefore, can tolerate nested datatypes.
-An example of this kind of problems is \emph{generic generic programming}~\cite{MagLoeh2014}, a technique of defining translations between different generic programming universes.
-In our setting, it is possible to define a translation from \textsf{generics-sop} to another library with an explicit encoding of recursive positions, such as \textsf{regular}~\cite{VanNoort2008}.
-This usually needs only to detect recursion points, because the most of generic views provide shallow conversion functions.
+An example of this kind of problems is ``generic generic programming'' (see Section~\ref{sec:ggp}).
+In our setting, it is possible to define a translation from \textsf{generics-sop} to another generic view that explicitly encodes recursive positions and supports polymorphic recursion---for example, \textsf{generic-deriving}~\cite{Magalhaes2010}.
+This needs only to detect recursion points, because this library provides shallow conversion functions.
 
@@
 There are several generic views that use certain forms of the fixed point operator to express recursion in a datatype structure~\cite{VanNoort2008,MuRec2009,Jansson1997,Loeh2011}.
-And there are a number of approaches that do not make use of fixed points~\cite{Chakravarty2009,Cheney2002,Magalhaes2010,Weirich2006}, but explicitly encode recursion in the datatype representation.
+And there are a number of approaches that do not make use of fixed points~\cite{Chakravarty2009,Magalhaes2010}, but explicitly encode recursion in the datatype representation.
 The SOP view~\cite{VriLoeh2014}, which we use to demonstrate our technique, is an approach to generic programming that does not reflect recursive positions in the generic representation of a datatype.

### 2018-11-17 lncs-paper.tex
A little bit more discussion on overlapping instances. Attempt to address more remarks

@@
 For handling recursion, several generic programming approaches express datatypes in the form of polynomial functors closed under fixed points~\cite{MuRec2009,Jansson1997,Loeh2011}.
-The SOP view naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise becomes unhandy.
+However, the SOP view does not reflect recursion points in generic representation types.
+So it naturally supports definitions of functions that do not require a knowledge about recursive occurrences, but otherwise becomes unhandy.
 
@@
 However, this may complicate the whole framework significantly.
-Besides, such a decision may lead to extra conversions between the generic views.
+Besides, such a decision may lead to extra conversions between the generic views: the original SOP encoding and the modified one.
 
@@
 Morris and Jones~\cite{Morris2010} extensively discuss the problems arising from overlapping instances.
-% and introduce the type system \textsf{ilab} with a feature called \emph{instance chains} that advances control of overlap. 
-The overlap problem also strikes in the security setting when code is compiled as \K{-XSafe} because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
-%Kiselyov et al.~\cite{Kiselyov2004} show how to localize overlap for access operations in a systematic way.
-Using \textit{closed type families}~\cite{Eisenberg2014}, we introduce an idiom that allows us to completely eliminate overlap for handling recursion.
-% -- this should go to Realted Work (if stays in the paper at all)
+One of those problems is late error reporting, which is a consequence of the fact that GHC resolves overlapping instances at call sites.
+That is, if there are more than one most specific instances for particular types, GHC does not determine that before one's attempt to use the class function with those types.
+Another notable problem is lack of specification for overlapping instances, so their behavior depends on a concrete compiler.
+The overlap problem also strikes in the security setting, when code is compiled as \K{-XSafe}, because GHC does not reflect unsafe overlaps and marks the module as safe~\cite{sh-overlapping}.
 
@@
 type    Rep   a = SOP I (Code a)
+newtype SOP f a = SOP {unSOP :: NS (NP f) a}
 
@@
 
-We leave out discussion of the \K{SListI} constraint in the \K{Generic} class definition, as it is irrelevant to our work.
+The \K{SListI} constraint in the \K{Generic} class definition is used in \textsf{generics-sop} to represent type-level index lists at the level of terms as singletons.
+We leave out discussion of this constraint, as it is irrelevant to our work.
 Although, we do use \K{All} constraint combinator (as in \K{All SListI}) in the following.
@@
 
-For every element in a list of all direct subterms of a term we shall provide a proof object witnessing its type (in)equality to the type of the term.
+For every element in a list of all direct subterms of a term, we shall provide a proof object witnessing its type (in)equality to the type of the term.
 This can be done by means of the \K{All} combinator and partially applied dedicated type class \K{ProofEq}, which abbreviates the heavy-weighted interface of \K{DecideEq}:
@@
 \end{lstlisting}
-The equality proof can then be employed to provide a type-safe cast between two equal types:
+The equality proof can then be employed to provide \K{castEq} (Section~\ref{subsec:recursion-problem}).
+The definition makes use of \K{castWith}, which performs a type-safe cast between two equal types, using the equality witness \K{b :==: a}:
 \begin{lstlisting}[style=fancy]
@@
 \begin{lstlisting}[style=fancy]
-data Expr = EAbs String Expr
-          | EApp Expr Expr
-          | EVar String
+data Expr = EAbs String Expr | EApp Expr Expr | EVar String
 
@@
 \begin{lstlisting}[style=fancy]
-data AExpr = EConst Int
-           | EAdd AExpr AExpr
-           | EMul AExpr AExpr
+data AExpr = EConst Int | EAdd AExpr AExpr | EMul AExpr AExpr
 
@@
 Also, it guarantees that \K{b} fulfils the constraint \K{c}.
-Again, we make \K{c} embedded in \K{Loc} by reason of existentiality of \K{b}, which precludes the enforcement of this restriction by means of external \K{All}.
 

### 2025-01-03 submissions/esop19/rebuttal.md
add TFP18 and ESOP19 submissions info

# Related Works (Rev_1_&_Rev_2)

Reviewers pointed out some of the relevant works missing in our discussion. We are thankful for the suggestions and will add the discussion of the works in future revisions. At the same time, we believe that none of those omissions are crucial. We discuss some of the most relevant works below.

Bahr’s work (“Composing and Decomposing…”) shows the usefulness of closed type families (CTF) in the datatype a la carte-line of work on extensible datatypes. While CTF are essential for us too, generic programming (GP) is a very different setting from the one of Bahr’s. Through the use of CTF, our work aims to advance GP-line of works, to which Bahr’s work does nothing.

In contrast, Cacciari Miraldo and Serrano made great steps advancing GP recently (ICFP’18 workshops), and we are well aware of those. “Generic programming of all kinds” paper aims at a much more ambitious task than ours, and quite orthogonal to it.

"Sums of products for mutually recursive datatypes" (Cacciari Miraldo and Serrano) is very relevant to our work, and the lack of the discussion of it in our text is very unfortunate, indeed! This is merely an oversight. For the purpose of the rebuttal, we summarize the difference with our work as follows: their approach leads to a whole new GP-library (dubbed generics-mrsop) which mixes the generic-sop “pointless” (as in not employing fixed points) approach with the multirec’s “pointful” (based on fixed points) one. In contrast, our approach stays within generic-sop and doesn’t require to abandon it. We believe our approach is more useful for those already adopted generic-sop as their main GP solution, as well as to the users of other widespread GP libraries not based on fixed points (mainly, GHC.Generics), see the next subsection.

# Relation to fixed-point and other approaches to GP (Rev_1_&_Rev_3)

Fixed-point approach to GP is very popular in the literature, so we omit its description. It was pointed out, this undermines the motivation for our work. In future revisions, we show it is more heavy-weight. We also try to do a statistical analysis of the GP libraries used in open-source code. We believe, that the most widespread approach is the one supported natively by the compiler — GHC.Generics. It is a binary sum-of-products, “pointless” approach. We took SOP instead just because SOP uses more recent compiler extensions allowing cleaner n-ary sum-of-products, but still “pointless” representation.

We are also going to support our claim about the applicability of our approach, as Reviewer_3 requested, by providing an example with GHC.Generics in future revisions.

# Misc

## Review_1

We admit that we could do a better job in the introductory part of the paper. At the same time, we are already running out of space, so this won’t come without the price of throwing out technical content. We agree that for ESOP audience the balance between introduction and technicalities is not ideal.

> Is a cascade of pairs really all that problematic? Why? The authors merely write "They form an expressive instrument for defining generic functions in a more succinct and high-level style

Binary sums of products have at least a minus that they allow ill-formed representations, e.g. products containing nested sums. This, in turn, may potentially lead to errors in a generic program. The SOP encoding prohibits this.

## Review_3

We agree that a formal description of the approach would be invaluable. Alas, we are even not aware of good examples formalizing GP idioms (we based our work mostly on what we see before in [30],[26],[22]). We believe, this is definitely a topic for a whole separate paper.
